---
title: "Learning Note: Parametric Survival Fitting for TA1092"
subtitle: "Step-by-step R workflow using reconstructed pseudo-IPD"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    css: fit_parametric_survival_1.css
    include-after-body:
      - text: |
          <script>
          document.addEventListener('DOMContentLoaded', function () {
            ['level2', 'level3'].forEach(function (lvl) {
              var tag = lvl === 'level2' ? 'h2' : 'h3';
              document.querySelectorAll('section.' + lvl).forEach(function (sec) {
                var hd = sec.querySelector(':scope > ' + tag);
                if (!hd) return;

                // arrow indicator
                var arrow = document.createElement('span');
                arrow.innerHTML = ' &#9660;';
                arrow.style.cssText = [
                  'font-size:0.6em', 'color:#bbb', 'vertical-align:middle',
                  'margin-left:6px', 'display:inline-block'
                ].join(';');
                hd.appendChild(arrow);
                hd.style.cursor = 'pointer';
                hd.title = 'Click to collapse / expand';

                var body = Array.from(sec.children).filter(function (el) {
                  return el !== hd;
                });
                var open = true;

                hd.addEventListener('click', function () {
                  open = !open;
                  body.forEach(function (el) {
                    el.style.display = open ? '' : 'none';
                  });
                  arrow.innerHTML = open ? ' &#9660;' : ' &#9654;';
                });
              });
            });
          });
          </script>
execute:
  warning: false
  message: false
---

## Purpose

This note turns the batch script `fit_parametric_survival.R` into a step-by-step learning document.

The goal is not just to run code. The goal is to understand:

- what the pseudo-IPD contains;
- how `Surv()` represents survival data;
- what `flexsurvreg()` estimates;
- how AIC/BIC are used for first-pass model comparison;
- why AIC-selected survival curves still need clinical and EAG-style critique.

## 1. Load Packages

This block loads the R packages used in this workflow.

```{r load-packages}
library(survival)
library(flexsurv)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
library(knitr)

tibble(
  package = c("survival", "flexsurv", "dplyr", "tidyr", "purrr", "readr", "ggplot2", "knitr"),
  role = c(
    "Create survival objects and Kaplan-Meier estimates",
    "Fit parametric survival models",
    "Data manipulation",
    "Unnest list-columns",
    "Loop over curves and distributions",
    "Read and write CSV files",
    "Plot KM curves and extrapolations",
    "Display readable tables"
  )
) |>
  kable()
```

## 2. Define File Paths

The main input is the combined long-format pseudo-IPD file.

```{r define-paths}
base_dir <- normalizePath("..", mustWork = TRUE)
input_path <- file.path(base_dir, "survival_inputs", "survival_ipd_long.csv")
output_dir <- file.path(base_dir, "survival_models")
plot_dir <- file.path(output_dir, "plots")

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

tibble(
  object = c("base_dir", "input_path", "output_dir", "plot_dir"),
  path = c(base_dir, input_path, output_dir, plot_dir),
  exists = c(dir.exists(base_dir), file.exists(input_path), dir.exists(output_dir), dir.exists(plot_dir))
) |>
  kable()
```

## 3. Read the Survival Input Dataset

The dataset is reconstructed pseudo-IPD. Each row is one pseudo-patient.

`status = 1` means the event happened. For PFS this means progression/death; for OS this means death.

```{r read-input-data}
ipd <- read_csv(input_path, show_col_types = FALSE)

ipd |>
  head(3) |>
  knitr::kable()
```

```{r transform-input-data}
ipd <- ipd |>
  mutate(
    subgroup = factor(subgroup, levels = c("dMMR", "pMMR")),
    endpoint = factor(endpoint, levels = c("PFS", "OS")),
    treatment = factor(treatment, levels = c("placebo_chemo", "pembro_chemo")),
    status = as.integer(status),
    time_months = as.numeric(time_months),
    curve_id = paste(subgroup, endpoint, treatment, sep = "_")
  )

ipd |>
  count(subgroup, endpoint, treatment, status_label) |>
  arrange(subgroup, endpoint, treatment, status_label) |>
  kable()
```

## 4. Inspect One Curve

Before fitting every curve, inspect one explicitly named curve: **pMMR PFS, pembro arm**.

```{r inspect-pmmr-pfs-pembro-ipd}
pmmr_pfs_pembro_ipd <- ipd |>
  filter(subgroup == "pMMR", endpoint == "PFS", treatment == "pembro_chemo")

pmmr_pfs_pembro_ipd |>
  head(3) |>
  knitr::kable()

tibble(
  curve = "pMMR PFS pembrolizumab + chemotherapy",
  n = nrow(pmmr_pfs_pembro_ipd),
  events = sum(pmmr_pfs_pembro_ipd$status),
  censored = sum(pmmr_pfs_pembro_ipd$status == 0),
  max_follow_up_months = max(pmmr_pfs_pembro_ipd$time_months)
) |>
  kable()
```

## 5. Create a Survival Object

`Surv(time, status)` is the standard R representation of right-censored survival data.

```{r create-surv-object}
pmmr_pfs_pembro_surv <- Surv(pmmr_pfs_pembro_ipd$time_months, pmmr_pfs_pembro_ipd$status)

tibble(
  raw_time = pmmr_pfs_pembro_ipd$time_months[1:5],
  raw_status = pmmr_pfs_pembro_ipd$status[1:5],
  surv_object = as.character(pmmr_pfs_pembro_surv[1:5])
) |>
  kable()
```

## 6. Plot the Reconstructed KM Curve

This is the Kaplan-Meier curve reconstructed from pseudo-IPD.

```{r plot-pmmr-pfs-pembro-km}
pmmr_pfs_pembro_km <- survfit(Surv(time_months, status) ~ 1, data = pmmr_pfs_pembro_ipd)

pmmr_pfs_pembro_km_df <- tibble(
  time_months = c(0, pmmr_pfs_pembro_km$time),
  survival = c(1, pmmr_pfs_pembro_km$surv)
)

ggplot(pmmr_pfs_pembro_km_df, aes(time_months, survival)) +
  geom_step(linewidth = 0.8, colour = "#008B8B") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Reconstructed KM: pMMR PFS, pembro arm",
    x = "Months",
    y = "Survival probability"
  ) +
  theme_minimal()
```

## 7. Candidate Parametric Distributions

These are common survival distributions used in oncology HTA submissions and EAG critiques.

```{r candidate-distributions}
candidate_dists <- c(
  "exponential" = "exp",
  "weibull" = "weibull",
  "gompertz" = "gompertz",
  "lognormal" = "lnorm",
  "loglogistic" = "llogis",
  "gamma" = "gamma",
  "generalized_gamma" = "gengamma"
)

tibble(
  display_name = names(candidate_dists),
  flexsurv_name = unname(candidate_dists)
) |>
  kable()
```

## 8. Fit One Distribution

This block fits a Weibull model to the pMMR PFS pembro curve.

```{r fit-one-weibull}
pmmr_pfs_pembro_weibull <- flexsurvreg(
  Surv(time_months, status) ~ 1,
  data = pmmr_pfs_pembro_ipd,
  dist = "weibull"
)

as.data.frame(pmmr_pfs_pembro_weibull$res) |>
  tibble::rownames_to_column("parameter") |>
  as_tibble() |>
  select(parameter, estimate = est, lower_95 = `L95%`, upper_95 = `U95%`, se) |>
  kable(digits = 4)
```

## 9. Define a Safe Fitting Function

Some distributions can fail for sparse or immature OS data. This function catches failures and returns a row explaining what happened.

```{r define-fitting-function}
`%||%` <- function(x, y) if (is.null(x)) y else x

fit_one_distribution <- function(dat, dist_label, dist_name) {
  tryCatch(
    {
      fit <- flexsurvreg(Surv(time_months, status) ~ 1, data = dat, dist = dist_name)
      n_par <- nrow(fit$res)
      fit_aic <- as.numeric(fit$AIC %||% NA_real_)
      fit_loglik <- as.numeric(fit$loglik %||% NA_real_)

      tibble(
        dist_label = dist_label,
        dist_name = dist_name,
        fit = list(fit),
        converged = TRUE,
        error = NA_character_,
        aic = fit_aic,
        bic = -2 * fit_loglik + log(nrow(dat)) * n_par,
        loglik = fit_loglik,
        n = nrow(dat),
        events = sum(dat$status)
      )
    },
    error = function(e) {
      tibble(
        dist_label = dist_label,
        dist_name = dist_name,
        fit = list(NULL),
        converged = FALSE,
        error = conditionMessage(e),
        aic = NA_real_,
        bic = NA_real_,
        loglik = NA_real_,
        n = nrow(dat),
        events = sum(dat$status)
      )
    }
  )
}

tibble(function_created = "fit_one_distribution()", purpose = "Fit one distribution and return diagnostics") |>
  kable()
```

### What this function is doing

This function is a small wrapper around `flexsurvreg()`. The reason to write a wrapper is that we want to fit many distributions across many curves. Some fits may fail, especially for immature OS curves with few events. If one fit fails, we do not want the whole document to stop.

The function has three inputs:

```r
fit_one_distribution(dat, dist_label, dist_name)
```

- `dat`: the pseudo-IPD for one curve, for example `pmmr_pfs_pembro_ipd`.
- `dist_label`: the readable name used in output tables, for example `"weibull"`.
- `dist_name`: the name required by `flexsurvreg()`, for example `"weibull"` or `"llogis"`.

The helper:

```r
`%||%` <- function(x, y) if (is.null(x)) y else x
```

means: if `x` is missing or `NULL`, use `y` instead. Here it prevents the code from breaking if a fitted object does not contain `AIC` or `loglik`.

The core model fit is:

```r
fit <- flexsurvreg(Surv(time_months, status) ~ 1, data = dat, dist = dist_name)
```

This means:

- use `time_months` and `status` from `dat`;
- fit one survival curve, because the right-hand side is `~ 1`;
- use the parametric distribution named in `dist_name`.

The object `fit` is a fitted survival model. It contains parameter estimates, log-likelihood, AIC, and other information. We store it inside a tibble using:

```r
fit = list(fit)
```

This creates a **list-column**. Without `list()`, a complex model object cannot fit cleanly inside one table cell.

The successful-output part returns one row:

```r
tibble(
  dist_label = dist_label,
  dist_name = dist_name,
  fit = list(fit),
  converged = TRUE,
  error = NA_character_,
  aic = fit_aic,
  bic = -2 * fit_loglik + log(nrow(dat)) * n_par,
  loglik = fit_loglik,
  n = nrow(dat),
  events = sum(dat$status)
)
```

Key fields:

- `converged = TRUE`: the model fitted successfully.
- `error = NA_character_`: there was no error message.
- `aic`: model fit criterion reported by `flexsurv`.
- `bic`: calculated manually using `-2 * logLik + log(n) * number_of_parameters`.
- `n`: number of pseudo-patients in this curve.
- `events`: number of observed events.

The `tryCatch()` part is the safety mechanism:

```r
tryCatch(
  { successful fitting code },
  error = function(e) { failure output }
)
```

If `flexsurvreg()` fails, the function still returns a row. That row has:

- `fit = list(NULL)`
- `converged = FALSE`
- `error = conditionMessage(e)`
- `aic`, `bic`, `loglik` as `NA`

This is useful for modelling workflow QA. Instead of silently losing failed fits, the output table records exactly which distribution failed and why.

In HEOR work, this matters because sparse OS data can make complex models unstable. A reviewer or EAG would want to know not only which model had the best AIC, but also whether some candidate models failed, behaved oddly, or were unsuitable for extrapolation.

## 10. Compare Candidate Fits for One Curve

For the pMMR PFS pembro curve, fit all candidate distributions and rank by AIC.

```{r compare-pmmr-pfs-pembro-fits}
pmmr_pfs_pembro_fit_comparison <- map2_dfr(
  names(candidate_dists),
  candidate_dists,
  ~fit_one_distribution(pmmr_pfs_pembro_ipd, .x, .y)
)

pmmr_pfs_pembro_fit_comparison |>
  select(dist_label, converged, n, events, loglik, aic, bic, error) |>
  arrange(aic) |>
  kable(digits = 3)
```

## 11. Fit Every Curve

Now repeat the same process for all subgroup-endpoint-treatment curves.

```{r fit-all-curves}
curve_groups <- ipd |>
  group_by(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id) |>
  group_split()

fit_tbl <- map_dfr(curve_groups, function(dat) {
  meta <- dat |>
    slice(1) |>
    select(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id)

  res <- map2_dfr(
    names(candidate_dists),
    candidate_dists,
    ~fit_one_distribution(dat, .x, .y)
  )

  bind_cols(res, meta[rep(1, nrow(res)), ])
})

fit_tbl |>
  count(subgroup, endpoint, treatment, converged, name = "number_of_candidate_distributions") |>
  kable()
```

## 12. Create the Fit Summary Table

AIC is useful for first-pass comparison, but not sufficient for final HTA model choice.

```{r fit-summary}
fit_summary <- fit_tbl |>
  select(
    subgroup, endpoint, treatment, arm_role, treatment_label, curve_id,
    dist_label, dist_name, converged, error, n, events, loglik, aic, bic
  ) |>
  arrange(subgroup, endpoint, treatment, aic)

top3_fit_summary <- fit_summary |>
  filter(converged) |>
  group_by(subgroup, endpoint, treatment) |>
  arrange(aic, .by_group = TRUE) |>
  slice_head(n = 3) |>
  ungroup() |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, bic)

top3_fit_summary |>
  kable(digits = 2)
```

## 13. Extract Model Parameters

The object `parameter_tbl` stores parameter estimates from **every converged model** across all subgroups, endpoints, treatment arms, and distributions.

The table displayed below is only an example: it filters `parameter_tbl` to show **pMMR PFS, pembro arm** parameters.

```{r extract-parameters}
extract_parameters <- function(fit) {
  if (is.null(fit)) {
    return(tibble(parameter = character(), estimate = numeric(), se = numeric()))
  }

  as.data.frame(fit$res) |>
    tibble::rownames_to_column("parameter") |>
    as_tibble() |>
    transmute(parameter, estimate = est, se = se)
}

parameter_tbl <- fit_tbl |>
  filter(converged) |>
  mutate(parameters = map(fit, extract_parameters)) |>
  select(subgroup, endpoint, treatment, curve_id, dist_label, parameters) |>
  unnest(parameters)

parameter_tbl |>
  filter(subgroup == "pMMR", endpoint == "PFS", treatment == "pembro_chemo") |>
  arrange(dist_label, parameter) |>
  transmute(
    distribution = dist_label,
    parameter,
    estimate = round(estimate, 4),
    std_error = round(se, 4)
  ) |>
  kable(caption = "Example display: parameter estimates for pMMR PFS, pembro arm")
```

## 14. Select the Best Fit by AIC

This is a mechanical first-pass choice. Later we will critique the tails.

```{r select-best-fit}
best_fits <- fit_tbl |>
  filter(converged, !is.na(aic)) |>
  group_by(subgroup, endpoint, treatment) |>
  slice_min(aic, n = 1, with_ties = FALSE) |>
  ungroup()

best_fits |>
  select(subgroup, endpoint, treatment, best_dist = dist_label, n, events, aic, bic) |>
  arrange(subgroup, endpoint, treatment) |>
  kable(digits = 2)
```

## 15. Predict Survival Over a Long Horizon

The extrapolation horizon is 360 months, or 30 years.

The object `extrapolated` still stores every best-fit curve over the full monthly time grid. To keep the document readable, the displayed output below is only a small sanity check for **pMMR PFS, pembro arm**.

```{r extrapolate-best-fit}
time_grid <- tibble(time_months = seq(0, 360, by = 1))

predict_survival <- function(fit, times) {
  s <- summary(fit, t = times, type = "survival")

  if (is.data.frame(s)) {
    return(as.numeric(s$est))
  }

  if (is.list(s) && length(s) == 1 && is.data.frame(s[[1]])) {
    return(as.numeric(s[[1]]$est))
  }

  map_dbl(s, ~.x$est)
}

extrapolated <- best_fits |>
  mutate(
    pred = map(
      fit,
      ~tibble(
        time_months = time_grid$time_months,
        survival = predict_survival(.x, time_grid$time_months)
      )
    )
  ) |>
  select(
    subgroup, endpoint, treatment, arm_role, treatment_label, curve_id,
    best_dist = dist_label, aic, bic, pred
  ) |>
  unnest(cols = c(pred))

extrapolated |>
  filter(
    subgroup == "pMMR",
    endpoint == "PFS",
    treatment == "pembro_chemo",
    time_months %in% c(12, 24, 60, 120)
  ) |>
  arrange(subgroup, endpoint, treatment, time_months) |>
  select(subgroup, endpoint, treatment, best_dist, time_months, survival) |>
  kable(digits = 4, caption = "Sanity-check display: pMMR PFS pembro extrapolated survival")
```

## 16. Visualise Selected Long-Run Extrapolations

Tables are not the best way to inspect extrapolated survival. For HTA work, the curve shape is more informative. Here we show two examples:

- pMMR PFS, because it is technically important for the partitioned survival model.
- dMMR OS, because it is immature and therefore useful for thinking about EAG-style tail critique.

```{r plot-selected-long-run-extrapolations}
selected_long_run <- extrapolated |>
  filter(
    (subgroup == "pMMR" & endpoint == "PFS") |
      (subgroup == "dMMR" & endpoint == "OS"),
    time_months <= 120
  ) |>
  mutate(
    curve = paste(subgroup, endpoint),
    treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
    line_label = paste0(treatment_short, " (", best_dist, ")")
  )

ggplot(selected_long_run, aes(time_months, survival, colour = line_label)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~curve) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 24)) +
  labs(
    title = "Selected AIC-best long-run extrapolations",
    subtitle = "Legend shows treatment arm and AIC-selected distribution",
    x = "Months",
    y = "Survival probability",
    colour = "Arm (distribution)"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.width = unit(1.2, "lines")
  )
```

## 17. Compare Reconstructed KM With the Best Fit

This plot is where HTA judgement begins. In this specific pMMR PFS example, the AIC-selected curves do not show an obvious visual problem. The point of the plot is more modest: it lets us check whether the fitted curve tracks the reconstructed KM reasonably before we start interpreting the long-term extrapolation.

The broader EAG-style warning is that a curve can have the best AIC and still become implausible later. Examples would include:

- a PFS curve that remains high for decades despite advanced cancer;
- an OS curve that falls below PFS, which is structurally impossible in a partitioned survival model;
- a treatment curve with an indefinitely widening benefit when clinical experts expect treatment effect waning;
- a curve tail that implies long-term survival far above age- and disease-plausible expectations.

Those problems are not necessarily visible in this pMMR PFS plot; they are the kinds of issues we will check in later model critique.

```{r plot-pmmr-pfs-best-fit}
km_points <- ipd |>
  group_by(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id) |>
  group_modify(~{
    km <- survfit(Surv(time_months, status) ~ 1, data = .x)
    tibble(time_months = c(0, km$time), survival = c(1, km$surv))
  }) |>
  ungroup()

plot_data <- extrapolated |>
  filter(subgroup == "pMMR", endpoint == "PFS", time_months <= 120) |>
  mutate(
    source = "Parametric best fit",
    treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
    curve_label = paste0(treatment_short, " (", best_dist, ")")
  ) |>
  select(subgroup, endpoint, treatment, treatment_short, curve_label, time_months, survival, source) |>
  bind_rows(
    km_points |>
      filter(subgroup == "pMMR", endpoint == "PFS") |>
      mutate(
        source = "Reconstructed KM",
        treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
        curve_label = treatment_short
      ) |>
      select(subgroup, endpoint, treatment, treatment_short, curve_label, time_months, survival, source)
  )

ggplot(plot_data, aes(time_months, survival, colour = curve_label)) +
  geom_step(
    data = filter(plot_data, source == "Reconstructed KM"),
    linewidth = 0.85,
    alpha = 0.55
  ) +
  geom_line(
    data = filter(plot_data, source == "Parametric best fit"),
    linewidth = 1.15
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 12)) +
  labs(
    title = "pMMR PFS: reconstructed KM vs AIC-selected extrapolation",
    subtitle = "Step curves are reconstructed KM; solid smooth curves are parametric best fits",
    x = "Months",
    y = "Survival probability",
    colour = "Curve"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.width = unit(1.2, "lines")
  )
```

## 18. Issue 3 Demonstration: Short Follow-up and Extrapolation Risk

The EAG's Issue 3 is not just saying that longer follow-up would be preferable. The practical concern is that KEYNOTE-868 provides relatively early survival evidence, while the economic model requires long-term PFS and OS predictions.

In the EAG report, several values are redacted in the Issue 3 table. Later consultation comments in the same committee papers refer to the median follow-up as **16.3 months**. I use that value below as a working interpretation, while noting that the value is redacted in the Issue 3 table itself.

This section uses **pMMR OS, pembro arm** as the worked example. OS is usually one of the most important drivers of life years and QALYs in oncology models, and short follow-up is especially important when OS benefit is immature.

### 18.1 EAG Logic Behind Issue 3

Before plotting anything, it is useful to translate the EAG's concern into modelling terms.

```{r issue3-eag-logic-table}
issue3_eag_logic <- tibble::tribble(
  ~`EAG point`, ~`Meaning for modelling`,
  "Median follow-up was about 16.3 months", "Most direct trial evidence covers early survival only",
  "Survival estimates for 2 years or more rely on extrapolation", "Beyond roughly 24 months, estimates are increasingly model-driven",
  "PFS and OS estimates extend out to 20 years", "The model must predict far beyond observed follow-up",
  "Median OS was not reached in the pembrolizumab arm", "The OS benefit is immature and long-term benefit is uncertain",
  "Active treatment lasts up to about 2.2 years", "Follow-up is short relative to the full treatment and post-treatment period"
)

issue3_eag_logic |>
  kable(
    caption = "Issue 3 translated into modelling questions",
    align = "l"
  )
```

The important point is that the EAG did not apply a simple universal rule such as "follow-up below X months is unacceptable". The concern comes from the combination of short follow-up, immature OS, and a model that needs to estimate survival far beyond the observed trial period.

Two judgement points are especially important here.

First, **median survival** is a maturity signal. If median OS has not been reached in the intervention arm, then fewer than half of patients in that arm have died by the data cut. That is clinically encouraging, but it also means the long-term OS curve is not well observed. The model must infer the later part of OS from limited event data.

Second, **median follow-up should be interpreted against treatment duration**. Pembrolizumab treatment in this regimen can last for around 2.2 years, but the median follow-up referred to in the committee papers is around 16.3 months. This means the median patient had not yet been followed for as long as the maximum planned active treatment period. From an EAG perspective, that creates uncertainty not only about survival during treatment, but also about what happens after treatment stops.

```{r issue3-maturity-checks}
issue3_maturity_checks <- tibble::tribble(
  ~`Maturity check`, ~`What it asks`, ~`Why it matters for Issue 3`,
  "Median OS reached?", "Has at least half the arm experienced the OS event?", "If median OS is not reached, long-term OS benefit is inferred rather than directly observed.",
  "Median follow-up vs active treatment duration", "Has the median patient been observed through the full planned treatment period and beyond?", "If median follow-up is shorter than maximum treatment duration, there is limited evidence on post-treatment survival and treatment-effect durability.",
  "2-year boundary", "Are estimates beyond roughly 24 months still supported by enough observed data?", "The EAG considered 2 years+ estimates highly susceptible to extrapolation error.",
  "20-year projection", "How far beyond trial follow-up does the economic model project?", "The longer the projection, the more long-term QALYs depend on model structure and tail assumptions."
)

issue3_maturity_checks |>
  kable(
    caption = "Issue 3 maturity checks used to interpret short follow-up",
    align = "l"
  )
```

### 18.2 Observed Evidence Versus the 20-Year Projection

The first technical check is descriptive: how much reconstructed follow-up do we have, and how does it compare with the 20-year projection horizon discussed by the EAG?

```{r issue3-observed-vs-modelled-horizon}
issue3_projection_horizon_months <- 240
issue3_median_follow_up_months <- 16.3
issue3_eag_two_year_boundary <- 24
issue3_active_treatment_months <- (30 + 84) * 7 / 30.4375

issue3_horizon_summary <- ipd |>
  group_by(subgroup, endpoint, treatment, treatment_label) |>
  summarise(
    n = n(),
    events = sum(status),
    censored = sum(status == 0),
    max_reconstructed_follow_up_months = max(time_months),
    model_horizon_months = issue3_projection_horizon_months,
    extrapolation_ratio = model_horizon_months / max_reconstructed_follow_up_months,
    .groups = "drop"
  ) |>
  arrange(endpoint, subgroup, treatment)

issue3_horizon_summary |>
  transmute(
    curve = paste(
      subgroup,
      endpoint,
      if_else(treatment == "pembro_chemo", "Pembro", "Placebo")
    ),
    events = paste0(events, "/", n),
    `max FU, mo` = round(max_reconstructed_follow_up_months, 1),
    `20-year horizon, mo` = model_horizon_months,
    `observed share of 20y` = paste0(round(100 * max_reconstructed_follow_up_months / model_horizon_months, 1), "%"),
    `20y / max FU` = paste0(round(extrapolation_ratio, 1), "x")
  ) |>
  kable(
    caption = "Observed reconstructed follow-up versus 20-year projection horizon",
    align = "l"
  )
```

This table is not a threshold test. It only shows scale. The more important question is not whether the ratio is "too high" in isolation, but whether the economic conclusion depends on the extrapolated part of the curve.

For this portfolio, the table sets up the next two checks:

- where the observed evidence ends and the modelled tail begins;
- whether different plausible parametric curves produce different long-term survival.

### 18.3 Extrapolation-Zone Plot

The next plot turns Issue 3 into a visual modelling problem. It marks:

- **16.3 months**: median follow-up referred to in the committee papers;
- **24 months**: the EAG's practical boundary for "2 years+" estimates;
- **26.2 months**: approximate maximum active treatment duration based on 30 weeks plus 84 weeks;
- **24 months onward**: the region where estimates are increasingly driven by extrapolation.

```{r issue3-extrapolation-zone-plot}
issue3_curve_id <- "pMMR_OS_pembro_chemo"

issue3_km <- km_points |>
  filter(curve_id == issue3_curve_id)

issue3_best_fit <- extrapolated |>
  filter(curve_id == issue3_curve_id, time_months <= 240)

issue3_observed_boundary <- max(
  ipd |>
    filter(curve_id == issue3_curve_id) |>
    pull(time_months)
)

issue3_best_dist <- issue3_best_fit |>
  distinct(best_dist) |>
  pull(best_dist)

ggplot() +
  annotate(
    "rect",
    xmin = issue3_eag_two_year_boundary,
    xmax = 240,
    ymin = 0,
    ymax = 1,
    fill = "#F2C94C",
    alpha = 0.18
  ) +
  geom_step(
    data = issue3_km,
    aes(time_months, survival),
    colour = "#008B8B",
    linewidth = 0.9,
    alpha = 0.8
  ) +
  geom_line(
    data = issue3_best_fit,
    aes(time_months, survival),
    colour = "#B22222",
    linewidth = 1.1
  ) +
  geom_vline(
    xintercept = issue3_median_follow_up_months,
    linetype = "dashed",
    colour = "#2F80ED"
  ) +
  geom_vline(
    xintercept = issue3_eag_two_year_boundary,
    linetype = "dashed",
    colour = "#B22222"
  ) +
  geom_vline(
    xintercept = issue3_active_treatment_months,
    linetype = "dotted",
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = issue3_median_follow_up_months + 2,
    y = 0.98,
    label = "Median FU\n16.3 mo",
    hjust = 0,
    size = 3.0,
    colour = "#2F80ED"
  ) +
  annotate(
    "text",
    x = issue3_eag_two_year_boundary + 2,
    y = 0.86,
    label = "2 years+\nmodel-driven",
    hjust = 0,
    size = 3.0,
    colour = "#B22222"
  ) +
  annotate(
    "text",
    x = issue3_active_treatment_months + 2,
    y = 0.74,
    label = "Approx max\nactive treatment",
    hjust = 0,
    size = 3.0,
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = 85,
    y = 0.95,
    label = "EAG concern: long-term extrapolated region",
    hjust = 0,
    size = 3.5,
    colour = "#7A5A00"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 240), breaks = seq(0, 240, 24)) +
  labs(
    title = "Issue 3 demonstration: observed pMMR OS evidence versus extrapolated model tail",
    subtitle = paste0("AIC-selected distribution: ", issue3_best_dist),
    x = "Months",
    y = "Overall survival probability",
    caption = "Teal step curve = reconstructed KM; red smooth curve = AIC-selected parametric extrapolation."
  ) +
  theme_minimal()
```

The point of this plot is not that the red curve is necessarily wrong. The point is that the area most relevant to 5-, 10-, and 20-year survival is not directly observed in the trial. It is a modelled continuation from early evidence.

### 18.4 Distribution Fan Plot

A second EAG-style check is to ask whether several distributions fit the observed period reasonably but imply different long-term tails.

```{r issue3-distribution-fan-plot}
issue3_all_distribution_predictions <- fit_tbl |>
  filter(curve_id == issue3_curve_id, converged) |>
  mutate(
    pred = map(
      fit,
      ~tibble(
        time_months = seq(0, 240, by = 1),
        survival = predict_survival(.x, seq(0, 240, by = 1))
      )
    )
  ) |>
  select(dist_label, aic, bic, pred) |>
  unnest(pred)

ggplot() +
  annotate(
    "rect",
    xmin = issue3_eag_two_year_boundary,
    xmax = 240,
    ymin = 0,
    ymax = 1,
    fill = "#F2C94C",
    alpha = 0.14
  ) +
  geom_step(
    data = issue3_km,
    aes(time_months, survival),
    colour = "black",
    linewidth = 0.9,
    alpha = 0.55
  ) +
  geom_line(
    data = issue3_all_distribution_predictions,
    aes(time_months, survival, colour = dist_label),
    linewidth = 0.9
  ) +
  geom_vline(
    xintercept = issue3_eag_two_year_boundary,
    linetype = "dashed",
    colour = "#B22222"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 240), breaks = seq(0, 240, 24)) +
  labs(
    title = "Issue 3 demonstration: plausible distributions can diverge after follow-up ends",
    subtitle = "The 2-year boundary is marked because the EAG considered 2 years+ estimates highly extrapolation-dependent",
    x = "Months",
    y = "Overall survival probability",
    colour = "Distribution"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
```

This is the visual reason why short follow-up matters. The observed KM can look compatible with several parametric shapes, but the long-term tail can differ materially once the model moves beyond the trial data.

### 18.5 Landmark Survival Table

The same point can be expressed numerically. This table compares predicted survival at landmark time points that are far beyond the observed follow-up.

```{r issue3-landmark-survival-table}
issue3_landmarks <- c(24, 36, 60, 120, 240)

issue3_landmark_table <- issue3_all_distribution_predictions |>
  filter(time_months %in% issue3_landmarks) |>
  mutate(
    survival_percent = round(100 * survival, 1),
    aic = round(aic, 1)
  ) |>
  select(distribution = dist_label, aic, time_months, survival_percent) |>
  tidyr::pivot_wider(
    names_from = time_months,
    values_from = survival_percent,
    names_prefix = "month_"
  ) |>
  arrange(aic)

issue3_landmark_table |>
  kable(
    caption = "pMMR OS pembro arm: landmark survival predictions by distribution (%)",
    align = "l"
  )
```

If the 10-year or 20-year predictions differ meaningfully across distributions, then the model has structural uncertainty. This is not solved by choosing the lowest AIC alone, because AIC mainly assesses fit to the available observed data. It does not prove that the extrapolated tail is clinically plausible.

### 18.6 Appraisal Interpretation

This section turns the EAG's Issue 3 into a reproducible modelling demonstration:

- the EAG's concern is translated into concrete modelling checks;
- the observed follow-up, 2-year boundary, and 20-year horizon are shown on the same scale;
- alternative distributions show the range of plausible long-term tails;
- landmark predictions quantify how much the tails differ;
- the interpretation links directly to NICE/EAG judgement about survival extrapolation uncertainty.

The key portfolio message is: short follow-up matters when the cost-effectiveness result depends on survival benefits generated beyond the observed trial period. The issue is not that extrapolation is avoidable; oncology economic models often require extrapolation. The EAG concern is that, with immature OS and limited follow-up, the long-term survival benefit may be driven more by modelling assumptions than by observed trial evidence.

## 19. Starter Cost-Effectiveness Module

The previous sections focus on survival modelling. To connect Issue 5 to the economic model, we need to show how PFS and OS extrapolations feed into cost-effectiveness.

This is not yet a NICE base-case replication. It is a transparent, simplified partitioned survival model (PSM) that uses extracted public parameters where I can currently verify them, and clearly labels remaining illustrative or unverified inputs. The purpose is to move from a pure survival-fitting exercise toward a mock cost-effectiveness case.

- PFS and OS curves determine health-state occupancy.
- Health-state occupancy determines QALYs and costs.
- OS tail assumptions can materially change incremental QALYs and the ICER.

### 19.1 Economic Inputs: Extracted Versus Still Illustrative

The table below separates the inputs into three groups:

- **Extracted:** taken from the NotebookLM extraction output and plausible to use in this mock case.
- **Needs source verification:** extracted or implied, but should be checked against the original PDF before being treated as final.
- **Illustrative:** used only to keep the model runnable until a verified parameter is abstracted.

```{r starter-ce-inputs}
ce_inputs <- tibble::tribble(
  ~Input, ~Value, ~Status, ~Note,
  "Time horizon", "35 years", "Extracted", "Lifetime horizon in the economic model",
  "Cycle length", "1 week", "Extracted", "Consistent with weekly cost and utility calculations",
  "Annual discount rate", "3.5%", "Extracted", "Costs and health effects",
  "Perspective", "UK NHS and Personal Social Services", "Extracted", "Not directly used in this simplified calculation",
  "Model structure", "3-state PSM", "Extracted", "Progression-free, progressed disease, death",
  "Pembro treatment duration", "20 cycles; 6 Q3W combination + 14 Q6W maintenance", "Extracted", "Simplified as 18 weeks + 84 weeks of active pembro cost",
  "Pembro stopping rule", "24 months / 2 years", "Extracted", "Approximated by the 102-week active dosing schedule in this mock model",
  "Chemotherapy duration", "6 cycles Q3W", "Extracted", "Applied as 18 weeks of active chemotherapy cost in this mock model",
  "Pembrolizumab list price", "&#163;2,630 per 100 mg vial", "Extracted", "List price; confidential commercial arrangement not available",
  "Pembrolizumab combination cost", "&#163;5,260 pembro drug + &#163;106 chemotherapy drug + &#163;277 admin per Q3W cycle", "Extracted and inferred", "Before confidential discount; chemotherapy drug cost is added using the comparator RDI-adjusted value",
  "Pembrolizumab maintenance cost", "&#163;10,520 drug + &#163;217 admin per Q6W cycle", "Extracted", "Before confidential discount; smoothed to weekly cost in this mock model",
  "Comparator chemotherapy cost", "&#163;106 drug + &#163;277 admin per Q3W cycle", "Extracted", "RDI-adjusted drug cost plus complex chemotherapy administration",
  "Terminal care cost", "&#163;8,829.07 one-off at death", "Extracted", "Added in this mock model through weekly death increments",
  "PF utility", "0.814", "Extracted", "Updated base-case PF utility in the NotebookLM extraction",
  "PD utility", "0.720", "Extracted", "Updated base-case PD utility in the NotebookLM extraction",
  "PF resource-use cost", "&#163;73.59 on-treatment Q3W; &#163;44.15 pembro Q6W maintenance (weeks 19-102); &#163;27.59 off-treatment year 1-2; &#163;13.79 off-treatment year 3+", "Corrected from extracted inputs", "NICE updated model frequencies: 0.33 (Q3W on-treatment), 0.17 (Q6W pembro maintenance), 0.08 (off-treatment year 1-2), 0.04 (off-treatment year 3+) for CT scan/outpatient/blood test",
  "PD resource-use cost", "&#163;26.67 per week", "Inferred from extracted inputs", "Computed from extracted unit costs and weekly resource frequencies",
  "AE cost", "One-off Grade 3+ anaemia and hypertension cost", "Inferred from extracted inputs", "Duration for AE QALY decrement remains redacted, so only one-off AE costs are included here",
  "Confidential pembrolizumab discount", "Scenario parameter", "Unavailable", "Handled through pembro_discount sensitivity rather than assumed in the base case"
)

ce_inputs |>
  kable(caption = "Economic inputs used in the mock PSM", align = "l", escape = FALSE)
```

```{r starter-ce-parameters}
ce_params <- list(
  horizon_weeks = 35 * 52,
  cycle_length_years = 1 / 52,
  annual_discount_rate = 0.035,
  utility_pf = 0.814,
  utility_pd = 0.720,
  weekly_cost_pf_initial = 0.08 * 160.83 + 0.33 * 179 + 0.33 * 5,
  weekly_cost_pf_pembro_maint = 0.08 * 160.83 + 0.17 * 179 + 0.17 * 5,
  weekly_cost_pf_offtreatment = 0.08 * 160.83 + 0.08 * 179 + 0.08 * 5,
  weekly_cost_pf_offtreatment_yr3 = 0.04 * 160.83 + 0.04 * 179 + 0.04 * 5,
  weekly_cost_pd = 0.04 * 160.83 + 0.11 * 179 + 0.11 * 5,
  weekly_cost_pembro_combination = (5260 + 106 + 277) / 3,
  weekly_cost_pembro_maintenance = (10520 + 217) / 6,
  weekly_cost_comparator_chemo = (106 + 277) / 3,
  pembro_discount = 0,
  pembro_combination_weeks = 18,
  pembro_maintenance_weeks = 84,
  pembro_treatment_cap_weeks = 18 + 84,
  comparator_treatment_cap_weeks = 18,
  terminal_care_cost = 8829.07,
  one_off_ae_cost_pembro = 0.169 * 565.40 + 0.056 * 735.07,
  one_off_ae_cost_comparator = 0.116 * 565.40 + 0.052 * 735.07
)

param_display <- tibble(
  Parameter = c(
    "horizon_weeks",
    "cycle_length_years",
    "annual_discount_rate",
    "utility_pf",
    "utility_pd",
    "weekly_cost_pf_initial",
    "weekly_cost_pf_pembro_maint",
    "weekly_cost_pf_offtreatment",
    "weekly_cost_pf_offtreatment_yr3",
    "weekly_cost_pd",
    "weekly_cost_pembro_combination",
    "weekly_cost_pembro_maintenance",
    "weekly_cost_comparator_chemo",
    "pembro_discount",
    "pembro_combination_weeks",
    "pembro_maintenance_weeks",
    "pembro_treatment_cap_weeks",
    "comparator_treatment_cap_weeks",
    "terminal_care_cost",
    "one_off_ae_cost_pembro",
    "one_off_ae_cost_comparator"
  ),
  Value = c(
    "1,820 weeks",
    "1/52 year per cycle",
    "3.5% per year",
    "0.814",
    "0.720",
    scales::dollar(ce_params$weekly_cost_pf_initial, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_pembro_maint, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment_yr3, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pd, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_combination, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_maintenance, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_comparator_chemo, prefix = "&#163;"),
    "0% in base case",
    "18 weeks",
    "84 weeks",
    "102 weeks",
    "18 weeks",
    "&#163;8,829.07",
    scales::dollar(ce_params$one_off_ae_cost_pembro, prefix = "&#163;"),
    scales::dollar(ce_params$one_off_ae_cost_comparator, prefix = "&#163;")
  )
)

tibble(
  Parameter_1 = c(
    "horizon_weeks", "cycle_length_years", "annual_discount_rate",
    "utility_pf", "utility_pd", "weekly_cost_pf_initial",
    "weekly_cost_pf_pembro_maint", "weekly_cost_pf_offtreatment",
    "weekly_cost_pf_offtreatment_yr3", "weekly_cost_pd",
    "weekly_cost_pembro_combination"
  ),
  Value_1 = c(
    "1,820 weeks", "1/52 year per cycle", "3.5% per year",
    "0.814", "0.720",
    scales::dollar(ce_params$weekly_cost_pf_initial,        prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_pembro_maint,   prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment,   prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment_yr3, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pd,                prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_combination, prefix = "&#163;")
  ),
  Parameter_2 = c(
    "weekly_cost_pembro_maintenance", "weekly_cost_comparator_chemo",
    "pembro_discount", "pembro_combination_weeks", "pembro_maintenance_weeks",
    "pembro_treatment_cap_weeks", "comparator_treatment_cap_weeks",
    "terminal_care_cost", "one_off_ae_cost_pembro", "one_off_ae_cost_comparator", ""
  ),
  Value_2 = c(
    scales::dollar(ce_params$weekly_cost_pembro_maintenance,  prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_comparator_chemo,    prefix = "&#163;"),
    "0% in base case", "18 weeks", "84 weeks", "102 weeks", "18 weeks",
    "&#163;8,829.07",
    scales::dollar(ce_params$one_off_ae_cost_pembro,          prefix = "&#163;"),
    scales::dollar(ce_params$one_off_ae_cost_comparator,      prefix = "&#163;"),
    ""
  )
) |>
  kable(
    col.names = c("Parameter", "Value", "Parameter", "Value"),
    caption   = "Parameter object used by the simplified PSM",
    align     = "l",
    escape    = FALSE
  )

ce_time_grid <- tibble(
  time_weeks  = 0:ce_params$horizon_weeks,
  time_months = time_weeks * 12 / 52
)

discount_factor <- function(week, annual_rate) {
  1 / ((1 + annual_rate) ^ (week / 52))
}
```

### 19.2 pMMR Subgroup: AIC-Selected Standard Parametric Models

The pMMR analysis uses the AIC-best standard parametric distribution for each of the four curves (CT PFS, CT OS, pembro PFS, pembro OS), selected from the fitting grid in Section 14. Survival curves, state occupancy, costs, and QALYs are computed over a 35-year horizon.

#### 19.2.1 Survival Curves to Health-State Occupancy

In a three-state PSM:

- progression-free = PFS;
- progressed disease = OS - PFS;
- dead = 1 - OS.

This is why survival extrapolation matters for cost-effectiveness. If OS is optimistic, patients remain alive longer and generate more life years and QALYs.

There is one structural rule in a PSM: **PFS should not exceed OS**. PFS is the proportion alive and progression-free. OS is the proportion alive. Since every progression-free patient must also be alive, PFS should always be less than or equal to OS.

The source materials indicate that the company model included a limit to ensure that PFS could not exceed OS, applied at the per-cycle hazard level. This matters because independently fitted PFS and OS curves can otherwise cross in long-term extrapolation. I cannot reproduce the exact hazard-level implementation without the company workbook, so this public-data teaching model uses a transparent probability-level approximation:

```r
PFS_constrained = min(PFS, OS)
```

This keeps the simplified PSM structurally coherent. It is not intended to replace full curve-pair selection in a formal submission.

```{r starter-ce-state-occupancy}
#| eval: false
psm_survival <- best_fits |>
  filter(subgroup == "pMMR", endpoint %in% c("PFS", "OS")) |>
  mutate(
    arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
    pred = map(
      fit,
      ~tibble(
        time_weeks = ce_time_grid$time_weeks,
        time_months = ce_time_grid$time_months,
        survival = predict_survival(.x, ce_time_grid$time_months)
      )
    )
  ) |>
  select(arm, endpoint, pred) |>
  unnest(pred) |>
  tidyr::pivot_wider(names_from = endpoint, values_from = survival) |>
  mutate(
    PFS = pmin(PFS, OS),  # Approximation of the company model's PFS <= OS constraint.
    PF = PFS,
    PD = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )
```

```{r starter-ce-state-occupancy-plot}
#| eval: false
psm_survival |>
  select(arm, time_weeks, PF, PD, Dead) |>
  pivot_longer(cols = c(PF, PD, Dead), names_to = "state", values_to = "proportion") |>
  mutate(time_years = time_weeks / 52) |>
  ggplot(aes(time_years, proportion, fill = state)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~arm) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 35), breaks = seq(0, 35, 5)) +
  labs(
    title = "Simplified pMMR partitioned survival model",
    subtitle = "Weekly PSM with a PFS <= OS structural constraint",
    x = "Years",
    y = "Proportion of cohort",
    fill = "Health state"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### 19.2.2 Discounted Costs and QALYs

This chunk turns health-state occupancy into QALYs and costs. The model is intentionally simple, so each step is visible.

```{r starter-ce-results}
#| eval: false
discount_factor <- function(week, annual_rate) {
  1 / ((1 + annual_rate) ^ (week / 52))
}

ce_trace <- psm_survival |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost +
        PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_results <- ce_trace |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  )

ce_incremental <- ce_results |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

ce_results |>
  mutate(
    discounted_qalys = round(discounted_qalys, 3),
    discounted_costs = round(discounted_costs, 0)
  ) |>
  transmute(
    Arm = arm,
    `Discounted QALYs` = discounted_qalys,
    `Discounted costs` = discounted_costs
  ) |>
  kable(caption = "Mock-case discounted results by arm", align = "l")

ce_incremental |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs` = incremental_costs,
    ICER = icer
  ) |>
  kable(caption = "Mock-case incremental cost-effectiveness result", align = "l")
```

The absolute ICER should still not be interpreted as a TA1092 result. The model now uses more extracted public inputs, including utility values, resource-use unit costs, treatment costs, administration costs, terminal care cost, and one-off AE costs. However, confidential pembrolizumab discounts, exact commercial-in-confidence implementation details, and some AE duration assumptions remain unavailable. The useful point is structural: survival assumptions now flow through to QALYs and costs.

#### 19.2.3 Confidential Discount Sensitivity

The pembrolizumab commercial arrangement is confidential. Instead of guessing it, the mock model treats the discount as a scenario parameter. This makes the limitation explicit and lets us ask a practical question: how sensitive is the ICER to the assumed pembrolizumab acquisition discount?

```{r starter-ce-discount-sensitivity}
#| eval: false
calculate_incremental_ce <- function(trace) {
  trace |>
    group_by(arm) |>
    summarise(
      discounted_qalys = sum(qaly),
      discounted_costs = sum(total_cost),
      .groups = "drop"
    ) |>
    summarise(
      incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
        discounted_qalys[arm == "Placebo + chemo"],
      incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
        discounted_costs[arm == "Placebo + chemo"],
      icer = incremental_costs / incremental_qalys
    )
}

run_discount_sensitivity <- function(discount_value) {
  psm_survival |>
    group_by(arm) |>
    arrange(time_weeks, .by_group = TRUE) |>
    mutate(
      discount = discount_factor(time_weeks, ce_params$annual_discount_rate),
      death_increment = pmax(Dead - lag(Dead, default = 0), 0),
      pf_resource_cost = case_when(
        time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_initial,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_pembro_maint,
        time_weeks <= 104 ~
          ce_params$weekly_cost_pf_offtreatment,
        TRUE ~
          ce_params$weekly_cost_pf_offtreatment_yr3
      ),
      qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
        ce_params$cycle_length_years * discount,
      disease_management_cost = (
        PF * pf_resource_cost +
          PD * ce_params$weekly_cost_pd
      ) * discount,
      treatment_cost = case_when(
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
          PF * ce_params$weekly_cost_pembro_combination *
            (1 - discount_value) * discount,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_pembro_maintenance *
            (1 - discount_value) * discount,
        arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_comparator_chemo * discount,
        TRUE ~ 0
      ),
      ae_cost = case_when(
        arm == "Pembro + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
        arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
        TRUE ~ 0
      ),
      terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
      total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
    ) |>
    ungroup() |>
    calculate_incremental_ce() |>
    mutate(pembro_discount = discount_value)
}

map_dfr(c(0, 0.25, 0.50, 0.75), run_discount_sensitivity) |>
  mutate(
    pembro_discount = scales::percent(pembro_discount, accuracy = 1),
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    `Pembro discount` = pembro_discount,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs` = incremental_costs,
    ICER = icer
  ) |>
  kable(caption = "Illustrative confidential-discount sensitivity", align = "l")
```

#### 19.2.4 OS Tail Sensitivity

Issue 5 is about the uncertain degree of OS benefit. A simple way to demonstrate this is to reduce the long-term OS separation between arms after the EAG's 2-year boundary.

This scenario is not an EAG preferred model. It is a teaching sensitivity: after 104 weeks, the pembro OS benefit is gradually reduced so that only 50% of the modelled OS separation remains by 10 years.

```{r starter-ce-os-tail-sensitivity}
#| eval: false
apply_os_tail_reduction <- function(dat, start_week = 104, end_week = 520, final_fraction = 0.5) {
  dat |>
    select(arm, time_weeks, time_months, PFS, OS) |>
    tidyr::pivot_wider(names_from = arm, values_from = c(PFS, OS)) |>
    mutate(
      reduction_fraction = case_when(
        time_weeks <= start_week ~ 1,
        time_weeks >= end_week ~ final_fraction,
        TRUE ~ 1 - (1 - final_fraction) * (time_weeks - start_week) / (end_week - start_week)
      ),
      `OS_Pembro + chemo` = `OS_Placebo + chemo` +
        (`OS_Pembro + chemo` - `OS_Placebo + chemo`) * reduction_fraction
    ) |>
    select(time_weeks, time_months, starts_with("PFS_"), starts_with("OS_")) |>
    pivot_longer(
      cols = -c(time_weeks, time_months),
      names_to = c("endpoint", "arm"),
      names_sep = "_",
      values_to = "survival"
    ) |>
    pivot_wider(names_from = endpoint, values_from = survival) |>
    mutate(
      PF = pmin(PFS, OS),
      PD = pmax(OS - PFS, 0),
      Dead = pmax(1 - OS, 0)
    )
}

psm_survival_os_sensitivity <- apply_os_tail_reduction(psm_survival)

ce_trace_os_sensitivity <- psm_survival_os_sensitivity |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost +
        PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_scenario_results <- ce_trace_os_sensitivity |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  ) |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

bind_rows(
  ce_incremental |> mutate(scenario = "AIC-selected OS tail"),
  ce_scenario_results |> mutate(scenario = "Reduced long-term OS separation")
) |>
  select(scenario, incremental_qalys, incremental_costs, icer) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Scenario = scenario,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs` = incremental_costs,
    ICER = icer
  ) |>
  kable(caption = "Illustrative sensitivity to long-term OS benefit", align = "l")
```

This is the bridge to Issue 5. If reducing the long-term OS separation reduces incremental QALYs and increases the ICER, then the OS tail is decision-relevant. That is why the EAG focuses on whether the company's long-term OS benefit is clinically plausible.

#### 19.2.5 EAG-Preferred Spline Models: Fitting Function

The EAG preferred spline models for pMMR (Table 2) are Royston-Parmar flexible parametric models fitted with `flexsurvspline()`. Unlike standard parametric distributions, spline models can accommodate non-monotonic hazard functions and are more flexible in the observed period.

```{r pmmr-spline-function}
fit_one_spline <- function(dat, k, scale, label) {
  tryCatch({
    fit     <- flexsurvspline(Surv(time_months, status) ~ 1, data = dat, k = k, scale = scale)
    fit_aic <- as.numeric(AIC(fit) %||% NA_real_)
    fit_ll  <- as.numeric(fit$loglik %||% NA_real_)
    tibble(
      dist_label = label,
      fit        = list(fit),
      converged  = TRUE,
      error      = NA_character_,
      aic        = fit_aic,
      loglik     = fit_ll,
      n          = nrow(dat),
      events     = sum(dat$status)
    )
  }, error = function(e) {
    tibble(
      dist_label = label,
      fit        = list(NULL),
      converged  = FALSE,
      error      = conditionMessage(e),
      aic        = NA_real_,
      loglik     = NA_real_,
      n          = nrow(dat),
      events     = sum(dat$status)
    )
  })
}

tibble(
  `Function` = "fit_one_spline(dat, k, scale, label)",
  `Backend`  = "flexsurvspline()",
  `Purpose`  = "Fit Royston-Parmar spline model for one curve"
) |> kable()
```

```{r pmmr-spline-specs}
eag_pmmr_spline_specs <- tibble::tribble(
  ~subgroup, ~endpoint, ~treatment,      ~k, ~scale,   ~dist_label,
  "pMMR",    "PFS",     "placebo_chemo", 2,  "hazard", "2-knot hazards",
  "pMMR",    "OS",      "placebo_chemo", 1,  "hazard", "1-knot hazards",
  "pMMR",    "PFS",     "pembro_chemo",  1,  "hazard", "1-knot hazards",
  "pMMR",    "OS",      "pembro_chemo",  1,  "normal", "1-knot normal"
)

eag_pmmr_spline_specs |>
  kable(caption = "EAG-preferred spline specifications for pMMR (Table 2)", align = "l")
```

```{r pmmr-spline-fits}
spline_fit_rows <- vector("list", nrow(eag_pmmr_spline_specs))
for (i in seq_len(nrow(eag_pmmr_spline_specs))) {
  spec_i <- eag_pmmr_spline_specs[i, ]
  dat_i  <- ipd |> filter(
    as.character(subgroup)  == spec_i$subgroup,
    as.character(endpoint)  == spec_i$endpoint,
    as.character(treatment) == spec_i$treatment
  )
  res_i <- fit_one_spline(dat_i, spec_i$k, spec_i$scale, spec_i$dist_label)
  spline_fit_rows[[i]] <- bind_cols(
    tibble(subgroup = spec_i$subgroup,
           endpoint = spec_i$endpoint,
           treatment = spec_i$treatment),
    res_i
  )
}
eag_pmmr_spline_fits <- bind_rows(spline_fit_rows)

eag_pmmr_spline_fits |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, converged, error) |>
  kable(digits = 2, caption = "EAG-preferred spline fit summary for pMMR", align = "l")

```

#### 19.2.6 State Occupancy with EAG Spline Models

```{r pmmr-spline-state-occupancy}
get_spline_surv <- function(fit_obj, times) {
  # Unwrap list(model_obj) wrapper only if not already a flexsurv model
  if (!inherits(fit_obj, "flexsurvreg")) fit_obj <- fit_obj[[1]]
  s <- summary(fit_obj, t = times, type = "survival", ci = FALSE)
  if (is.data.frame(s))                    return(as.numeric(s$est))
  if (is.list(s) && is.data.frame(s[[1]])) return(as.numeric(s[[1]]$est))
  if (is.list(s) && is.numeric(s[[1]]))    return(as.numeric(s[[1]]))
  stop(paste("Cannot extract survival: class(s)=", class(s), "class(s[[1]])=", class(s[[1]])))
}

pmmr_sp_curves <- eag_pmmr_spline_fits |>
  filter(converged) |>
  mutate(arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo")) |>
  select(arm, endpoint, fit)

surv_rows <- vector("list", nrow(pmmr_sp_curves))
for (i in seq_len(nrow(pmmr_sp_curves))) {
  surv_rows[[i]] <- tibble(
    arm         = pmmr_sp_curves$arm[[i]],
    endpoint    = pmmr_sp_curves$endpoint[[i]],
    time_weeks  = ce_time_grid$time_weeks,
    time_months = ce_time_grid$time_months,
    surv        = get_spline_surv(pmmr_sp_curves$fit[[i]], ce_time_grid$time_months)
  )
}

psm_pmmr_spline <- bind_rows(surv_rows) |>
  tidyr::pivot_wider(
    id_cols     = c(arm, time_weeks, time_months),
    names_from  = endpoint,
    values_from = surv
  ) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

psm_pmmr_spline |>
  select(arm, time_weeks, PF, PD, Dead) |>
  pivot_longer(cols = c(PF, PD, Dead), names_to = "state", values_to = "proportion") |>
  mutate(time_years = time_weeks / 52) |>
  ggplot(aes(time_years, proportion, fill = state)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~arm) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 35), breaks = seq(0, 35, 5)) +
  labs(
    title    = "pMMR partitioned survival model: EAG-preferred spline models",
    subtitle = "CT PFS: 2-knot hazards | CT OS: 1-knot hazards | Pembro PFS: 1-knot hazards | Pembro OS: 1-knot normal",
    x = "Years", y = "Proportion of cohort", fill = "Health state"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### 19.2.7 Discounted Costs and QALYs (EAG Spline Models)

```{r pmmr-spline-ce-results}
ce_trace_pmmr_spline <- psm_pmmr_spline |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount         = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment  = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_results_pmmr_spline <- ce_trace_pmmr_spline |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  )

ce_incremental_pmmr_spline <- ce_results_pmmr_spline |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

ce_results_pmmr_spline |>
  mutate(
    discounted_qalys = round(discounted_qalys, 3),
    discounted_costs = round(discounted_costs, 0)
  ) |>
  transmute(Arm = arm,
            `Discounted QALYs` = discounted_qalys,
            `Discounted costs (&#163;)` = discounted_costs) |>
  kable(caption = "pMMR (EAG spline): discounted results by arm", align = "l", escape = FALSE)

ce_incremental_pmmr_spline |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer              = round(icer, 0)
  ) |>
  transmute(`Incremental QALYs`     = incremental_qalys,
            `Incremental costs (&#163;)` = incremental_costs,
            `ICER (&#163;/QALY)`        = icer) |>
  kable(caption = "pMMR (EAG spline): incremental cost-effectiveness", align = "l", escape = FALSE)
```

### 19.3 dMMR Subgroup: EAG-Preferred Models (Table 2)

This section computes the dMMR ICER using the EAG's preferred distributions from Table 2 of the EAG supplementary report.

Two structural caveats apply before interpreting the output:

- The EAG's pMMR preferred models are now implemented as spline models in Section 19.2 above.
- Utility values (PF = 0.814, PD = 0.720) were collected from the pMMR cohort only. KEYNOTE-868 did not collect dMMR HRQoL data due to insufficient power in the smaller dMMR cohort. Applying pMMR utilities to dMMR patients is an approximation with additional uncertainty.

#### 19.3.1 Select EAG-Preferred Distributions

```{r dmmr-eag-model-selection}
eag_dmmr_selections <- tibble::tribble(
  ~subgroup, ~endpoint, ~treatment,      ~dist_label,
  "dMMR",    "PFS",     "placebo_chemo", "generalized_gamma",
  "dMMR",    "OS",      "placebo_chemo", "exponential",
  "dMMR",    "PFS",     "pembro_chemo",  "loglogistic",
  "dMMR",    "OS",      "pembro_chemo",  "loglogistic"
)

eag_dmmr_fits <- fit_tbl |>
  inner_join(eag_dmmr_selections, by = c("subgroup", "endpoint", "treatment", "dist_label")) |>
  filter(converged)

eag_dmmr_fits |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, bic) |>
  arrange(endpoint, treatment) |>
  kable(digits = 2, caption = "EAG-preferred distributions for dMMR (Table 2)", align = "l")
```

#### 19.3.2 State Occupancy

```{r dmmr-state-occupancy}
psm_dmmr <- eag_dmmr_fits |>
  mutate(
    arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
    pred = map(
      fit,
      ~tibble(
        time_weeks = ce_time_grid$time_weeks,
        time_months = ce_time_grid$time_months,
        survival = predict_survival(.x, ce_time_grid$time_months)
      )
    )
  ) |>
  select(arm, endpoint, pred) |>
  unnest(pred) |>
  tidyr::pivot_wider(names_from = endpoint, values_from = survival) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

psm_dmmr |>
  select(arm, time_weeks, PF, PD, Dead) |>
  pivot_longer(cols = c(PF, PD, Dead), names_to = "state", values_to = "proportion") |>
  mutate(time_years = time_weeks / 52) |>
  ggplot(aes(time_years, proportion, fill = state)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~arm) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 35), breaks = seq(0, 35, 5)) +
  labs(
    title = "dMMR partitioned survival model (EAG-preferred distributions)",
    subtitle = "EAG: generalised gamma / exponential / log-logistic / log-logistic",
    x = "Years",
    y = "Proportion of cohort",
    fill = "Health state"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### 19.3.3 Discounted Costs and QALYs

```{r dmmr-ce-results}
ce_trace_dmmr <- psm_dmmr |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount        = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_results_dmmr <- ce_trace_dmmr |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  )

ce_incremental_dmmr <- ce_results_dmmr |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

ce_results_dmmr |>
  mutate(
    discounted_qalys = round(discounted_qalys, 3),
    discounted_costs = round(discounted_costs, 0)
  ) |>
  transmute(
    Arm = arm,
    `Discounted QALYs` = discounted_qalys,
    `Discounted costs (&#163;)` = discounted_costs
  ) |>
  kable(caption = "dMMR: discounted results by arm (EAG-preferred models)", align = "l", escape = FALSE)

ce_incremental_dmmr |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs (&#163;)` = incremental_costs,
    `ICER (&#163;/QALY)` = icer
  ) |>
  kable(caption = "dMMR: incremental cost-effectiveness (EAG-preferred models)", align = "l", escape = FALSE)
```

#### 19.3.4 pMMR vs dMMR Summary

```{r pmmr-dmmr-comparison}
bind_rows(
  ce_incremental_pmmr_spline |> mutate(subgroup = "pMMR", model = "EAG-preferred spline (Table 2)"),
  ce_incremental_dmmr        |> mutate(subgroup = "dMMR", model = "EAG-preferred standard (Table 2)")
) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Subgroup = subgroup,
    `Curve selection` = model,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs (&#163;)` = incremental_costs,
    `ICER (&#163;/QALY)` = icer
  ) |>
  kable(
    caption = "pMMR (EAG spline) vs dMMR (EAG standard) cost-effectiveness comparison",
    align = "l",
    escape = FALSE
  )
```

Interpretation notes:

- pMMR curves are AIC-selected standard parametric distributions; EAG preferred spline models for pMMR are not yet implemented in this document.
- dMMR curves follow EAG Table 2 preferences: generalised gamma (CT PFS), exponential (CT OS), log-logistic (pembro PFS and OS).
- Both subgroups apply pMMR-derived utility values (0.814 / 0.720) because dMMR HRQoL data were not collected in KEYNOTE-868.
- Pembrolizumab confidential discount is set to 0% in both calculations; the real ICER under PAS pricing would be lower.

## 20. Save Outputs

This final block writes the same output files as the batch script.

```{r save-outputs}
write_csv(fit_summary, file.path(output_dir, "parametric_fit_summary.csv"))
write_csv(parameter_tbl, file.path(output_dir, "parametric_fit_parameters.csv"))
write_csv(extrapolated, file.path(output_dir, "best_fit_extrapolated_survival.csv"))
write_csv(top3_fit_summary, file.path(output_dir, "top3_fit_summary_by_curve.csv"))

tibble(
  output_file = c(
    "parametric_fit_summary.csv",
    "parametric_fit_parameters.csv",
    "best_fit_extrapolated_survival.csv",
    "top3_fit_summary_by_curve.csv"
  ),
  exists = file.exists(file.path(output_dir, output_file))
) |>
  kable()
```

## What to Critique Next

The next analytical step is not more code. It is interpretation:

- Which best-fit curves produce clinically plausible tails?
- Are OS curves mature enough for long-term extrapolation?
- Does PFS ever exceed OS?
- Do fitted curves align with company, EAG, and committee preferred assumptions?
- Where would an EAG challenge a purely AIC-driven model choice?
