---
title: "Targeted Maximum Likelihood Estimation & Sensitivity Analysis"
subtitle: "Doubly Robust Causal Inference & Unmeasured Confounding Audits"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "R Code"
    number-sections: true
---

*Advanced Causal Inference for External Control Arms and Observational Real-World Evidence*

## Introduction: The Limits of Single Robustness & AIPW

In health technology assessment (HTA), observational real-world data (RWD) is frequently used to construct External Control Arms (ECAs) when randomized controlled trials (RCTs) are unfeasible. Standard propensity score adjustments, such as Propensity Score Matching (PSM) and Inverse Probability Treatment Weighting (IPTW), rely entirely on the correct specification of the treatment assignment model:
$$g(X) = P(D = 1 \mid X)$$

If the propensity score model omits crucial confounders or misspecifies their functional forms (e.g., ignoring non-linearities or interactions), these single-robust estimators yield biased estimates of the treatment effect.

### Doubly Robust Estimators

To address this, doubly robust estimators combine the treatment model with an outcome regression model:
$$m(D, X) = E[Y \mid D, X]$$

This ensures that the treatment effect estimate is unbiased if *either* the treatment model *or* the outcome model is correctly specified.

### Why TMLE Over AIPW?

While Augmented Inverse Probability Weighting (AIPW) is also doubly robust, **Targeted Maximum Likelihood Estimation (TMLE)**—developed by Mark van der Laan and colleagues—is preferred in HTA and regulatory submissions for several reasons:

- **Substitution Estimator (Respecting Bounds)**: AIPW is an influence function-based estimator. Because it adds a weighted residual correction, it can yield predicted counterfactual values that lie outside the physical bounds of the outcome space. For example, in binary or time-to-event outcomes, AIPW can predict survival probabilities that are greater than $1$ or less than $0$. TMLE is a *substitution estimator*, which mathematically guarantees that the estimated probabilities respect the natural bounds of the parameter space (i.e., restricting survival estimates strictly within $[0, 1]$).
- **Semiparametric Efficiency**: TMLE is asymptotically efficient, meaning it achieves the lowest possible variance (the semiparametric efficiency bound) among all regular estimators when both models are correctly specified.
- **Ensemble Integration**: TMLE natively integrates machine learning ensembles (like `SuperLearner`) to model both the outcome and treatment mechanisms, minimizing the risk of model misspecification bias.

---

## The 5-Step TMLE Algorithm

TMLE operates by taking an initial estimate of the outcome regression and "targeting" it using the treatment assignment model to directly optimize the parameter of interest (e.g., the Average Treatment Effect, ATE).

The mathematical implementation follows five steps:

### Step 1: Initial Estimation of the Outcome Regression ($Q_0$)

We estimate the conditional expectation of the outcome $Y$ given treatment $D$ and baseline covariates $W$:
$$Q_0(D, W) = E[Y \mid D, W]$$

This model is fitted using a machine learning library or parametric regression. We then generate two predicted values for each individual $i$:

- The predicted outcome under treatment: $Q_0(1, W_i)$
- The predicted outcome under control: $Q_0(0, W_i)$

### Step 2: Estimation of the Propensity Score ($g_0$)

We estimate the probability of treatment assignment given covariates:
$$g(W) = P(D = 1 \mid W)$$

For each individual $i$, we compute their estimated propensity score $g(W_i)$.

### Step 3: Computation of the Clever Covariate ($H$)

We construct a secondary covariate, known as the **Clever Covariate** $H(D, W)$, which represents the inverse probability weight adjusted for the treatment actually received:
$$H(D, W) = \frac{D}{g(W)} - \frac{1 - D}{1 - g(W)}$$

For each individual $i$, we evaluate two values of the Clever Covariate:

- The covariate value under treatment ($D_i = 1$): $H(1, W_i) = \frac{1}{g(W_i)}$
- The covariate value under control ($D_i = 0$): $H(0, W_i) = -\frac{1}{1 - g(W_i)}$

### Step 4: The Fluctuation (Update) Step

We update the initial outcome prediction $Q_0$ by fitting a logistic regression of the observed outcome $Y$ on the Clever Covariate $H(D, W)$, using the logit of the initial prediction $Q_0(D, W)$ as a fixed offset:
$$\text{logit}\left(Q_1(D, W)\right) = \text{logit}\left(Q_0(D, W)\right) + \epsilon \cdot H(D, W)$$

Here, the coefficient $\epsilon$ (the fluctuation parameter) is estimated via Maximum Likelihood. 

If the initial outcome regression $Q_0$ is already perfectly calibrated, the residual error $(Y - Q_0)$ will be orthogonal to the weights, and the estimated coefficient $\epsilon$ will be close to $0$. If $Q_0$ is biased, $\epsilon$ will deviate from $0$, shifting the predictions to remove the bias.

We then update the counterfactual predictions for all individuals:

- Updated survival under treatment: $Q_1(1, W_i) = \text{logit}^{-1}\left( \text{logit}(Q_0(1, W_i)) + \epsilon \cdot H(1, W_i) \right)$
- Updated survival under control: $Q_1(0, W_i) = \text{logit}^{-1}\left( \text{logit}(Q_0(0, W_i)) + \epsilon \cdot H(0, W_i) \right)$

### Step 5: Parameter Estimation via Substitution

Finally, we calculate the doubly robust ATE by averaging the updated counterfactual predictions across the sample:
$$\hat{ATE}_{TMLE} = \frac{1}{n} \sum_{i=1}^n \left[ Q_1(1, W_i) - Q_1(0, W_i) \right]$$

Because the update step satisfies the efficient score equation, standard errors and confidence intervals are computed asymptotically using the variance of the influence function.

---

## Sensitivity Analysis: The E-Value

A critical assumption in causal inference is **Exchangeability (Unconfoundedness)**—the assumption that all confounders ($W$) that simultaneously drive treatment assignment and clinical outcomes have been measured and adjusted for:
$$Y^{(1)}, Y^{(0)} \perp D \mid W$$

In real-world observational registries, this assumption is unverifiable and frequently threatened by **Unmeasured Confounding** (e.g., unrecorded tumor biomarkers, sub-clinical frailty, or physician subjective preferences).

### The E-Value Formulation

The **E-Value** (VanderWeele and Ding, 2017) quantifies the minimum strength of association that an unmeasured confounder ($U$) would need to have with both the treatment ($D$) and the outcome ($Y$), on the Risk Ratio (RR) scale, to fully explain away an observed treatment effect.

For a point estimate representing a Relative Risk ($RR_{obs} > 1$), the E-value is calculated as:
$$E = RR_{obs} + \sqrt{RR_{obs} (RR_{obs} - 1)}$$

If the observed treatment effect is protective ($RR_{obs} < 1$), we first invert the estimate ($1/RR_{obs}$) before applying the formula.

### Interpretation in HTA Appraisals

The E-value provides an intuitive benchmark for HTA committees:

- If the E-value is small (e.g., $E = 1.2$), a weak unmeasured confounder that increases both the probability of treatment and the probability of survival by only 1.2-fold could nullify the treatment effect. The evidence is highly sensitive to hidden bias.
- If the E-value is large (e.g., $E = 2.8$), a hidden confounder would need to have a strong 2.8-fold association with both treatment and outcome to explain away the effect. If known, measured confounders (like ECOG score) only have associations of 1.5-fold to 2.0-fold, it is highly unlikely that an unmeasured factor of such extreme magnitude exists, establishing the robustness of the causal claim.

---

## R Implementation Pipeline

This pipeline simulates a cohort ($N=500$) with baseline confounding, fits the ATE using TMLE with ensemble learners, and conducts an E-value sensitivity analysis.

```{r}
#| message: false
#| warning: false

library(WeightIt)
library(survey)
library(tmle)
library(SuperLearner)
library(EValue)

set.seed(2026)
n <- 500

# 1. Simulate Confounders
age <- rnorm(n, mean = 65, sd = 10)
ecog <- rbinom(n, 1, 0.4) # 1 = Poor performance status

# 2. Simulate Treatment Assignment (Confounding by Indication)
# Elderly and sicker patients are more likely to receive the treatment (D = 1)
ps_true <- plogis(-2 + 0.05 * age + 1.2 * ecog)
treatment <- rbinom(n, 1, ps_true)

# 3. Simulate Survival Outcome (1-Year Survival)
# True Treatment Effect (Risk Difference) = +15.5% (coefficient = 0.8 in log-odds scale)
y_prob_1 <- plogis(1 - 0.03 * age - 1.0 * ecog + 0.8 * 1)
y_prob_0 <- plogis(1 - 0.03 * age - 1.0 * ecog + 0.8 * 0)
true_ate <- mean(y_prob_1) - mean(y_prob_0)

survival_1yr <- ifelse(treatment == 1,
                       rbinom(n, 1, y_prob_1),
                       rbinom(n, 1, y_prob_0))

df <- data.frame(age, ecog, treatment, survival_1yr)

# Calculate Naive (Unadjusted) Estimate
naive_treated <- mean(df$survival_1yr[df$treatment == 1])
naive_control <- mean(df$survival_1yr[df$treatment == 0])
naive_ate <- naive_treated - naive_control
```

### Doubly Robust Estimation via TMLE

We estimate the treatment effect using the `tmle` package. We specify `SuperLearner` ensemble models (with GLM and Mean libraries) to predict both the treatment assignment and outcome regressions.

```{r}
# Fit TMLE and suppress package-level output
sl_lib <- c("SL.glm", "SL.mean")

tmle_fit <- suppressMessages(tmle(
  Y             = df$survival_1yr,
  A             = df$treatment,
  W             = df[, c("age", "ecog")],
  Q.SL.library  = sl_lib,
  g.SL.library  = sl_lib,
  family        = "binomial"
))

# Extract results
tmle_est <- tmle_fit$estimates$ATE$psi
tmle_se  <- sqrt(tmle_fit$estimates$ATE$var.psi)
tmle_ci  <- tmle_fit$estimates$ATE$CI
tmle_pval <- tmle_fit$estimates$ATE$pvalue
epsilon  <- tmle_fit$epsilon

# Format TMLE parameter table
tmle_tbl <- data.frame(
  Parameter = c("Risk under Control (EY0)", "Risk under Exposure (EY1)", "Risk Difference (ATE)"),
  Estimate = c(tmle_fit$estimates$EY0$psi, tmle_fit$estimates$EY1$psi, tmle_est),
  "Std. Error" = c(sqrt(tmle_fit$estimates$EY0$var.psi), sqrt(tmle_fit$estimates$EY1$var.psi), tmle_se),
  "95% LCL" = c(tmle_fit$estimates$EY0$CI[1], tmle_fit$estimates$EY1$CI[1], tmle_ci[1]),
  "95% UCL" = c(tmle_fit$estimates$EY0$CI[2], tmle_fit$estimates$EY1$CI[2], tmle_ci[2]),
  "p-value" = c(NA, NA, tmle_pval),
  check.names = FALSE
)

# Round values for display
tmle_tbl[, 2:5] <- round(tmle_tbl[, 2:5], 4)
tmle_tbl[3, 6] <- format.pval(tmle_pval, eps = 0.001, digits = 3)

knitr::kable(tmle_tbl, caption = "Targeted Maximum Likelihood Estimation (TMLE) Results", align = "lrrrrr")
```

::: {.callout-note}
## Interpretation of the Fluctuation Parameter
The estimated fluctuation parameter $\epsilon$ is **`r round(epsilon, 6)`**. This value represents the degree of covariate adjustment applied during the targeting step. A non-zero $\epsilon$ indicates that the initial outcome predictions were shifted to successfully correct for treatment selection bias.
:::

### Sensitivity Analysis: E-Value Calculation

To evaluate sensitivity to unmeasured confounding, we transform the Risk Difference obtained from TMLE to the Relative Risk (RR) scale and calculate the corresponding E-value.

```{r}
# 1. Transform Risk Difference to Relative Risk
ey0_est <- tmle_fit$estimates$EY0$psi
rr_est  <- (ey0_est + tmle_est) / ey0_est
rr_lo   <- (ey0_est + tmle_ci[1]) / ey0_est
rr_hi   <- (ey0_est + tmle_ci[2]) / ey0_est

# 2. Compute E-value
ev <- evalues.RR(est = rr_est, lo = rr_lo, hi = rr_hi)
ev_df <- as.data.frame(ev)

# 3. Format E-value table
ev_tbl <- data.frame(
  "Metric / Parameter" = c("Relative Risk of Treatment (RR)", "E-value (Treatment Effect Bound)"),
  "Point Estimate" = c(sprintf("%.3f", ev_df["RR", "point"]), sprintf("%.3f", ev_df["E-values", "point"])),
  "95% CI Lower Bound" = c(sprintf("%.3f", ev_df["RR", "lower"]), sprintf("%.3f", ev_df["E-values", "lower"])),
  "95% CI Upper Bound" = c(sprintf("%.3f", ev_df["RR", "upper"]), "N/A"),
  check.names = FALSE
)

knitr::kable(ev_tbl, caption = "E-Value Analysis for Unmeasured Confounding Sensitivity", align = "lrrr")
```

The point estimate E-value of **`r round(ev_df["E-values", "point"], 2)`** indicates that an unmeasured confounder would need to have a Relative Risk association of at least **`r round(ev_df["E-values", "point"], 2)`**-fold with both treatment assignment and survival to fully explain away the observed causal survival benefit.

### E-Value Contour Visualization

The contour curve illustrates the required strength of association between the unmeasured confounder and the treatment (x-axis) and outcome (y-axis) to reduce the observed effect to the null.

```{r}
#| label: evalue-plot
#| fig-align: center
#| fig-width: 6
#| fig-height: 5

plot(ev, type = "line", 
     main = "E-value Contour: Sensitivity to Unmeasured Confounding")
```

---

## Causal Evidence Reconciliation

We reconcile the point estimates and standard errors across our observational causal pipeline compared to the simulated population ground truth.

```{r}
# Fit IPTW for comparison (from Page 2 workflow)
W <- weightit(treatment ~ age + ecog, data = df, method = "ps", estimand = "ATE")
fit_iptw <- lm_weightit(survival_1yr ~ treatment, data = df, weightit = W)
iptw_summary <- summary(fit_iptw)
iptw_est <- iptw_summary$coefficients["treatment", "Estimate"]
iptw_se  <- iptw_summary$coefficients["treatment", "Std. Error"]

# Combine into reconciliation matrix
reconciliation_tbl <- data.frame(
  Estimator = c("Ground Truth (Omniscient)", "Naive (Unadjusted)", "IPTW (Weighting Only)", "TMLE (Doubly Robust)"),
  "ATE Point Estimate" = c(true_ate, naive_ate, iptw_est, tmle_est),
  "Standard Error" = c(NA, NA, iptw_se, tmle_se),
  "95% LCL" = c(NA, NA, iptw_est - 1.96 * iptw_se, tmle_ci[1]),
  "95% UCL" = c(NA, NA, iptw_est + 1.96 * iptw_se, tmle_ci[2]),
  "Bias vs. Truth" = c(0.0000, naive_ate - true_ate, iptw_est - true_ate, tmle_est - true_ate),
  check.names = FALSE
)

# Format values for display
reconciliation_tbl[, 2:6] <- round(reconciliation_tbl[, 2:6], 4)
reconciliation_tbl[is.na(reconciliation_tbl)] <- "N/A"

knitr::kable(reconciliation_tbl, caption = "Causal Evidence Reconciliation Across Estimators", align = "lrrrrr")
```

---

## EAG Defense Strategy

When defending causal estimates derived via TMLE and E-value sensitivity analysis during EAG reviews, modellers should leverage the following arguments:

### Justifying TMLE Over Parametric Regression

> "Rather than relying on arbitrary functional forms that assume linear log-odds relations for confounding covariates, we implemented TMLE coupled with the SuperLearner ensemble. This allows the data to non-parametrically determine the correct functional relations, minimizing model misspecification bias. 
> 
> Furthermore, TMLE is a substitution estimator, meaning that unlike influence function-based estimators (like AIPW) which can yield survival predictions outside the valid $[0, 1]$ interval, TMLE is mathematically restricted to remain within the physical bounds of probability space. This ensures clinical plausibility in all extrapolated subgroups."

### Auditing Sensitivity via E-Value

> "The E-value of `r round(ev_df["E-values", "point"], 2)` demonstrates that our causal treatment effect is robust against unmeasured confounding. To invalidate this survival difference, an unmeasured baseline covariate would need to increase both the likelihood of treatment and the likelihood of survival by at least `r round(ev_df["E-values", "point"], 2)`-fold. 
> 
> In our adjusted models, the strongest measured clinical confounders (such as poor ECOG performance status) only display hazard associations of approximately 1.8-fold. It is highly implausible that a missing, unmeasured confounder of greater magnitude exists that is completely uncorrelated with the clinical indicators adjusted for in this analysis."
