---
title: "Targeted Maximum Likelihood Estimation & Sensitivity Analysis"
subtitle: "Doubly Robust Causal Inference & Unmeasured Confounding Audits"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "R Code"
    number-sections: true
---

*Advanced Causal Inference for External Control Arms and Observational Real-World Evidence*

## Introduction: The Limits of Single Robustness & AIPW

In health technology assessment (HTA), observational real-world data (RWD) is frequently used to construct External Control Arms (ECAs) when randomized controlled trials (RCTs) are unfeasible. Standard propensity score adjustments, such as Propensity Score Matching (PSM) and Inverse Probability Treatment Weighting (IPTW), rely entirely on the correct specification of the treatment assignment model:
$$g(X) = P(D = 1 \mid X)$$

If the propensity score model **omits crucial confounders** or **misspecifies their functional forms (e.g., ignoring non-linearities or interactions)**, these single-robust estimators yield biased estimates of the treatment effect.

### Doubly Robust Estimators

To address this, doubly robust estimators combine the treatment model with an outcome regression model. For a detailed algebraic proof of the double robustness property, see the [AIPW section in Page 2](./02_PSM_IPTW_AIPW.qmd#augmented-inverse-probability-weighting-aipw):
$$m(D, X) = E[Y \mid D, W]$$

This ensures that the treatment effect estimate is unbiased if *either* the treatment model *or* the outcome model is correctly specified.

### Why TMLE Over AIPW?

While Augmented Inverse Probability Weighting (AIPW) is also doubly robust, **Targeted Maximum Likelihood Estimation (TMLE)**—developed by Mark van der Laan and colleagues—is preferred in HTA and regulatory submissions for several reasons:

- **Substitution Estimator (Respecting Bounds)**: AIPW is an influence function-based estimator. Because it adds a weighted residual correction, it can yield predicted counterfactual values that lie outside the physical bounds of the outcome space. For example, in binary or time-to-event outcomes, **AIPW can predict survival probabilities that are greater than $1$ or less than $0$.** TMLE is a *substitution estimator*, which mathematically guarantees that the estimated probabilities respect the natural bounds of the parameter space (i.e., restricting survival estimates strictly within $[0, 1]$).
- **Semiparametric Efficiency**: TMLE is asymptotically **efficient**, meaning it achieves the lowest possible variance (the semiparametric efficiency bound) among all regular estimators when both models are correctly specified.
- **Ensemble Integration**: TMLE natively integrates machine learning ensembles (like `SuperLearner`) to model **both** the outcome and treatment mechanisms, **minimizing the risk of model misspecification bias**.

## The 5-Step TMLE Algorithm with Integrated R Code

TMLE operates by taking an initial estimate of the outcome regression and "targeting" it using the treatment assignment model to directly optimize the parameter of interest (e.g., the Average Treatment Effect, ATE).

The mathematical implementation follows five steps. To illustrate the mechanics transparently, we will run the manual calculations in R alongside each step using a simulated cohort ($N=500$) with baseline confounding.

### Cohort Simulation Setup & Baseline Characteristics

The simulated cohort represents an observational registry of 500 patients:

- **Baseline Confounders**: Age (normally distributed around 65) and ECOG performance status (binary indicator where 1 denotes poor status, with 40% prevalence).
- **Confounding by Indication**: Older and sicker patients are preferentially prescribed the active treatment (probability of treatment increases with both age and poor ECOG status).
- **Survival Outcome**: 1-year survival probability is driven by both confounders (older age and poor ECOG status lower survival) and treatment (which increases survival by a true risk difference of +15.5%).

The baseline characteristics table below highlights the treatment assignment bias: older, sicker patients are heavily concentrated in the treated group, which will mask the true efficacy in naive unadjusted analyses.

```{r}
#| message: false
#| warning: false

library(WeightIt)
library(survey)
library(tmle)
library(SuperLearner)
library(EValue)

set.seed(2026)
n <- 500

# 1. Simulate Confounders
age <- rnorm(n, mean = 65, sd = 10)
ecog <- rbinom(n, 1, 0.4) # 1 = Poor performance status

# 2. Simulate Treatment Assignment (Confounding by Indication)
# Elderly and sicker patients are more likely to receive the treatment (D = 1)
ps_true <- plogis(-2 + 0.05 * age + 1.2 * ecog)
treatment <- rbinom(n, 1, ps_true)

# 3. Simulate Survival Outcome (1-Year Survival)
# True Treatment Effect (Risk Difference) = +15.5% (coefficient = 0.8 in log-odds scale)
y_prob_1 <- plogis(1 - 0.03 * age - 1.0 * ecog + 0.8 * 1)
y_prob_0 <- plogis(1 - 0.03 * age - 1.0 * ecog + 0.8 * 0)
true_ate <- mean(y_prob_1) - mean(y_prob_0)

survival_1yr <- ifelse(treatment == 1,
                       rbinom(n, 1, y_prob_1),
                       rbinom(n, 1, y_prob_0))

df <- data.frame(age, ecog, treatment, survival_1yr)

# Calculate Naive (Unadjusted) Estimate
naive_treated <- mean(df$survival_1yr[df$treatment == 1])
naive_control <- mean(df$survival_1yr[df$treatment == 0])
naive_ate <- naive_treated - naive_control

# Construct baseline descriptive statistics
get_stats <- function(sub_df) {
  n_sub <- nrow(sub_df)
  mean_age <- mean(sub_df$age)
  sd_age <- sd(sub_df$age)
  n_ecog <- sum(sub_df$ecog == 1)
  pct_ecog <- (n_ecog / n_sub) * 100
  n_surv <- sum(sub_df$survival_1yr == 1)
  pct_surv <- (n_surv / n_sub) * 100
  
  c(
    sprintf("%d", n_sub),
    sprintf("%.2f (%.2f)", mean_age, sd_age),
    sprintf("%d (%.1f%%)", n_ecog, pct_ecog),
    sprintf("%d (%.1f%%)", n_surv, pct_surv)
  )
}

desc_tbl <- data.frame(
  Characteristic = c("Sample Size (N)", "Age, Mean (SD)", "ECOG Status: Poor (1), N (%)", "Observed 1-Year Survival, N (%)"),
  "Control (D = 0)" = get_stats(df[df$treatment == 0, ]),
  "Treated (D = 1)" = get_stats(df[df$treatment == 1, ]),
  "Total Cohort" = get_stats(df),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

knitr::kable(desc_tbl, caption = "Baseline Characteristics & Raw Outcomes of the Simulated Cohort", align = "lrrr")
```

---

### Step 1: Initial Estimation of the Outcome Regression ($Q_0$)

We estimate the conditional expectation of the outcome $Y$ given treatment $D$ and baseline covariates $W$:
$$Q_0(D, W) = E[Y \mid D, W]$$

We fit a logistic regression model for $Y$ and generate two counterfactual predicted values for each individual $i$:

::: {.column-margin}
Note: Specifying `family = binomial` in R's `glm()` function defaults to a logistic regression model (Logit Model) to model binary outcome probabilities.
:::

- The predicted outcome under treatment: $Q_0(1, W_i)$
- The predicted outcome under control: $Q_0(0, W_i)$

To prevent boundary logit explosions, the initial outcome predictions are bounded to $[0.005, 0.995]$ (corresponding to the default `alpha = 0.995` threshold).

```{r}
# Fit the initial outcome regression model (Q0)
fit_Q <- glm(survival_1yr ~ treatment + age + ecog, family = binomial, data = df)

# Predict counterfactual outcomes under treatment and control
Q0_0_raw <- predict(fit_Q, newdata = data.frame(treatment = 0, age, ecog), type = "response")
Q0_1_raw <- predict(fit_Q, newdata = data.frame(treatment = 1, age, ecog), type = "response")

# Bound predictions to [0.005, 0.995]
Q0_0 <- pmin(pmax(Q0_0_raw, 0.005), 0.995)
Q0_1 <- pmin(pmax(Q0_1_raw, 0.005), 0.995)

# Factual prediction under actual observed treatment (no counterfactual component).
# Because we only observe outcome Y under the actual treatment A, we must use Q0_A as the baseline offset
# to calculate residuals and fit the fluctuation parameter (epsilon) in Step 4.
Q0_A <- ifelse(treatment == 1, Q0_1, Q0_0)

# Format summary table
q0_summary <- data.frame(
  Metric = c("Mean Prediction", "Minimum Prediction", "Maximum Prediction"),
  "Control (Q0_0)" = c(mean(Q0_0), min(Q0_0), max(Q0_0)),
  "Treatment (Q0_1)" = c(mean(Q0_1), min(Q0_1), max(Q0_1)),
  check.names = FALSE
)
knitr::kable(q0_summary, digits = 4, caption = "Summary of Initial Outcome Regression Predictions (Q0)", align = "lrr")
```

---

### Step 2: Estimation of the Propensity Score ($g_0$)

We estimate the probability of treatment assignment given covariates:
$$g(W) = P(D = 1 \mid W)$$

We fit a logistic regression model for treatment assignment to predict propensity scores, and apply a dynamic lower bound threshold to guarantee that inverse weights do not blow up.

::: {.column-margin}
If not specified, the `tmle` package adaptively calculates the propensity score lower bound based on the sample size:
$$\text{gbound\_lower} = \frac{5}{\sqrt{n} \cdot \ln(n)}$$

For $N=500$, the lower bound is approximately `0.03598`. This bound is applied independently to the propensity scores of the treated ($g(W)$) and control ($1 - g(W)$) groups.
:::


```{r}
# Fit propensity score model
fit_g <- glm(treatment ~ age + ecog, family = binomial, data = df)
g1W_raw <- predict(fit_g, type = "response")

# Calculate dynamic lower bound: 5 / (sqrt(n) * log(n))
gbound_lower <- 5 / (sqrt(n) * log(n))

# Apply bounding independently for g1W and g0W
g1W <- pmax(g1W_raw, gbound_lower)
g0W <- pmax(1 - g1W_raw, gbound_lower)

# Format summary table
g_summary <- data.frame(
  Metric = c("Lower Bound Value", "Mean Propensity Score", "Minimum Propensity Score"),
  "g1W (Treated)" = c(gbound_lower, mean(g1W), min(g1W)),
  "g0W (Control)" = c(gbound_lower, mean(g0W), min(g0W)),
  check.names = FALSE
)
knitr::kable(g_summary, digits = 4, caption = "Summary of Bounded Propensity Scores", align = "lrr")
```

---

### Step 3: Computation of the Clever Covariate ($H$)

We construct the **Clever Covariates** $H_1(W)$ and $H_0(W)$ representing the inverse probability weights adjusted for treatment received:

- For treated: $H_1(W) = \frac{D}{g(W)}$
- For controls: $H_0(W) = \frac{1 - D}{1 - g(W)}$

```{r}
# Calculate clever covariates
H1W <- treatment / g1W
H0W <- (1 - treatment) / g0W

# Display first few rows
h_tbl <- data.frame(
  Patient = 1:5,
  Treatment = treatment[1:5],
  "g1W (Treated)" = g1W[1:5],
  "g0W (Control)" = g0W[1:5],
  "H0W (Clever Control)" = H0W[1:5],
  "H1W (Clever Treated)" = H1W[1:5],
  check.names = FALSE
)
knitr::kable(h_tbl, digits = 4, caption = "Clever Covariates (First 5 Patients)", align = "lrrrrr")
```

---

### Step 4: The Fluctuation (Update) Step

We update the initial outcome prediction $Q_0$ by fitting a logistic regression of the observed outcome $Y$ on the Clever Covariates $H_0(W)$ and $H_1(W)$ without an intercept, using the logit of the initial prediction $Q_0(D, W)$ as a fixed offset:
$$\text{logit}\left(Q_1(D, W)\right) = \text{logit}\left(Q_0(D, W)\right) + \epsilon_0 \cdot H_0(W) + \epsilon_1 \cdot H_1(W)$$

Fitting this regression via Maximum Likelihood estimates the fluctuation parameters $\epsilon_0$ and $\epsilon_1$, which represent the residual bias correction. We then update the counterfactual predictions and bound them.

```{r}
# Fit fluctuation regression (offset-only model)
fit_eps <- glm(survival_1yr ~ -1 + H0W + H1W, offset = qlogis(Q0_A), family = binomial)
eps <- coef(fit_eps)

# Update counterfactual predictions
Q1_0 <- plogis(qlogis(Q0_0) + eps["H0W"] / g0W)
Q1_1 <- plogis(qlogis(Q0_1) + eps["H1W"] / g1W)

# Bound final updated predictions to [0.005, 0.995]
Q1_0 <- pmin(pmax(Q1_0, 0.005), 0.995)
Q1_1 <- pmin(pmax(Q1_1, 0.005), 0.995)

# Format epsilon table
eps_tbl <- data.frame(
  Parameter = c("epsilon_0 (Control, H0W)", "epsilon_1 (Treatment, H1W)"),
  Estimate = c(eps["H0W"], eps["H1W"]),
  "Std. Error" = summary(fit_eps)$coefficients[, "Std. Error"],
  "z-value" = summary(fit_eps)$coefficients[, "z value"],
  "p-value" = summary(fit_eps)$coefficients[, "Pr(>|z|)"],
  check.names = FALSE
)
knitr::kable(eps_tbl, digits = 6, caption = "Estimated Fluctuation Coefficients (Epsilons)", align = "lrrrr")
```

---

### Step 5: Parameter Estimation via Substitution

Finally, we calculate the doubly robust ATE by averaging the updated counterfactual predictions across the sample:
$$\hat{ATE}_{TMLE} = \frac{1}{n} \sum_{i=1}^n \left[ Q_1(1, W_i) - Q_1(0, W_i) \right]$$

```{r}
# Calculate manual ATE
ate_manual <- mean(Q1_1 - Q1_0)
```

The manual step-by-step TMLE yields a Risk Difference (ATE) of **`r sprintf("%.6f", ate_manual)`**.

---

### Verification: Comparing Manual TMLE with the Package Output

To confirm the accuracy of our step-by-step manual implementation, we fit the ATE using the `tmle` package, turning off cross-validation (`cvQinit = FALSE`) to align the initial outcome estimation.

```{r}
# Fit Package TMLE (without cross-validation for direct verification)
tmle_fit_verify <- tmle(
  Y             = df$survival_1yr,
  A             = df$treatment,
  W             = df[, c("age", "ecog")],
  Q.SL.library  = "SL.glm",
  g.SL.library  = "SL.glm",
  cvQinit       = FALSE,
  family        = "binomial"
)

# Format verification comparison table
comparison_tbl <- data.frame(
  Parameter = c("Epsilon_0 (Control Coefficient)", "Epsilon_1 (Treatment Coefficient)", "ATE Point Estimate"),
  "Manual Step-by-Step" = c(eps["H0W"], eps["H1W"], ate_manual),
  "tmle Package Output" = c(tmle_fit_verify$epsilon["H0W"], tmle_fit_verify$epsilon["H1W"], tmle_fit_verify$estimates$ATE$psi),
  "Absolute Difference" = c(
    abs(eps["H0W"] - tmle_fit_verify$epsilon["H0W"]),
    abs(eps["H1W"] - tmle_fit_verify$epsilon["H1W"]),
    abs(ate_manual - tmle_fit_verify$estimates$ATE$psi)
  ),
  check.names = FALSE
)
knitr::kable(comparison_tbl, digits = 8, caption = "Validation of Manual TMLE Against tmle Package", align = "lrrr")
```

As demonstrated, the manual step-by-step implementation matches the `tmle` package output exactly to machine precision (absolute difference $\approx 0$), validating the underlying mechanics of the TMLE algorithm.

---

## Production R Implementation Pipeline (Package-level Fit & Plots)

While the manual walkthrough utilizes simple parametric models to illustrate the underlying mechanics, HTA submissions require production-grade models using machine learning ensembles and cross-validation to minimize misspecification bias.

Here, we fit the final ATE using the `tmle` package with cross-validation (`cvQinit = TRUE`) and SuperLearner ensembles containing GLM and Mean libraries.

```{r}
#| message: false
#| warning: false

# Fit production-grade TMLE and suppress package-level output
sl_lib <- c("SL.glm", "SL.mean")

tmle_fit <- suppressMessages(tmle(
  Y             = df$survival_1yr,
  A             = df$treatment,
  W             = df[, c("age", "ecog")],
  Q.SL.library  = sl_lib,
  g.SL.library  = sl_lib,
  family        = "binomial"
))

# Extract results
tmle_est <- tmle_fit$estimates$ATE$psi
tmle_se  <- sqrt(tmle_fit$estimates$ATE$var.psi)
tmle_ci  <- tmle_fit$estimates$ATE$CI
tmle_pval <- tmle_fit$estimates$ATE$pvalue
epsilon  <- tmle_fit$epsilon

# Format TMLE parameter table
tmle_tbl <- data.frame(
  Parameter = c("Risk under Control (EY0)", "Risk under Exposure (EY1)", "Risk Difference (ATE)"),
  Estimate = c(tmle_fit$estimates$EY0$psi, tmle_fit$estimates$EY1$psi, tmle_est),
  "Std. Error" = c(sqrt(tmle_fit$estimates$EY0$var.psi), sqrt(tmle_fit$estimates$EY1$var.psi), tmle_se),
  "95% LCL" = c(tmle_fit$estimates$EY0$CI[1], tmle_fit$estimates$EY1$CI[1], tmle_ci[1]),
  "95% UCL" = c(tmle_fit$estimates$EY0$CI[2], tmle_fit$estimates$EY1$CI[2], tmle_ci[2]),
  "p-value" = c(NA, NA, tmle_pval),
  check.names = FALSE
)

# Round values for display
tmle_tbl[, 2:5] <- round(tmle_tbl[, 2:5], 4)
tmle_tbl[3, 6] <- format.pval(tmle_pval, eps = 0.001, digits = 3)

knitr::kable(tmle_tbl, caption = "Production TMLE Results with SuperLearner & Cross-Validation", align = "lrrrrr")
```

::: {.callout-note}
## Interpretation of the Fluctuation Parameters
The estimated fluctuation parameters under the production fit are $\epsilon_0 = $ **`r round(epsilon[1], 6)`** (Control) and $\epsilon_1 = $ **`r round(epsilon[2], 6)`** (Treated). These non-zero coefficients represent the targeted adjustments applied during the fluctuation step to remove residual treatment selection bias.
:::

### Sensitivity Analysis: E-Value Calculation

To evaluate sensitivity to unmeasured confounding, we transform the Risk Difference obtained from TMLE to the Relative Risk (RR) scale and calculate the corresponding E-value.

```{r}
# 1. Transform Risk Difference to Relative Risk
ey0_est <- tmle_fit$estimates$EY0$psi
rr_est  <- (ey0_est + tmle_est) / ey0_est
rr_lo   <- (ey0_est + tmle_ci[1]) / ey0_est
rr_hi   <- (ey0_est + tmle_ci[2]) / ey0_est

# 2. Compute E-value
ev <- evalues.RR(est = rr_est, lo = rr_lo, hi = rr_hi)
ev_df <- as.data.frame(ev)

# 3. Format E-value table
ev_tbl <- data.frame(
  "Metric / Parameter" = c("Relative Risk of Treatment (RR)", "E-value (Treatment Effect Bound)"),
  "Point Estimate" = c(sprintf("%.3f", ev_df["RR", "point"]), sprintf("%.3f", ev_df["E-values", "point"])),
  "95% CI Lower Bound" = c(sprintf("%.3f", ev_df["RR", "lower"]), sprintf("%.3f", ev_df["E-values", "lower"])),
  "95% CI Upper Bound" = c(sprintf("%.3f", ev_df["RR", "upper"]), "N/A"),
  check.names = FALSE
)

knitr::kable(ev_tbl, caption = "E-Value Analysis for Unmeasured Confounding Sensitivity", align = "lrrr")
```

The point estimate E-value of **`r round(ev_df["E-values", "point"], 2)`** indicates that an unmeasured confounder would need to have a Relative Risk association of at least **`r round(ev_df["E-values", "point"], 2)`**-fold with both treatment assignment and survival to fully explain away the observed causal survival benefit.

### E-Value Contour Visualization

The contour curve illustrates the required strength of association between the unmeasured confounder and the treatment (x-axis) and outcome (y-axis) to reduce the observed effect to the null.

```{r}
#| label: evalue-plot
#| fig-align: center
#| fig-width: 6
#| fig-height: 5

plot(ev, type = "line", 
     main = "E-value Contour: Sensitivity to Unmeasured Confounding")
```

---

## Causal Evidence Reconciliation

We reconcile the point estimates and standard errors across our observational causal pipeline compared to the simulated population ground truth.

```{r}
# Fit IPTW for comparison (from Page 2 workflow)
W <- weightit(treatment ~ age + ecog, data = df, method = "ps", estimand = "ATE")
fit_iptw <- lm_weightit(survival_1yr ~ treatment, data = df, weightit = W)
iptw_summary <- summary(fit_iptw)
iptw_est <- iptw_summary$coefficients["treatment", "Estimate"]
iptw_se  <- iptw_summary$coefficients["treatment", "Std. Error"]

# Combine into reconciliation matrix
reconciliation_tbl <- data.frame(
  Estimator = c("Ground Truth (Omniscient)", "Naive (Unadjusted)", "IPTW (Weighting Only)", "TMLE (Doubly Robust)"),
  "ATE Point Estimate" = c(true_ate, naive_ate, iptw_est, tmle_est),
  "Standard Error" = c(NA, NA, iptw_se, tmle_se),
  "95% LCL" = c(NA, NA, iptw_est - 1.96 * iptw_se, tmle_ci[1]),
  "95% UCL" = c(NA, NA, iptw_est + 1.96 * iptw_se, tmle_ci[2]),
  "Bias vs. Truth" = c(0.0000, naive_ate - true_ate, iptw_est - true_ate, tmle_est - true_ate),
  check.names = FALSE
)

# Format values for display
reconciliation_tbl[, 2:6] <- round(reconciliation_tbl[, 2:6], 4)
reconciliation_tbl[is.na(reconciliation_tbl)] <- "N/A"

knitr::kable(reconciliation_tbl, caption = "Causal Evidence Reconciliation Across Estimators", align = "lrrrrr")
```

---

## EAG Defense Strategy

When defending causal estimates derived via TMLE and E-value sensitivity analysis during EAG reviews, modellers should leverage the following arguments:

### Justifying TMLE Over Parametric Regression

> "Rather than relying on arbitrary functional forms that assume linear log-odds relations for confounding covariates, we implemented TMLE coupled with the SuperLearner ensemble. This allows the data to non-parametrically determine the correct functional relations, minimizing model misspecification bias. 
> 
> Furthermore, TMLE is a substitution estimator, meaning that unlike influence function-based estimators (like AIPW) which can yield survival predictions outside the valid $[0, 1]$ interval, TMLE is mathematically restricted to remain within the physical bounds of probability space. This ensures clinical plausibility in all extrapolated subgroups."

### Auditing Sensitivity via E-Value

> "The E-value of `r round(ev_df["E-values", "point"], 2)` demonstrates that our causal treatment effect is robust against unmeasured confounding. To invalidate this survival difference, an unmeasured baseline covariate would need to increase both the likelihood of treatment and the likelihood of survival by at least `r round(ev_df["E-values", "point"], 2)`-fold. 
> 
> In our adjusted models, the strongest measured clinical confounders (such as poor ECOG performance status) only display hazard associations of approximately 1.8-fold. It is highly implausible that a missing, unmeasured confounder of greater magnitude exists that is completely uncorrelated with the clinical indicators adjusted for in this analysis."
