---
title: "Survival Analysis & Immortal Time Bias in Real-World Evidence"
subtitle: "Target Trial Emulation, Landmark Analysis, and Time-Varying Cox Regression"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "R Code"
    number-sections: true
    reference-location: margin
---

*Methodological guidelines and implementation templates for survival analysis, delayed treatment initiation, and target trial emulation in health technology assessment.*

## Introduction: Causal Inference in Time-to-Event Data

In health technology assessment (HTA), clinical endpoints are frequently time-to-event outcomes, such as Overall Survival (OS) or Progression-Free Survival (PFS). When randomized controlled trials (RCTs) are unfeasible or ethical constraints limit comparator arms, observational real-world data (RWD) is analyzed to construct External Control Arms (ECAs). 

Unlike binary outcomes where logistic regression or propensity score matching are applied statically, survival data introduces a critical dimension: **Time**. 

If time-to-event analysis is not bounded by strict causal constraints, observational analyses can yield severe selection biases. A primary methodological challenge in retrospective registries is **Immortal Time Bias (ITB)**. 

### The Coffee & Heart Disease Paradox (Reverse Causality)

To understand why simple observational survival analysis fails, consider the "Coffee & Heart Disease Paradox":

- **The Observation**: A cross-sectional analysis or simple regression shows that people who drink coffee have significantly lower rates of heart disease (e.g., Hazard Ratio < 1.0).
- **The Reality**: This is driven by reverse causality. Patients who *already* have heart problems or are at high risk are often advised by their doctors to quit coffee.
- **The Trap**: If we run a regression on cross-sectional data, we see "Coffee = Health." 

To resolve this paradox, we cannot rely on simple regression adjustments; we must emulate a randomized trial to establish a clear chronological sequence of events.

---

## The Target Trial Emulation (TTE) Framework

Target Trial Emulation is a structured design framework developed by Miguel Hernán and colleagues to guide the design and analysis of observational studies. The core principle is that **we should design our observational study to mimic a hypothetical randomized trial as closely as possible.**

### The 7 Pillars of Target Trial Emulation

To translate observational data into robust evidence, researchers must define a protocol addressing seven protocol components:

| Pillar | NICE RWE Framework Requirement | Application to Survival Cohorts |
| :--- | :--- | :--- |
| **1. Eligibility** | Define clear inclusion/exclusion criteria at baseline. | Patients must meet all criteria at Time Zero ($T_0$), ensuring no baseline selection bias. |
| **2. Treatment Strategy** | Specify the interventions being compared. | **Arm A:** Initiate active treatment within a defined window. **Arm B:** Initiate Standard of Care (SOC) / no treatment. |
| **3. Assignment** | Define how treatment groups are formed. | Emulate randomization at $T_0$ using Propensity Scores (IPTW or matching) based on baseline covariates. |
| **4. Follow-up** | Align follow-up start exactly with $T_0$. | Clock starts ticking for all patients at $T_0$. Ends at the event, death, censoring, or study end. |
| **5. Outcome** | Define a standardized clinical endpoint. | First occurrence of the event (e.g., death, progression) during the follow-up period. |
| **6. Causal Contrast** | Specify the effect of interest. | **Intention-to-Treat (ITT):** Effect of treatment assignment at $T_0$, regardless of later compliance. |
| **7. Analysis Plan** | Specify the statistical estimator. | Cox Proportional Hazards model or parametric survival models, adjusted for baseline confounding. |

### Deep Dive: The New User Design

A fundamental rule of TTE is the **New User Design**. In observational studies, researchers often mistakenly compare "prevalent users" (patients already on the drug for some time) with "never users." This introduces three major issues:

- **Survivor Bias**: Prevalent users are, by definition, "survivors" who have tolerated the drug and survived long enough to be observed in the database. Patients who died or discontinued the drug early due to adverse events are systematically excluded from the treated group, creating a false survival advantage.
- **Baseline Contamination**: To adjust for confounding, we must measure covariates *before* treatment starts. For prevalent users, their baseline measurements (e.g., blood pressure, lab values) are already affected by the treatment, making it impossible to measure their true pre-treatment state.
- **$T_0$ Synchronization**: We cannot align the start of follow-up with the start of treatment if treatment began years before the observation period.

A New User Design enforces that patients are only entered into the cohort at the exact moment they **initiate** the treatment of interest.

### Deep Dive: Intention-to-Treat (ITT) in RWD (Cloning & Censoring)

In randomized trials, the ITT principle analyzes patients in the group they were randomized to, regardless of whether they complied. In retrospective RWD, we often face a "grace period"—a window of time (e.g., 30 or 60 days) after diagnosis during which patients are allowed to initiate treatment.

If we assign patients to the "treated" group because they initiated treatment on day 30, we violate the ITT principle by using "future knowledge" at $T_0$. To solve this, TTE uses **Cloning**:

1. **Cloning at $T_0$**: At the moment of eligibility ($T_0$), we create two identical copies of each patient. One copy is assigned to Arm A (Initiate Treatment), and one copy is assigned to Arm B (Standard of Care).
2. **Follow-up**: Both clones begin follow-up at $T_0$.
3. **Censoring**: If a patient assigned to Arm B initiates treatment on day 30, their Arm B clone is **censored** at day 30 (because they violated the "no treatment" protocol). The Arm A clone continues to be followed.
4. **IPCW Adjustment**: To correct for selection bias introduced by this artificial censoring, we apply **Inverse Probability of Censoring Weights (IPCW)**, which upweights the remaining uncensored clones who share similar baseline characteristics.

### Auditor's View: Why TTE > Simple Regression

NICE prefers TTE because it enforces **Transparency**. 

In a simple Cox regression, you often see:
$$\text{Surv}(\text{time}, \text{event}) \sim \text{Treatment} + \text{Confounders}$$

An auditor cannot see:

- When was the treatment started?
- Were the confounders measured *before* or *after* the drug was given?
- Is there a "Healthy User Bias"?

By using the **DataSAT** tool (Data Suitability Assessment Tool), NICE auditors look for **Provenance**—the clear lineage of how a patient moved from a raw database record to a study row. TTE forces the researcher to write a "Trial Protocol" *before* touching the data, making it obvious where $T_0$ is. If an auditor sees that you measured covariates after the drug was given and used them as baseline, the TTE framework will flag this as a "Post-baseline Bias" immediately.

---

## HTA Case Study: Drug X (METex14) Case Study

To see how these principles apply to a real submission, we examine the case of **Drug X**, a precision oncology therapy for METex14-mutated non-small cell lung cancer (NSCLC).

### Case Overview: The Drug X Conflict

- **The Innovation**: Due to the rarity of the METex14 mutation and ethical constraints, only a **Single-arm Trial (N=40)** was conducted for Drug X.
- **The Objection**: The Evidence Assessment Group (EAG) flagged the submission as "high risk" because there was no randomized head-to-head comparison. Natural history for METex14 is aggressive, and simple benchmarking is insufficient.
- **The Strategy**: We leverage the UK Hospital Episode Statistics (HES) database to construct an External Control Arm (ECA) of 200 patients, using the Target Trial Emulation (TTE) framework to satisfy NICE's demand for rigorous causal inference.

### Step 1: Data Suitability Audit (DataSAT)

According to Sections 3.1–3.3 of the NICE RWE Framework, we must prove the data is fit for purpose across three core dimensions:

| Dimension | HTA Audit Response (HES Database) | NICE Framework Requirement |
| :--- | :--- | :--- |
| **Provenance** | Data sourced from NHS Digital with a clear audit trail from ICD-10 coding to the final analytical dataset. | Evaluate the **origin and trustworthiness** of data sources, methods of collection, and governance. (Section 3.1) |
| **Quality** | METex14 status confirmed via secondary linkage to SACT and genomic records; completeness checked for key covariates. | Assess the **completeness and accuracy** of eligibility criteria, exposures, and outcomes. (Section 3.2) |
| **Relevance** | HES covers 99% of NHS hospital activity, matching the SOC pathways in the UK. | Ensure findings **generalise to the target population** and clinical care pathways in the NHS. (Section 3.3) |

### Defining $T_0$ (Zero-time Alignment)

To avoid Immortal Time Bias, we align $T_0$ as the moment a patient meets all eligibility criteria **AND** is assigned to treatment (Section 5.2).

#### Why Immortal Time Bias is a "Statistical Killer"

1. **Logic Paradox**: In RWD studies, patients from diagnosis to actual treatment initiation (Drug X) face delays (e.g., 1 month). If follow-up begins at diagnosis, but grouping is based on whether they ever received treatment, the period before receiving the drug is **immortal**—if the patient died on Day 15, they would never have lived long enough to receive the drug, and would automatically be classified into the control group.
2. **Case Simulation**:

   - **Patient A**: Diagnosed Jan 1, scheduled for Drug X, dies Jan 15 (unexposed). Classified as Control.
   - **Patient B**: Diagnosed Jan 1, initiates Drug X Feb 1, survives until June.
   - **Result**: The Drug Y/X group artificially filters out those who die before drug administration, inflating the active group's survival.
3. **TTE Surgery**: The clock must start ticking exactly at treatment initiation, or grace periods must be modeled using cloning.

#### The Control Arm $T_0$

Where does the control arm's $T_0$ start if they never receive the active treatment?

- **Synchronicity Principle**: $T_0$ is the moment the patient meets all clinical eligibility criteria (e.g., METex14 diagnosis + ECOG status).
- **Timeline Alignment**:
  - *Incorrect*: Treated group starts at "first drug administration", while controls start at "diagnosis". This gives the treated group an implicit survival benefit.
  - *Correct TTE Approach*: Both groups start follow-up at the exact date they meet all eligibility criteria (the clinical decision point, or clinical equipoise). 

#### Handling the Grace Period

In RWD, a patient has a "window" of time (grace period) to initiate treatment (e.g., 30 days after diagnosis). 

- If a patient initiates treatment on Day 20, they contribute 20 days of "unexposed" person-time to the control group.
- Using cloning, we create two clones at diagnosis. The "Control Clone" is followed until Day 20, where it is censored because the patient initiated treatment. The "Treated Clone" starts contributing treated time from Day 20 onwards.

### Step 2: Cloning & Censoring Mechanics

To maintain ITT principles:

- **Anti-Future-Knowledge**: We cannot look ahead to see what the patient did. We clone them at $T_0$ into both paths.
- **Artificial Censoring**: If the patient's actual path deviates from a clone's assigned arm, that clone is censored.
- **IPCW Weighting**: Because censoring might be non-random, we estimate the probability of being censored at time $t$ based on baseline covariates and apply Inverse Probability of Censoring Weights to the remaining patients, dynamically restoring the balance.

---

## Immortal Time Bias (ITB): Theory & Landmark Analysis

### The Meditation & Longevity Paradox

Consider a study evaluating whether a 12-week intensive meditation course increases overall survival (OS):

- **Cohort**: 1,000 participants followed for 5 years.
- **Exposure Group ($Z=1$)**: Completed the full 12-week program.
- **Control Group ($Z=0$)**: Did not complete the program (includes those who died or dropped out before week 12).
- **The Result**: Meditators show a Hazard Ratio ($HR$) of 0.30 (70% risk reduction).
- **The Error**: Since a patient had to survive 12 weeks to complete the program, the meditation group is guaranteed 0% mortality during the first 12 weeks. Sicker patients who died in weeks 1–11 were automatically allocated to the control group.

### Violating the "Non-anticipatory Constraint"

Mathematically, let $\mathcal{F}_t$ be the filtration (information history) up to time $t$. A treatment assignment indicator $Z_t$ must be adapted to the filtration:
$$Z_t \in \mathcal{F}_t$$

If we define a static indicator $Z$ based on whether the event occurred in the future:
$$Z = f(Y_{t \in [0, 12]})$$
we condition on a post-baseline outcome $Y$. This violates the non-anticipatory constraint, leading to "time travel" in the statistical model.

### Landmark Analysis

Landmark Analysis is a classical approach to resolve this:

1. **Select Landmark Time ($T_L$)**: E.g., Week 12.
2. **Exclude Early Events**: Exclude all patients who died or were censored before $T_L$.
3. **Re-classify**: Classify survivors at $T_L$ based on their exposure status at $T_L$.
4. **Follow-up**: Set $T_L$ as the new baseline ($T_{start} = T_L$) and follow patients from that point.

#### HTA Audit of Landmark Analysis

- **Pros**: Simple, highly intuitive, and removes the immortal period.
- **Cons**: 
  - **Information Loss**: Discards all events and person-time before $T_L$. If the drug has high early toxicity, this signal is completely lost.
  - **Selection Bias**: The populations at $T_L$ may no longer be balanced, as sicker patients in the control group may have died, while sicker patients in the treated group were protected by the "immortal" definition.

---

## Time-Dependent Cox Model & HTA Practice

The time-dependent Cox model is the industry standard for auditing and correcting immortal time bias in HTA submissions.

### Counting Process Data Structure

Rather than one row per patient, the counting process format splits each patient's follow-up into multiple intervals `(tstart, tstop]` where treatment status remains constant.

#### Restructuring Example (Patient #101)
Patient #101 has a treatment initiation lag of 60 days and dies on Day 500. We split this patient's record into two lines:

| Patient ID | Start (Day) | Stop (Day) | Status (Event) | Treatment ($Z$) |
| :--- | :--- | :--- | :--- | :--- |
| #101 | 0 | 60 | 0 (Alive) | 0 (Unexposed) |
| #101 | 60 | 500 | 1 (Death) | 1 (Exposed) |

At Day 60, the patient is right-censored in the control group and enters the treated group.

### Landmark vs. Time-Dependent Cox

| Dimension | Time-dependent Cox | Landmark Analysis |
| :--- | :--- | :--- |
| **Data Retention** | Retains all patients and events from Day 0. | Discards patients/events prior to Landmark $T_L$. |
| **Causal Question** | Causal effect starting from Day 0. | Conditional effect given survival to $T_L$. |
| **Early Toxicity** | Captures early adverse event signals. | Masked if events occur before $T_L$. |
| **HTA Audit Preference** | Preferred for primary analysis. | Used primarily as a sensitivity analysis. |

---

## R Simulation: Validation of Estimators

We simulate a cohort of 500 patients where treatment initiation is delayed due to an administrative "Approval Lag."

### Simulating the Cohort and Biased Analysis

We simulate a true treatment Hazard Ratio of **0.70** (30% risk reduction). Age and ECOG status act as baseline confounders.

```{r}
#| message: false
#| warning: false

library(survival)
library(EValue)
set.seed(42)
n <- 500

# Helper function to render Cox model summaries as kable tables
show_cox <- function(fit, caption) {
  sum_fit <- summary(fit)
  coefs <- sum_fit$coefficients
  cis <- sum_fit$conf.int
  tbl <- data.frame(
    Covariate = rownames(coefs),
    `Hazard Ratio` = round(coefs[, 2], 4),
    `Log HR (coef)` = round(coefs[, 1], 4),
    `Std Error` = round(coefs[, 3], 4),
    `z-value` = round(coefs[, 4], 3),
    `p-value` = format.pval(coefs[, 5], digits = 4),
    `95% LCL` = round(cis[, 3], 4),
    `95% UCL` = round(cis[, 4], 4),
    check.names = FALSE
  )
  knitr::kable(tbl, caption = caption, align = "lrrrrrrr")
}

# 1. Simulate Baseline Confounders
age <- rnorm(n, mean = 60, sd = 10)
ecog <- rbinom(n, 1, 0.4)

# 2. Simulate Treatment Selection (Confounding by Indication)
# Sicker (ECOG=1) and older patients are more likely to receive treatment
ps_prob <- plogis(-1.5 + 0.02 * age + 0.5 * ecog)
z_treatment <- rbinom(n, 1, ps_prob)

# For those treated, add a random initiation delay between 30 and 120 days
t_init <- ifelse(z_treatment == 1, runif(n, 30, 120), Inf)

# 3. Simulate Survival Times (True treatment HR = 0.70)
true_hr <- 0.70
lambda_base <- 0.001 * exp(0.02 * age + 0.5 * ecog)

u <- runif(n)
log_u <- -log(u)

t_event <- numeric(n)
for(i in 1:n) {
  l0 <- lambda_base[i]
  l1 <- l0 * true_hr
  t_tr <- t_init[i]
  
  if (log_u[i] <= l0 * t_tr) {
    t_event[i] <- log_u[i] / l0
  } else {
    t_event[i] <- t_tr + (log_u[i] - l0 * t_tr) / l1
  }
}

# Censoring between 150 and 730 days
t_censor <- runif(n, 150, 730)
time <- pmin(t_event, t_censor)
event <- as.integer(t_event <= t_censor)

df <- data.frame(id = 1:n, age, ecog, z_treatment, t_init, time, event)

# Determine treatment status based on actual follow-up (Biased approach)
df$ever_treated <- ifelse(df$t_init < df$time, 1, 0)

# --- Biased Analysis: Static Cox Model ---
fit_biased <- coxph(Surv(time, event) ~ ever_treated + age + ecog, data = df)
show_cox(fit_biased, "Biased Static Cox Proportional Hazards Model Results (Immortal Time Unadjusted)")
```

The biased static Cox model overestimates drug efficacy due to the unadjusted immortal time.

### Method 1: Landmark Analysis (Day 90)

We apply Landmark Analysis at Day 90:

```{r}
t_landmark <- 90

# Exclude patients who died or were censored before Day 90
df_landmark <- df[df$time > t_landmark, ]

# Classify status at Landmark
df_landmark$treated_at_landmark <- ifelse(df_landmark$t_init <= t_landmark, 1, 0)

# Shift time
df_landmark$time_shifted <- df_landmark$time - t_landmark

fit_landmark <- coxph(Surv(time_shifted, event) ~ treated_at_landmark + age + ecog, data = df_landmark)
show_cox(fit_landmark, "Landmark Analysis (Day 90) Cox Proportional Hazards Model Results")
```

### Method 2: Time-Varying Cox Model

We restructure the dataset into a counting process format using `tmerge()` and fit the time-varying Cox model:

```{r}
# 1. Initialize dataset with the primary event time
df_tv <- tmerge(df, df, id = id, event = event(time, event))

# 2. Merge the time-dependent treatment initiation event
df_tv <- tmerge(df_tv, df, id = id, treatment_status = tdc(t_init))

# Display the split intervals for Patient ID 1 to 3 in table format
knitr::kable(
  head(df_tv[df_tv$id %in% 1:3, c("id", "tstart", "tstop", "event", "treatment_status")]),
  caption = "Restructured Counting Process Intervals (Patient ID 1-3)",
  align = "c"
)
```

As shown, if a patient receives treatment at a delayed time, their follow-up is split:

- **Interval 1 `[0, t_init]`**: `treatment_status = 0` (contributes person-time to the control risk set).
- **Interval 2 `[t_init, time]`**: `treatment_status = 1` (contributes person-time to the treated risk set after the first infusion).

We fit the Cox model on the split intervals:

```{r}
# Fit the time-varying Cox model
fit_tv <- coxph(Surv(tstart, tstop, event) ~ treatment_status + age + ecog, data = df_tv)
show_cox(fit_tv, "Time-Varying Cox Proportional Hazards Model Results (Immortal Time Corrected)")
```

The time-varying Cox model successfully recovers the true treatment effect ($HR = 0.70$) without discarding data.

---

## Causal Evidence Reconciliation

The table below summarizes the estimates across all three models compared to the true parameter.

```{r}
# Extract statistics
hr_biased  <- exp(coef(fit_biased)["ever_treated"])
se_biased  <- summary(fit_biased)$coefficients["ever_treated", "se(coef)"]

hr_land    <- exp(coef(fit_landmark)["treated_at_landmark"])
se_land    <- summary(fit_landmark)$coefficients["treated_at_landmark", "se(coef)"]

hr_tv      <- exp(coef(fit_tv)["treatment_status"])
se_tv      <- summary(fit_tv)$coefficients["treatment_status", "se(coef)"]

# Combine into reconciliation matrix
reconciliation_tbl <- data.frame(
  Estimator = c("Ground Truth (True Parameter)", "Biased Model (Static Baseline)", "Landmark Analysis (Day 90)", "Time-Varying Cox Model"),
  "Hazard Ratio (HR)" = c(true_hr, hr_biased, hr_land, hr_tv),
  "Standard Error" = c(NA, se_biased, se_land, se_tv),
  "95% LCL" = c(NA, exp(confint(fit_biased)["ever_treated", 1]), exp(confint(fit_landmark)["treated_at_landmark", 1]), exp(confint(fit_tv)["treatment_status", 1])),
  "95% UCL" = c(NA, exp(confint(fit_biased)["ever_treated", 2]), exp(confint(fit_landmark)["treated_at_landmark", 2]), exp(confint(fit_tv)["treatment_status", 2])),
  "Bias vs. Truth (Log HR)" = c(0.0000, coef(fit_biased)["ever_treated"] - log(true_hr), coef(fit_landmark)["treated_at_landmark"] - log(true_hr), coef(fit_tv)["treatment_status"] - log(true_hr)),
  check.names = FALSE
)

# Format values for display
reconciliation_tbl[, 2:6] <- round(reconciliation_tbl[, 2:6], 4)
reconciliation_tbl[is.na(reconciliation_tbl)] <- "N/A"

knitr::kable(reconciliation_tbl, caption = "Survival Parameter Reconciliation Across Estimators", align = "lrrrrr")
```

---

## Quantitative Bias Analysis: E-Value Sensitivity Analysis

When submitting RWE to HTA bodies, we must audit the robustness of our findings against unmeasured confounding. We compute the E-value for the time-varying Hazard Ratio ($HR = 0.693$).

### E-Value Math & Formula Setup

The E-value is defined as the minimum strength of association on the risk ratio scale that an unmeasured confounder ($U$) must have with both:
1. The treatment assignment ($Z$)
2. The clinical outcome ($Y$)

to fully explain away the observed treatment-outcome association.

The formula depends on the cumulative incidence of the outcome:

1. **Rare Outcome Assumption** (cumulative incidence < 15%):

   The Hazard Ratio approximates the Relative Risk ($RR \approx HR$). For protective effects ($HR < 1.0$), we first invert the HR ($HR^* = 1 / HR$) to get a value $> 1.0$:
   $$\text{E-value} = HR^* + \sqrt{HR^* \cdot (HR^* - 1)}$$

2. **Common Outcome Correction** (cumulative incidence > 15%):

   The rare outcome assumption is violated, and we must apply the square root approximation:
   $$RR \approx \sqrt{HR}$$
   The corresponding E-value is calculated as:
   $$\text{E-value} = \sqrt{HR^*} + \sqrt{\sqrt{HR^*} \cdot (\sqrt{HR^*} - 1)}$$

In our cohort, $379$ out of $500$ patients ($75.8\%$) experienced the event, meaning the outcome is common. We must apply the common outcome correction.

```{r}
# 1. Extract HR and confidence limits from the time-varying Cox model
hr_est_val <- exp(coef(fit_tv)["treatment_status"])
hr_lcl_val <- exp(confint(fit_tv)["treatment_status", 1])
hr_ucl_val <- exp(confint(fit_tv)["treatment_status", 2])

# 2. Compute E-value (common outcome correction: rare = FALSE)
ev_survival <- evalues.HR(est = hr_est_val, lo = hr_lcl_val, hi = hr_ucl_val, rare = FALSE)
ev_survival_df <- as.data.frame(ev_survival)

# 3. Format E-value table
ev_survival_tbl <- data.frame(
  "Metric / Parameter" = c("Hazard Ratio (HR)", "E-value (Treatment Effect Bound)"),
  "Point Estimate" = c(sprintf("%.3f", ev_survival_df["HR", "point"]), sprintf("%.3f", ev_survival_df["E-values", "point"])),
  "95% CI Lower Bound" = c(sprintf("%.3f", ev_survival_df["HR", "lower"]), sprintf("%.3f", ev_survival_df["E-values", "lower"])),
  "95% CI Upper Bound" = c(sprintf("%.3f", ev_survival_df["HR", "upper"]), "N/A"),
  check.names = FALSE
)

knitr::kable(ev_survival_tbl, caption = "E-Value Analysis for Survival Sensitivity to Unmeasured Confounding", align = "lrrr")
```

The point estimate E-value of **`r round(ev_survival_df["E-values", "point"], 3)`** means that an unmeasured confounder would need to increase the probability of treatment assignment by **`r round(ev_survival_df["E-values", "point"], 3)`-fold** and increase the hazard of death by **`r round(ev_survival_df["E-values", "point"], 3)`-fold** to fully explain away the observed survival benefit.

### Algebraic Derivation of the E-Value (VanderWeele & Ding 2016)

To understand the mathematical strength of the E-value, we trace its algebraic derivation from the **Bias Factor** model.

Let $U$ be a binary unmeasured confounder. The observed relative risk $RR_{obs}$ is related to the true relative risk $RR_{true}$ by:
$$RR_{obs} = RR_{true} \times \text{Bias}$$

where the Bias Factor is defined as:
$$\text{Bias} = \frac{RR_{ZU} \times RR_{UY}}{RR_{ZU} + RR_{UY} - 1}$$

Here:

- $RR_{ZU}$ is the risk ratio associating the treatment $Z$ with the unmeasured confounder $U$.
- $RR_{UY}$ is the risk ratio associating the unmeasured confounder $U$ with the outcome $Y$.

To explain away an observed effect (i.e., to show that the true effect is null, $RR_{true} = 1.0$), the Bias Factor must equal or exceed the observed relative risk:
$$\text{Bias} \ge RR_{obs}$$

To find the minimum strength of association required for both relationships, we define a single threshold $E = RR_{ZU} = RR_{UY}$. Substituting $E$ into the Bias formula:
$$\frac{E^2}{2E - 1} \ge RR_{obs}$$

This can be rearranged into a quadratic inequality:
$$E^2 - 2RR_{obs} \cdot E + RR_{obs} \ge 0$$

Solving the quadratic equation $E^2 - 2RR_{obs} \cdot E + RR_{obs} = 0$ using the quadratic formula:
$$E = \frac{2RR_{obs} \pm \sqrt{4RR_{obs}^2 - 4RR_{obs}}}{2} = RR_{obs} + \sqrt{RR_{obs}(RR_{obs} - 1)}$$

This is the exact mathematical derivation of the E-value formula. It represents the absolute minimum joint relationship required for an unmeasured confounder to explain away the observed treatment effect.

---

## HTA Audit & Interview Guide

### Interview Audit Checklist

When auditing a manufacturer's RWE survival submission (or preparing for an HTA interview), follow these three key checks:

1. **Zero-time Alignment Check**:

   - *Question*: Identify the exact calendar date of $T_0$ for both cohorts. Is $T_0$ defined by the clinical event (diagnosis) or the administrative event (first prescription)?
   - *Audit Rule*: Follow-up must start for both cohorts when they meet the eligibility criteria. If the treated group follow-up starts at drug initiation and controls start at diagnosis, the analysis is invalid.

2. **Immortal Time Bias Screen**:

   - *Question*: Check the initial segment of the Kaplan-Meier curve. Does the treated arm show a "zero-risk" flat line at the start of follow-up?
   - *Audit Rule*: A flat line on the treatment curve indicates that patients who died early were excluded from the treated group, indicating Immortal Time Bias.

3. **Statistical Correction Audit**:

   - *Question*: If a treatment initiation lag exists, was it addressed using a time-varying covariate or a Landmark approach?
   - *Audit Rule*: Standard Cox models using a static baseline treatment indicator are statistically invalid in the presence of treatment initiation delays.

### Expert HTA Articulation

Use the following professional statements to articulate survival analysis defense strategies during HTA reviews or interviews:

- **Justifying Time-Varying Cox Models Over Static Models**:
  > "To prevent the introduction of lead-time and immortal time biases, we did not define treatment statically at baseline based on retrospective treatment status. Instead, we modeled treatment as a time-varying exposure using a counting process framework with start-stop intervals via the `tmerge` protocol. This ensures that the risk sets are updated dynamically at each event time, preventing the misattribution of waiting-period survival to the active drug's efficacy."

- **Defending the Conditional Exchangeability Assumption**:
  > "Second, to maintain conditional exchangeability over time despite treatment delays, we utilized a cloning-and-censoring approach with IPCW, ensuring the counterfactual baseline characteristics remained balanced as the retrospective follow-up progressed."

- **Empirically Validating the Positivity Assumption**:
  > "Finally, we empirically validated the positivity assumption by confirming that patients across all covariate strata in our HES data had a non-zero probability of receiving standard of care, preventing structurally empty cells in our propensity score distributions."

- **Defending against Unmeasured Confounding via E-Values**:
  > "Although our real-world registry lacks smoking history, our E-value of 1.70 demonstrates that to nullify the observed survival benefit, an unmeasured confounder would need to increase both the likelihood of treatment and the hazard of death by at least 1.70-fold. This is larger than the hazard associated with major clinical indicators like age in our adjusted models, making it highly implausible that a missing confounder of this magnitude exists that is completely independent of the clinical variables adjusted for in our primary analysis."
