---
title: "Survival Analysis & Immortal Time Bias in Real-World Evidence"
subtitle: "Target Trial Emulation, Landmark Analysis, Time-Dependent Cox Models, and IPCW"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "R Code"
    number-sections: true
    reference-location: margin
---

*Methodological guidelines and implementation templates for survival analysis, delayed treatment initiation, and target trial emulation in health technology assessment.*

## Introduction: Causal Inference in Time-to-Event Data

In health technology assessment (HTA), clinical endpoints are frequently **time-to-event outcomes**. While aggregate summaries like Overall Survival (OS) or Progression-Free Survival (PFS) are represented as curves showing the *proportion* or *probability* of surviving patients at specific points in time, the underlying outcome measured at the patient level is the actual **duration of time** (e.g., months or days) elapsed from a defined baseline (such as diagnosis or treatment initiation) until the clinical event of interest (e.g., death, relapse) occurs.

Unlike binary outcomes where logistic regression or propensity score matching are applied statically, survival data introduces a critical dimension: **Time**. 

If time-to-event analysis is not bounded by strict causal constraints, observational analyses can yield severe selection biases. A primary methodological challenge in retrospective registries is **Immortal Time Bias (ITB)**. 

### The Coffee & Heart Disease Paradox: Temporal Ordering and Selection Bias

To understand why a strict time-to-event framework is necessary, consider the "Coffee & Heart Disease Paradox":

- **The Paradox**: Observational studies often show that coffee drinkers have lower rates of heart disease than non-coffee drinkers (Hazard Ratio < 1.0).
- **The Reverse Causality**: In reality, patients who develop heart symptoms or are diagnosed with high risk are advised by doctors to quit coffee.
- **The Methodological Connection**: If we classify exposure statically based on whether a patient drinks coffee during follow-up, we violate the **chronological ordering** of causal inference. A patient who develops heart disease on **Day 50** and **subsequently** quits coffee will be misclassified as a "non-coffee drinker" at the **end of the study**. This attributes their heart event (which occurred **before** quitting) to the non-exposure group, violating the non-anticipatory constraint of time-to-event outcomes. 

To resolve this paradox, we must align the study baseline ($T_0$), measure confounders (like baseline heart risk) strictly *before* exposure initiation, and model exposure dynamically over time. This is exactly what Target Trial Emulation and Time-Varying survival models achieve.

## The Target Trial Emulation (TTE) Framework

Target Trial Emulation is a structured design framework developed by Miguel Hernán and colleagues to **guide the design** and analysis of observational studies. The core principle is that we should design our observational study **to mimic a hypothetical randomized trial as closely as possible.**

### The 7 Dimensions of the Target Trial Approach

According to the official NICE RWE Framework (Section 5.2)[^1], a target trial must be defined across 7 core dimensions to prevent bias and ensure transparency:

[^1]: **Official Source:** NICE Real-World Evidence Framework (2022), Section 5.2 ("The target trial approach"). **Methodological Reference:** Hernán MA, Robins JM (2016). *Using Big Data to Emulate a Target Trial When a Randomized Trial Is Not Available*. Am J Epidemiol.

| Dimension | NICE RWE Framework Requirement | Application to Survival Cohorts |
| :--- | :--- | :--- |
| **1. Eligibility criteria** | Define clear inclusion/exclusion criteria at baseline. | Patients must meet all criteria at Time Zero ($T_0$), ensuring no baseline selection bias. |
| **2. Treatment strategies** | Specify the interventions being compared. | **Arm A:** Initiate active treatment within a defined window. **Arm B:** Initiate Standard of Care (SOC) / no treatment. |
| **3. Assignment procedure** | Define how treatment groups are formed. | Emulate randomization at $T_0$ using Propensity Scores (IPTW or matching) based on baseline covariates. |
| **4. Follow-up period** | Align follow-up start exactly with $T_0$. | Clock starts ticking for all patients at $T_0$. Ends at the event, death, censoring, or study end. |
| **5. Outcomes** | Define a standardized clinical endpoint. | First occurrence of the event (e.g., death, progression) during the follow-up period. |
| **6. Causal effect of interest** | Specify the causal contrast of interest. | **Intention-to-Treat (ITT):** Effect of treatment assignment at $T_0$, regardless of later compliance. |
| **7. Analysis plan** | Specify the statistical estimator. | Cox Proportional Hazards model or parametric survival models, adjusted for baseline confounding. |

### Deep Dive: The New User Design

A fundamental rule of TTE is the **New User Design**[^2]. In observational studies, researchers often mistakenly compare "prevalent users" (patients already on the drug for some time) with "never users." This introduces three major issues:

[^2]: **Defining a "New User":** A patient is classified as a "new user" only if they initiate the drug of interest for the first time at $T_0$. This requires establishing a **washout period** (e.g., 180 or 365 days prior to $T_0$) during which they have no records of the drug in the database, ensuring they are not "prevalent users."

- **Survivor Bias**: Prevalent users are, by definition, "survivors" who have tolerated the drug and survived long enough to be observed in the database. Patients who died or discontinued the drug early due to adverse events are systematically excluded from the treated group, creating a false survival advantage.
- **Baseline Contamination**: To adjust for confounding, we must measure covariates **before** treatment starts. For prevalent users, their baseline measurements (e.g., blood pressure, lab values) are already affected by the treatment, making it impossible to measure their true pre-treatment state.
- **$T_0$ Synchronization**: We cannot align the start of follow-up with the start of treatment if treatment began years before the observation period.

A New User Design enforces that patients are only entered into the cohort at the exact moment they **initiate** the treatment of interest.

### Deep Dive: Intention-to-Treat (ITT) in RWD (Cloning & Censoring)

In randomized trials, the ITT principle analyzes patients in the group they were randomized to, regardless of whether they complied. In retrospective RWD, we often face a "grace period"—a window of time (e.g., 30 or 60 days) after diagnosis during which patients are allowed to initiate treatment.

If we assign patients to the "treated" group because they initiated treatment on day 30, we violate the ITT principle by using "future knowledge" at $T_0$. To solve this, TTE uses **Cloning**:

1. **Cloning at $T_0$**: At the moment of eligibility ($T_0$), we create two identical copies of each patient. One copy is assigned to Arm A (Initiate Treatment), and one copy is assigned to Arm B (Standard of Care).
2. **Follow-up**: Both clones begin follow-up at $T_0$.
3. **Censoring**: If a patient assigned to Arm B initiates treatment on day 30, their Arm B clone is **censored** at day 30 (because they violated the "no treatment" protocol). The Arm A clone continues to be followed.
4. **IPCW Adjustment**: To correct for selection bias introduced by this artificial censoring, we apply **Inverse Probability of Censoring Weights (IPCW)**, which upweights the remaining uncensored clones who share similar baseline characteristics.

### Causal Transparency: Why TTE Aligns with the DataSAT Tool

NICE prefers TTE because it enforces **Transparency**. 

In a simple Cox regression, you often see:
$$\text{Surv}(\text{time}, \text{event}) \sim \text{Treatment} + \text{Confounders}$$

An auditor cannot see:

- When was the treatment started?
- Were the confounders measured *before* or *after* the drug was given?
- Is there a "Healthy User Bias"?

By using the **DataSAT** tool (Data Suitability Assessment Tool), NICE auditors look for **Provenance**—the clear lineage of how a patient moved from a raw database record to a study row. TTE forces the researcher to write a "Trial Protocol" *before* touching the data, making it obvious where $T_0$ is. If an auditor sees that you measured covariates after the drug was given and used them as baseline, the TTE framework will flag this as a "Post-baseline Bias" immediately.

## Applied Scenario: Zero-Time Alignment in an Oncology External Control Arm

To see how these principles apply in an HTA-facing RWE setting, consider **Drug X**, a precision oncology therapy for METex14-mutated non-small cell lung cancer (NSCLC).

::: {.callout-note collapse="true"}
### Declaration: Simulated Proof-of-Methodology Scenario
This METex14 scenario is a **self-generated proof-of-methodology example**, not a real NICE submission, manufacturer dossier, or clinical claim about an identifiable medicine. Drug X, the sample sizes, data source details, and audit responses are simplified or fictionalized to demonstrate how target trial emulation, immortal time bias correction, and DataSAT-style reasoning can be documented in an HTA-facing RWE workflow.
:::

### Scenario Overview: The Drug X Challenge

- **The Innovation**: Due to the rarity of the METex14 mutation and ethical constraints, only a **Single-arm Trial (N=40)** was conducted for Drug X.
- **The Challenge**: A single-arm submission faces a high risk of rejection due to the lack of a randomized head-to-head comparison. Given the aggressive natural history of METex14 NSCLC, simple unadjusted benchmarking is insufficient for demonstrating comparative effectiveness.
- **The Strategy**: We leverage the UK **Hospital Episode Statistics (HES)** database to construct an External Control Arm (ECA) of 200 patients, using the Target Trial Emulation (TTE) framework to satisfy NICE's demand for rigorous causal inference.

### Data Suitability Considerations (DataSAT)

According to Sections 3.1–3.3 of the NICE RWE Framework, we must prove the data is fit for purpose across three core dimensions:

| Dimension | HTA Audit Response (HES Database) | NICE Framework Requirement |
| :--- | :--- | :--- |
| **Provenance** | Data sourced from NHS Digital with a clear audit trail from ICD-10 coding to the final analytical dataset. | Evaluate the **origin and trustworthiness** of data sources, methods of collection, and governance. (Section 3.1) |
| **Quality** | METex14 status confirmed via secondary linkage to SACT and genomic records; completeness checked for key covariates. | Assess the **completeness and accuracy** of eligibility criteria, exposures, and outcomes. (Section 3.2) |
| **Relevance** | HES covers 99% of NHS hospital activity, matching the SOC pathways in the UK. | Ensure findings **generalise to the target population** and clinical care pathways in the NHS. (Section 3.3) |

### Defining $T_0$ (Zero-time Alignment)

To avoid Immortal Time Bias, we align $T_0$ as the moment a patient meets all eligibility criteria **AND** is assigned to treatment, following the target trial approach described in **Section 5.2 of the NICE RWE Framework**. Basically, the index date and treatment strategy assignment must be defined **without using future information**. Treatment status observed after $T_0$ cannot be used retrospectively to create baseline groups unless it is handled through a pre-specified grace-period strategy, such as cloning and censoring.

#### Why Immortal Time Bias is a "Statistical Killer" {.unnumbered}

1. **Logic Paradox**: In RWD studies, patients from diagnosis to actual treatment initiation (Drug X) face delays (e.g., 1 month). If follow-up begins at diagnosis, but grouping is based on whether they ever received treatment, the period before receiving the drug is **immortal**—if the patient **died on Day 15**, they would **never** have lived long enough to receive the drug, and would automatically be classified into the control group (since they are not "treated").
2. **Case Simulation**:

   - **Patient A**: Diagnosed Jan 1, scheduled for Drug X, dies Jan 15 (unexposed). Classified as Control.
   - **Patient B**: Diagnosed Jan 1, initiates Drug X Feb 1, survives until June.
   - **Result**: The Drug X group artificially filters out those who die before treatment initiation, inflating the active group's survival.
3. **TTE Correction**: The survival follow-up clock must be aligned to a clearly defined index date that is comparable across arms. If the causal question starts at diagnosis but treatment can begin during a grace period, the waiting time must not be retrospectively credited to the treated group; it should be handled using cloning-and-censoring, a time-varying exposure model, or a pre-specified landmark strategy.

#### The Control Arm $T_0$ {.unnumbered}

Where does the control arm's $T_0$ start if they never receive the active treatment?

- **Synchronicity Principle**: $T_0$ is the moment the patient meets all clinical eligibility criteria (e.g., METex14 diagnosis + ECOG status).
- **Timeline Alignment**:
  - *Incorrect*: Treated group starts at the "first recorded treatment start date", while controls start at "diagnosis". This gives the treated group an implicit survival benefit.
  - *Correct TTE Approach*: Both groups start follow-up at the exact date they meet all eligibility criteria (the clinical decision point, or clinical equipoise). 

#### Handling the Grace Period {.unnumbered}

In RWD, a patient has a "window" of time (grace period) to initiate treatment (e.g., 30 days after diagnosis). 

- If a patient initiates treatment on Day 20, they contribute 20 days of "unexposed" person-time to the control group.
- Using cloning, we create two clones at diagnosis. The "Control Clone" is followed until Day 20, where it is censored because the patient initiated treatment. The "Treated Clone" starts contributing treated time from Day 20 onwards.

### Implication for Cloning & Censoring

To maintain ITT principles:

- **Anti-Future-Knowledge**: We cannot look ahead to see what the patient did. We clone them at $T_0$ into both paths.
- **Artificial Censoring**: If the patient's actual path deviates from a clone's assigned arm, that clone is censored.
- **IPCW Weighting**: Because censoring might be non-random, we estimate the probability of being censored at time $t$ based on baseline covariates and apply Inverse Probability of Censoring Weights to the remaining patients, dynamically restoring the balance.

## Simulated RWD Cohort: Delayed Treatment Initiation

To make the Drug X scenario concrete, we simulate a simplified RWD cohort with delayed treatment initiation. The simulated dataset is intentionally small enough to audit line by line, but it contains the core features that create immortal time bias:

- baseline prognostic covariates: age and ECOG status;
- treatment selection: older and sicker patients are more likely to receive Drug X;
- treatment delay: treated patients have an observed treatment-initiation delay of 30 to 120 days; later, a 90-day grace period is used to define the target trial strategy;
- time-to-event outcome: death or censoring during follow-up.

We simulate a true treatment Hazard Ratio of **0.70** (30% risk reduction). The purpose of the simulation is not to recreate a real METex14 dataset, but to create a controlled setting where the true parameter is known and different analytic choices can be compared.

```{r}
#| message: false
#| warning: false

library(survival)
library(EValue)
set.seed(42)
n <- 500

# 1. Simulate baseline confounders
age <- rnorm(n, mean = 60, sd = 10)
ecog <- rbinom(n, 1, 0.4)

# 2. Simulate treatment selection (confounding by indication)
# Sicker (ECOG=1) and older patients are more likely to receive Drug X
ps_prob <- plogis(-1.5 + 0.02 * age + 0.5 * ecog)
z_treatment <- rbinom(n, 1, ps_prob)

# Untreated patients are assigned Inf to indicate that treatment initiation never occurs
t_init <- ifelse(z_treatment == 1, runif(n, 30, 120), Inf)

# 3. Simulate survival times (true treatment HR = 0.70)
true_hr <- 0.70
lambda_base <- 0.001 * exp(0.02 * age + 0.5 * ecog)

u <- runif(n)
log_u <- -log(u)

pre_treatment_hazard <- lambda_base * t_init
event_before_treatment <- log_u <= pre_treatment_hazard

t_event <- ifelse(
  event_before_treatment,
  log_u / lambda_base,
  t_init + (log_u - pre_treatment_hazard) / (lambda_base * true_hr)
)

# Censoring between 150 and 730 days
t_censor <- runif(n, 150, 730)
time <- pmin(t_event, t_censor)
event <- as.integer(t_event <= t_censor)

df <- data.frame(id = 1:n, age, ecog, z_treatment, t_init, time, event)

# Future-informed treatment status: this is intentionally biased
df$ever_treated <- ifelse(df$t_init < df$time, 1, 0)
df$t_init_display <- ifelse(is.infinite(df$t_init), NA, df$t_init)

event_count <- sum(df$event)
event_pct <- mean(df$event) * 100

# Helper function used later to render Cox model summaries as kable tables
show_cox <- function(fit, caption) {
  sum_fit <- summary(fit)
  coefs <- sum_fit$coefficients
  cis <- sum_fit$conf.int
  se_col <- if ("robust se" %in% colnames(coefs)) "robust se" else "se(coef)"
  z_col <- if ("z" %in% colnames(coefs)) "z" else colnames(coefs)[4]
  tbl <- data.frame(
    Covariate = rownames(coefs),
    `Hazard Ratio` = round(coefs[, "exp(coef)"], 4),
    `Log HR (coef)` = round(coefs[, "coef"], 4),
    `Std Error` = round(coefs[, se_col], 4),
    `z-value` = round(coefs[, z_col], 3),
    `p-value` = format.pval(coefs[, "Pr(>|z|)"], digits = 4),
    `95% LCL` = round(cis[, 3], 4),
    `95% UCL` = round(cis[, 4], 4),
    check.names = FALSE
  )
  knitr::kable(tbl, caption = caption, align = "lrrrrrrr", row.names = FALSE)
}
```

::: {.callout-note collapse="true"}
### How to Read the Survival-Time Simulation

The simulation uses a simple **piecewise hazard** structure. Before Drug X starts, each patient has an untreated baseline hazard, `lambda_base`. After Drug X starts, the hazard becomes `lambda_base * true_hr`. Because `true_hr = 0.70`, treatment reduces the hazard to 70% of the untreated level.

The quantity `log_u = -log(u)` can be read as each patient's randomly assigned **cumulative hazard threshold**. An event occurs when the cumulative hazard experienced by the patient reaches this threshold.

Before treatment starts, cumulative hazard accumulates at the untreated rate:

$$
\text{pre-treatment cumulative hazard} = \lambda_{\text{base}} \times t_{\text{init}}
$$

If `log_u <= pre_treatment_hazard`, the patient uses up their cumulative hazard threshold before treatment begins, so the event occurs during the untreated period:

$$
t_{\text{event}} = \frac{\text{log\_u}}{\lambda_{\text{base}}}
$$

If the patient survives until treatment initiation, only part of the cumulative hazard threshold has been used. The remaining hazard budget is:

$$
\text{log\_u} - \text{pre-treatment cumulative hazard}
$$

After treatment starts, hazard continues to accumulate, but at the lower treated rate, $\lambda_{\text{base}} \times HR$. Therefore, the total event time is:

$$
t_{\text{event}} =
t_{\text{init}} +
\frac{\text{log\_u} - \text{pre-treatment cumulative hazard}}
{\lambda_{\text{base}} \times HR}
$$

In short, the code simulates a patient who first experiences untreated risk, then, if they live long enough to initiate Drug X, continues accumulating hazard at a lower treated rate.
:::

### Description of the Simulated Cohort

Before fitting any model, we inspect the simulated cohort. The key feature is that treatment is not assigned at time zero. Patients who eventually receive Drug X have a waiting period before their first recorded treatment start date.

```{r}
cohort_summary <- data.frame(
  Metric = c(
    "Total patients",
    "Ever received Drug X during observed follow-up",
    "Observed deaths",
    "Mean age",
    "ECOG 1 proportion",
    "Median observed follow-up (days)",
    "Median treatment initiation delay among treated (days)"
  ),
  Value = c(
    nrow(df),
    sum(df$ever_treated),
    sprintf("%s (%.1f%%)", event_count, event_pct),
    sprintf("%.1f", mean(df$age)),
    sprintf("%.1f%%", mean(df$ecog == 1) * 100),
    sprintf("%.1f", median(df$time)),
    sprintf("%.1f", median(df$t_init_display[df$ever_treated == 1], na.rm = TRUE))
  )
)

knitr::kable(
  cohort_summary,
  caption = "Simulated RWD Cohort Description",
  align = "ll"
)
```

```{r}
baseline_by_group <- aggregate(
  cbind(age, ecog, time, event) ~ ever_treated,
  data = df,
  FUN = mean
)

baseline_by_group$Group <- ifelse(
  baseline_by_group$ever_treated == 1,
  "Received Drug X",
  "Did not receive Drug X"
)

baseline_display <- data.frame(
  Group = baseline_by_group$Group,
  `Mean&nbsp;Age` = round(baseline_by_group$age, 1),
  `ECOG&nbsp;1&nbsp;(%)` = round(baseline_by_group$ecog * 100, 1),
  `Mean&nbsp;Follow-up&nbsp;(days)` = round(baseline_by_group$time, 1),
  `Event&nbsp;Rate&nbsp;(%)` = round(baseline_by_group$event * 100, 1),
  check.names = FALSE
)

knitr::kable(
  baseline_display,
  caption = "Naive Future-Informed Treatment Groups",
  align = "lrrrr",
  escape = FALSE
)
```

This table is deliberately labelled as "future-informed" because the group definition uses information that was not available at $T_0$. A patient who died before treatment initiation could never become "ever treated," so this baseline grouping already embeds immortal time bias.

### Naive Static Cox Model: The Biased Starting Point

The simplest analysis treats `ever_treated` as if it were a baseline treatment variable:

```{r}
fit_biased <- coxph(Surv(time, event) ~ ever_treated + age + ecog, data = df)
show_cox(fit_biased, "Naive Static Cox Model Results (Immortal Time Unadjusted)")
```

This model is intentionally wrong. The `ever_treated` hazard ratio is far below the true simulated HR of 0.70 because patients had to survive long enough to initiate Drug X before they could be classified as treated. Early deaths before treatment initiation are pushed into the untreated group, so the model incorrectly credits this survival selection to the drug.

## Immortal Time Bias in the Simulated Drug X Cohort

This specific problem is known in epidemiology and survival analysis as **Immortal Time Bias (ITB)**. The naive Cox result above illustrates the mechanism: survival time before Drug X initiation is incorrectly credited to the treated group because treatment status was defined using information observed after $T_0$.

Now that the bias mechanism is visible, we evaluate three ways to handle delayed treatment initiation: landmark analysis, time-dependent Cox regression, and cloning-censoring-IPCW.

### First Correction: Landmark Analysis

Landmark Analysis is a classical approach to resolve this:

1. **Select Landmark Time ($T_L$)**: E.g., Day 90.
2. **Exclude Early Events**: Exclude all patients who died or were censored before $T_L$.
3. **Re-classify**: Classify survivors at $T_L$ based on their treatment status at $T_L$.
4. **Follow-up**: Set $T_L$ as the new baseline ($T_{start} = T_L$) and follow patients from that point.

#### HTA Audit of Landmark Analysis {.unnumbered}

- **Pros**: Simple, highly intuitive, and removes the immortal period.
- **Cons**: 
  - **Information Loss**: Discards all events and person-time before $T_L$. If the drug **has high early toxicity**, this signal is completely lost.
  - **Selection Bias**: The populations at $T_L$ may no longer be balanced, as sicker patients in the control group may have died, while sicker patients in the treated group were protected by the "immortal" definition.
  - **Arbitrary Landmark Choice**: The selected landmark time is a design decision, not a statistical truth. If $T_L$ is too early, many later initiators remain classified as untreated; if $T_L$ is too late, more early events are discarded.

Therefore, landmark analysis is usually better positioned as an intuitive diagnostic or sensitivity analysis, rather than the preferred primary analysis for HTA submissions. It is most defensible when the clinical question is explicitly conditional on surviving and remaining eligible until the landmark time; otherwise, time-varying exposure models or cloning-censoring-IPCW are generally more aligned with the original target trial question.

### Applying Landmark Analysis to the Simulated Cohort

We now apply a Day 90 landmark analysis to the simulated Drug X cohort. Patients who died or were censored before Day 90 are excluded. Among those who remain under observation at Day 90, treatment status is re-defined based on whether Drug X had started by that date.

```{r}
t_landmark <- 90

# Exclude patients who died or were censored before Day 90
df_landmark <- df[df$time > t_landmark, ]

# Classify treatment status at the landmark
df_landmark$treated_at_landmark <- ifelse(df_landmark$t_init <= t_landmark, 1, 0)

# Shift follow-up time so Day 90 becomes the new time zero
df_landmark$time_shifted <- df_landmark$time - t_landmark

fit_landmark <- coxph(
  Surv(time_shifted, event) ~ treated_at_landmark + age + ecog,
  data = df_landmark
)

show_cox(fit_landmark, "Landmark Analysis (Day 90) Cox Model Results")
```

The landmark model avoids using Day 90 treatment status to define baseline groups at Day 0. However, it changes the causal question: the result applies only to patients who survived and remained observable until the landmark time.

## Time-Dependent Cox Model & HTA Practice

The time-dependent Cox model is an extension of the standard Cox proportional hazards model that allows a patient's treatment status to **change during follow-up**. This makes it useful for auditing and correcting immortal time bias when treatment initiation is delayed.

### What the Model Does

In a standard Cox model, treatment is usually treated as fixed at baseline:

$$
h_i(t) = h_0(t)\exp(\beta Z_i + \gamma L_i)
$$

where $Z_i$ is a single treatment indicator for patient $i$, and $L_i$ represents baseline covariates such as age or ECOG status.

The problem is that delayed treatment is not fixed at baseline. A patient may be untreated from Day 0 to Day 60, then treated from Day 60 onward. A time-dependent Cox model replaces the fixed treatment indicator $Z_i$ with a time-updated treatment indicator $Z_i(t)$:

$$
h_i(t) = h_0(t)\exp(\beta Z_i(t) + \gamma L_i)
$$

Here, $Z_i(t)=0$ before treatment starts and $Z_i(t)=1$ after treatment starts. At each event time, the Cox model compares patients according to their treatment status **at that moment**, not according to information observed later in follow-up.

This is the key correction for immortal time bias: the waiting period before treatment is counted as untreated person-time, rather than being retrospectively assigned to the treated group.

### Counting Process Data Structure

To estimate this model, the data usually need to be written in counting process format. Rather than one row per patient, the dataset is split into multiple intervals `(tstart, tstop]` where treatment status remains constant.

#### Start-Stop Data Format Example (Patient #101) {.unnumbered}
Patient #101 has a treatment initiation lag of 60 days and dies on Day 500. We split this patient's record into two lines:

| Patient ID | Start (Day) | Stop (Day) | Status (Event) | Treatment ($Z$) |
| :--- | :--- | :--- | :--- | :--- |
| #101 | 0 | 60 | 0 (Alive) | 0 (Not treated) |
| #101 | 60 | 500 | 1 (Death) | 1 (Treated) |

At Day 60, the patient's treatment status switches from not treated to treated. The patient is not removed from follow-up; instead, their contribution to the risk set is updated from untreated person-time to treated person-time.

::: {.callout-note collapse="true"}
### Why This Is Not Duplicating the Patient

In a Cox model, the partial likelihood is built at each observed event time. At each event time, the model rebuilds the risk set: who is still alive, still under follow-up, and what are their covariate values at that moment?

In a standard Cox model, treatment status is fixed, so the model always reads the same $Z_i$ for patient $i$. In a time-dependent Cox model, the model reads $Z_i(t)$, meaning the patient's treatment status at the event time being evaluated.

For Patient #101, this means:

- before Day 60, the patient can enter risk sets as `Z = 0`;
- after Day 60, the same patient can enter risk sets as `Z = 1`;
- at no event time do both rows enter the same risk set simultaneously.

The two rows therefore represent two non-overlapping risk intervals for the same patient, not two independent patients. The patient contributes untreated person-time before Day 60, treated person-time after Day 60, and one death event at Day 500.
:::

### Landmark vs. Time-Dependent Cox

| Dimension | Time-dependent Cox | Landmark Analysis |
| :--- | :--- | :--- |
| **Data Retention** | Retains all patients and events from Day 0. | Discards patients/events prior to Landmark $T_L$. |
| **Causal Question** | Causal effect starting from Day 0. | Conditional effect given survival to $T_L$. |
| **Early Toxicity** | Captures early adverse event signals. | Masked if events occur before $T_L$. |
| **HTA Audit Preference** | Preferred for primary analysis. | Used primarily as a sensitivity analysis. |

### Applying the Time-Dependent Cox Model

We now restructure the same simulated cohort into counting process format using `tmerge()`.

```{r}
# 1. Initialize dataset with the primary event time
df_tv <- tmerge(df, df, id = id, event = event(time, event))

# 2. Merge the time-dependent treatment initiation event
df_tv <- tmerge(df_tv, df, id = id, treatment_status = tdc(t_init))

knitr::kable(
  head(df_tv[df_tv$id %in% 1:3, c("id", "tstart", "tstop", "event", "treatment_status")]),
  caption = "Restructured Counting Process Intervals (Patient ID 1-3)",
  align = "c"
)
```

As shown, if a patient receives treatment at a delayed time, their follow-up is split:

- **Interval 1 `[0, t_init]`**: `treatment_status = 0`, contributing untreated person-time.
- **Interval 2 `[t_init, time]`**: `treatment_status = 1`, contributing treated person-time after Drug X starts.

```{r}
fit_tv <- coxph(
  Surv(tstart, tstop, event) ~ treatment_status + age + ecog,
  data = df_tv
)

show_cox(fit_tv, "Time-Dependent Cox Model Results (Immortal Time Corrected)")
```

The time-dependent Cox model keeps the early follow-up time and assigns it to the correct treatment state. In this simulation, it is expected to recover a hazard ratio closer to the true treatment effect than the naive static model.

[^strategy-vs-status]: **Treatment status vs. treatment strategy:** Imagine Drug X has strong biological efficacy once started, but testing and approval delays mean many patients die before receiving it. A time-dependent Cox model may show a large hazard reduction once patients are on Drug X. A cloning-censoring-IPCW analysis may still find limited benefit for the overall "start within 90 days" strategy unless treatment can actually be started fast enough. This is the difference between estimating an on-treatment effect and evaluating a baseline treatment-initiation strategy.

## Cloning, Censoring, and IPCW in Target Trial Emulation

Time-dependent Cox models are useful for auditing whether delayed treatment initiation has created immortal time bias. However, when the target trial question is defined as a comparison of **treatment strategies**—for example, "initiate Drug X within 90 days of diagnosis" versus "do not initiate Drug X within 90 days"—a trial-emulation analysis often requires **cloning, artificial censoring, and inverse probability of censoring weighting (IPCW)**. Here, the 90-day grace period is not a minor technical detail; it is part of the strategy definition itself and should be justified, audited, and tested in sensitivity analyses.[^strategy-vs-status]

### Why Artificial Censoring Creates Bias

In a cloned analysis, each patient is copied into every treatment strategy at $T_0$:

- One clone is assigned to the "initiate within grace period" strategy.
- One clone is assigned to the "do not initiate within grace period" strategy.

Each clone is followed from the same $T_0$. If the patient's observed behavior becomes incompatible with a clone's assigned strategy, that clone is artificially censored. For example, if the "no treatment" clone initiates Drug X on Day 20, that clone is censored on Day 20.

This artificial censoring is not necessarily random. Patients who start treatment early may differ systematically from patients who do not start treatment, and patients who remain compatible with a strategy may no longer represent the original target population. IPCW is used to reduce this selection bias.

### IPCW Intuition

The "C" in IPCW stands for **censoring**. The core idea is to weight each remaining uncensored clone by the inverse of its probability of remaining uncensored:

$$
w_i(t) = \frac{1}{P(C_i(t)=0 \mid L_i)}
$$

where $C_i(t)=0$ means clone $i$ remains uncensored at time $t$, and $L_i$ represents baseline or time-updated covariate history.

The intuition is simple: if a type of patient is highly likely to be censored, but one such patient remains uncensored, that patient receives more weight because they represent similar patients who were censored.

### Stabilized Weights

In practice, stabilized weights are often preferred because they reduce variance:

$$
SW_i(t) =
\frac{P(C_i(t)=0 \mid A_i)}
{P(C_i(t)=0 \mid A_i, L_i)}
$$

The denominator models censoring using strategy assignment and prognostic covariates. The numerator uses a simpler model, often including strategy assignment only. Extreme weights should be inspected and, if necessary, truncated using a pre-specified rule.

### HTA Audit Checks

For an HTA-facing analysis, IPCW should not be presented as a black box. The submission should document:

- the protocol rule that triggers artificial censoring;
- the censoring model covariates and timing;
- the distribution of stabilized weights;
- any truncation threshold and its sensitivity analysis;
- whether positivity is plausible across clinically relevant subgroups.

### Applying Cloning-Censoring-IPCW

We now emulate a strategy-level target trial with a 90-day grace period:

- **Strategy 1**: initiate Drug X within 90 days of diagnosis.
- **Strategy 0**: do not initiate Drug X within 90 days of diagnosis.

Each patient is cloned into both strategies. A clone is artificially censored when the observed treatment path becomes incompatible with its assigned strategy.

```{r}
grace_period <- 90

# 1. Clone each patient into both treatment strategies
df_clone <- rbind(
  transform(df, strategy = 1, strategy_label = "Initiate within 90 days"),
  transform(df, strategy = 0, strategy_label = "No initiation within 90 days")
)

# 2. Define artificial censoring rules
df_clone$artificial_censor_time <- Inf

# Strategy 1 violation: no treatment initiation by the end of the grace period
df_clone$artificial_censor_time[
  df_clone$strategy == 1 &
    df_clone$t_init > grace_period &
    df_clone$time > grace_period
] <- grace_period

# Strategy 0 violation: treatment initiation during the grace period
df_clone$artificial_censor_time[
  df_clone$strategy == 0 &
    df_clone$t_init <= grace_period &
    df_clone$t_init < df_clone$time
] <- df_clone$t_init[
  df_clone$strategy == 0 &
    df_clone$t_init <= grace_period &
    df_clone$t_init < df_clone$time
]

# 3. Observed follow-up under each assigned strategy
df_clone$obs_time <- pmin(df_clone$time, df_clone$artificial_censor_time)
df_clone$outcome_event <- as.integer(df_clone$event == 1 & df_clone$time <= df_clone$artificial_censor_time)
df_clone$artificial_censored <- as.integer(df_clone$artificial_censor_time < df_clone$time)
df_clone$uncensored <- 1 - df_clone$artificial_censored

knitr::kable(
  head(df_clone[, c("id", "strategy_label", "time", "event", "t_init", "obs_time", "outcome_event", "artificial_censored")], 8),
  caption = "Cloned Target Trial Dataset with Artificial Censoring Rules",
  align = "c"
)
```

Next, we estimate stabilized IPCW. This is a simplified subject-level implementation for demonstration; production analyses commonly estimate censoring probabilities over time using discrete-time pooled logistic models or start-stop data.

```{r}
# 4. Estimate stabilized inverse probability of censoring weights
censor_den <- glm(uncensored ~ strategy + age + ecog, family = binomial(), data = df_clone)
censor_num <- glm(uncensored ~ strategy, family = binomial(), data = df_clone)

df_clone$p_uncens_den <- predict(censor_den, type = "response")
df_clone$p_uncens_num <- predict(censor_num, type = "response")
df_clone$ipcw <- df_clone$p_uncens_num / df_clone$p_uncens_den

weight_summary <- data.frame(
  Metric = c("Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum"),
  Stabilized_IPCW = as.numeric(summary(df_clone$ipcw))
)

knitr::kable(
  weight_summary,
  caption = "Stabilized IPCW Distribution",
  digits = 3,
  align = "lr"
)
```

Finally, we fit a weighted Cox model comparing the two grace-period strategies. The `cluster(id)` term accounts for the fact that each original patient contributes two cloned records.

```{r}
fit_ipcw <- coxph(
  Surv(obs_time, outcome_event) ~ strategy + age + ecog + cluster(id),
  data = df_clone,
  weights = ipcw
)

show_cox(fit_ipcw, "Cloning-Censoring-IPCW Cox Model Results")
```

## Causal Evidence Reconciliation

An **estimand** means the exact causal question being estimated. The methods below all address delayed treatment initiation, but they do not answer precisely the same question:

- **Naive static Cox** asks a biased future-informed question: did patients who eventually received Drug X survive longer than those who did not?
- **Landmark analysis** estimates a conditional effect among patients who survived and remained observable until Day 90.
- **Time-dependent Cox** estimates the hazard contrast associated with current treatment status at each event time.
- **Cloning-censoring-IPCW** estimates a strategy contrast, such as "initiate Drug X within 90 days" versus "do not initiate within 90 days," starting from the original eligibility date.

The table below summarizes the estimates across the four analyses compared with the true simulated treatment effect.

```{r}
# Extract statistics
hr_biased  <- exp(coef(fit_biased)["ever_treated"])
se_biased  <- summary(fit_biased)$coefficients["ever_treated", "se(coef)"]

hr_land    <- exp(coef(fit_landmark)["treated_at_landmark"])
se_land    <- summary(fit_landmark)$coefficients["treated_at_landmark", "se(coef)"]

hr_tv      <- exp(coef(fit_tv)["treatment_status"])
se_tv      <- summary(fit_tv)$coefficients["treatment_status", "se(coef)"]

hr_ipcw    <- exp(coef(fit_ipcw)["strategy"])
se_ipcw    <- summary(fit_ipcw)$coefficients["strategy", "robust se"]

# Combine into reconciliation matrix
reconciliation_tbl <- data.frame(
  Estimator = c("Ground Truth (True Parameter)", "Biased Model (Static Baseline)", "Landmark Analysis (Day 90)", "Time-Dependent Cox Model", "Cloning-Censoring-IPCW"),
  "Hazard Ratio (HR)" = c(true_hr, hr_biased, hr_land, hr_tv, hr_ipcw),
  "Standard Error" = c(NA, se_biased, se_land, se_tv, se_ipcw),
  "95% LCL" = c(NA, exp(confint(fit_biased)["ever_treated", 1]), exp(confint(fit_landmark)["treated_at_landmark", 1]), exp(confint(fit_tv)["treatment_status", 1]), exp(confint(fit_ipcw)["strategy", 1])),
  "95% UCL" = c(NA, exp(confint(fit_biased)["ever_treated", 2]), exp(confint(fit_landmark)["treated_at_landmark", 2]), exp(confint(fit_tv)["treatment_status", 2]), exp(confint(fit_ipcw)["strategy", 2])),
  "Bias vs. Truth (Log HR)" = c(0.0000, coef(fit_biased)["ever_treated"] - log(true_hr), coef(fit_landmark)["treated_at_landmark"] - log(true_hr), coef(fit_tv)["treatment_status"] - log(true_hr), coef(fit_ipcw)["strategy"] - log(true_hr)),
  check.names = FALSE
)

# Format values for display
reconciliation_tbl[, 2:6] <- round(reconciliation_tbl[, 2:6], 4)
reconciliation_tbl[is.na(reconciliation_tbl)] <- "N/A"

knitr::kable(reconciliation_tbl, caption = "Survival Parameter Reconciliation Across Estimators", align = "lrrrrr")
```

## Quantitative Bias Analysis: E-Value Sensitivity Analysis

When submitting RWE to HTA bodies, we must audit the robustness of our findings against unmeasured confounding. We compute the E-value for the time-dependent Cox Hazard Ratio ($HR = `r round(exp(coef(fit_tv)["treatment_status"]), 3)`$).

### E-Value Math & Formula Setup

The E-value is defined as the minimum strength of association on the risk ratio scale that an unmeasured confounder ($U$) must have with both:
1. The treatment assignment ($Z$)
2. The clinical outcome ($Y$)

to fully explain away the observed treatment-outcome association.

The formula depends on the cumulative incidence of the outcome:

1. **Rare Outcome Assumption** (cumulative incidence < 15%):

   The Hazard Ratio approximates the Relative Risk ($RR \approx HR$). For protective effects ($HR < 1.0$), we first invert the HR ($HR^* = 1 / HR$) to get a value $> 1.0$:
   $$\text{E-value} = HR^* + \sqrt{HR^* \cdot (HR^* - 1)}$$

2. **Common Outcome Correction** (cumulative incidence > 15%):

   The rare outcome assumption is violated, and we must apply the square root approximation:
   $$RR \approx \sqrt{HR}$$
   The corresponding E-value is calculated as:
   $$\text{E-value} = \sqrt{HR^*} + \sqrt{\sqrt{HR^*} \cdot (\sqrt{HR^*} - 1)}$$

In our cohort, `r event_count` out of `r n` patients (`r round(event_pct, 1)`%) experienced the event, meaning the outcome is common. We must apply the common outcome correction.

```{r}
# 1. Extract HR and confidence limits from the time-varying Cox model
hr_est_val <- exp(coef(fit_tv)["treatment_status"])
hr_lcl_val <- exp(confint(fit_tv)["treatment_status", 1])
hr_ucl_val <- exp(confint(fit_tv)["treatment_status", 2])

# 2. Compute E-value (common outcome correction: rare = FALSE)
ev_survival <- evalues.HR(est = hr_est_val, lo = hr_lcl_val, hi = hr_ucl_val, rare = FALSE)
ev_survival_df <- as.data.frame(ev_survival)

# 3. Format E-value table
ev_survival_tbl <- data.frame(
  "Metric / Parameter" = c("Hazard Ratio (HR)", "E-value (Treatment Effect Bound)"),
  "Point Estimate" = c(sprintf("%.3f", ev_survival_df["HR", "point"]), sprintf("%.3f", ev_survival_df["E-values", "point"])),
  "95% CI Lower Bound" = c(sprintf("%.3f", ev_survival_df["HR", "lower"]), sprintf("%.3f", ev_survival_df["E-values", "lower"])),
  "95% CI Upper Bound" = c(sprintf("%.3f", ev_survival_df["HR", "upper"]), "N/A"),
  check.names = FALSE
)

knitr::kable(ev_survival_tbl, caption = "E-Value Analysis for Survival Sensitivity to Unmeasured Confounding", align = "lrrr")
```

The point estimate E-value of **`r round(ev_survival_df["E-values", "point"], 3)`** means that an unmeasured confounder would need to increase the probability of treatment assignment by **`r round(ev_survival_df["E-values", "point"], 3)`-fold** and increase the hazard of death by **`r round(ev_survival_df["E-values", "point"], 3)`-fold** to fully explain away the observed survival benefit.

### Algebraic Derivation of the E-Value (VanderWeele & Ding 2016)

To understand the mathematical strength of the E-value, we trace its algebraic derivation from the **Bias Factor** model.

Let $U$ be a binary unmeasured confounder. The observed relative risk $RR_{obs}$ is related to the true relative risk $RR_{true}$ by:
$$RR_{obs} = RR_{true} \times \text{Bias}$$

where the Bias Factor is defined as:
$$\text{Bias} = \frac{RR_{ZU} \times RR_{UY}}{RR_{ZU} + RR_{UY} - 1}$$

Here:

- $RR_{ZU}$ is the risk ratio associating the treatment $Z$ with the unmeasured confounder $U$.
- $RR_{UY}$ is the risk ratio associating the unmeasured confounder $U$ with the outcome $Y$.

To explain away an observed effect (i.e., to show that the true effect is null, $RR_{true} = 1.0$), the Bias Factor must equal or exceed the observed relative risk:
$$\text{Bias} \ge RR_{obs}$$

To find the minimum strength of association required for both relationships, we define a single threshold $E = RR_{ZU} = RR_{UY}$. Substituting $E$ into the Bias formula:
$$\frac{E^2}{2E - 1} \ge RR_{obs}$$

This can be rearranged into a quadratic inequality:
$$E^2 - 2RR_{obs} \cdot E + RR_{obs} \ge 0$$

Solving the quadratic equation $E^2 - 2RR_{obs} \cdot E + RR_{obs} = 0$ using the quadratic formula:
$$E = \frac{2RR_{obs} \pm \sqrt{4RR_{obs}^2 - 4RR_{obs}}}{2} = RR_{obs} + \sqrt{RR_{obs}(RR_{obs} - 1)}$$

This is the exact mathematical derivation of the E-value formula. It represents the absolute minimum joint relationship required for an unmeasured confounder to explain away the observed treatment effect.

## HTA Audit & Interview Guide

### Interview Audit Checklist

When auditing a manufacturer's RWE survival submission (or preparing for an HTA interview), follow these three key checks:

1. **Zero-time Alignment Check**:

   - *Question*: Identify the exact calendar date of $T_0$ for both cohorts. Is $T_0$ defined by the clinical event (diagnosis) or the administrative event (first prescription)?
   - *Audit Rule*: Follow-up must start for both cohorts when they meet the eligibility criteria. If the treated group follow-up starts at drug initiation and controls start at diagnosis, the analysis is invalid.

2. **Immortal Time Bias Screen**:

   - *Question*: Check the initial segment of the Kaplan-Meier curve. Does the treated arm show a "zero-risk" flat line at the start of follow-up?
   - *Audit Rule*: A flat line on the treatment curve indicates that patients who died early were excluded from the treated group, indicating Immortal Time Bias.

3. **Statistical Correction Audit**:

   - *Question*: If a treatment initiation lag exists, was it addressed using a time-varying covariate or a Landmark approach?
   - *Audit Rule*: Standard Cox models using a static baseline treatment indicator are statistically invalid in the presence of treatment initiation delays.

### Expert HTA Articulation

Use the following professional statements to articulate survival analysis defense strategies during HTA reviews or interviews:

- **Justifying Time-Varying Cox Models Over Static Models**:
  > "To prevent the introduction of lead-time and immortal time biases, we did not define treatment statically at baseline based on retrospective treatment status. Instead, we modeled treatment as a time-varying exposure using a counting process framework with start-stop intervals via the `tmerge` protocol. This ensures that the risk sets are updated dynamically at each event time, preventing the misattribution of waiting-period survival to the active drug's efficacy."

- **Defending the Conditional Exchangeability Assumption**:
  > "Second, to maintain conditional exchangeability over time despite treatment delays, we utilized a cloning-and-censoring approach with IPCW, ensuring the counterfactual baseline characteristics remained balanced as the retrospective follow-up progressed."

- **Empirically Validating the Positivity Assumption**:
  > "Finally, we empirically validated the positivity assumption by confirming that patients across all covariate strata in our HES data had a non-zero probability of receiving standard of care, preventing structurally empty cells in our propensity score distributions."

- **Defending against Unmeasured Confounding via E-Values**:
  > "Although our real-world registry lacks smoking history, our E-value of 1.70 demonstrates that to nullify the observed survival benefit, an unmeasured confounder would need to increase both the likelihood of treatment and the hazard of death by at least 1.70-fold. This is larger than the hazard associated with major clinical indicators like age in our adjusted models, making it highly implausible that a missing confounder of this magnitude exists that is completely independent of the clinical variables adjusted for in our primary analysis."
