---
title: "TA1092 EAG Appraisal Case: Pembrolizumab + Chemotherapy in Endometrial Cancer"
subtitle: "Survival Extrapolation, Subgroup Uncertainty, and the EAG Review Lens"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    code-annotations: below
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

set.seed(2026)
```

::: {.callout-important icon=false}
## Case Study Status

This is a working Quarto template for a NICE TA1092 case study. The R code uses synthetic placeholder parameters until the public NICE committee papers and appraisal tables are manually abstracted. The final version should replace the placeholder inputs with values traceable to the NICE guidance, company submission, and EAG report.
:::

## Project Scope Statement

**In scope:**

- Reconstruct a simplified three-state Partitioned Survival Model (PSM): progression-free, progressed disease, death.
- Implement a technical R replication of survival extrapolation and state occupancy.
- Compare a company-style base case with EAG-style alternative assumptions.
- Add an explicit **EAG critique lens**: what the EAG challenged, why it mattered, and how it affected the committee's preferred assumptions.
- Translate appraisal disagreements into practical HEOR modelling lessons.

**Out of scope:**

- Full confidential ICER replication.
- Patient-level reconstruction from Kaplan-Meier curves unless digitised data are later added.
- Full probabilistic sensitivity analysis (PSA).
- Commercial arrangement or confidential discount reconstruction.
- Recreating the exact manufacturer model workbook.

**Time budget:** ~12-15 hours total

| Task | Allocated time |
|---|---:|
| NICE document abstraction | 2.0 h |
| Decision problem and evidence summary | 1.5 h |
| EAG critique extraction | 3.0 h |
| R technical replication | 5.0 h |
| Interpretation and portfolio write-up | 3.0 h |

**Deliberate trade-off:** The goal is not to reproduce NICE's confidential base-case ICER. The goal is to demonstrate how an EAG interrogates survival modelling, subgroup evidence, and economic assumptions under uncertainty.

## Background

### Clinical Context

**Disease area:** untreated primary advanced or recurrent endometrial cancer.

**Intervention:** pembrolizumab with carboplatin and paclitaxel, followed by pembrolizumab maintenance.

**Comparator:** carboplatin with paclitaxel.

**Key clinical trial:** KEYNOTE-868 / NRG-GY018.

**Main modelling issue:** the intervention combines immunotherapy with platinum-taxane chemotherapy. The health economic question is whether the additional cost of pembrolizumab is justified by improved progression-free survival (PFS), overall survival (OS), and quality-adjusted life years (QALYs), especially when long-term OS data are immature.

### Decision Problem

| Element | Working abstraction |
|---|---|
| Population | Adults with untreated primary advanced or recurrent endometrial cancer |
| Subgroups | dMMR/MSI-H and pMMR/MSS populations |
| Intervention | Pembrolizumab + carboplatin + paclitaxel, then pembrolizumab maintenance |
| Comparator | Carboplatin + paclitaxel |
| Outcomes | PFS, OS, adverse events, health-related quality of life, costs, QALYs |
| Model type | Three-state PSM |
| NICE decision tension | Whether extrapolated survival benefit is clinically plausible and cost-effective across molecular subgroups |

### Documents to Abstract

| Source | Status | Notes to extract |
|---|---|---|
| NICE guidance | TODO | Final recommendation, stopping rule, subgroup restrictions |
| Committee discussion | TODO | Committee reasoning and preferred assumptions |
| Committee papers | TODO | Company base case, EAG critique, scenario analyses |
| Company submission | TODO | Survival models, utilities, costs, treatment duration |
| EAG report | TODO | Concerns, alternative assumptions, exploratory ICERs |

## My Independent Assessment

> Written before reading the full EAG report in detail.

### What I Would Check First

From a reviewer perspective, this appraisal raises four immediate questions:

1. **Survival maturity:** Is OS mature enough to support a lifetime extrapolation?
2. **Subgroup credibility:** Are dMMR/MSI-H and pMMR/MSS results stable enough for separate modelling?
3. **Curve behaviour:** Do the selected PFS and OS curves produce clinically plausible long-term survival?
4. **Treatment effect duration:** Does the pembrolizumab benefit persist indefinitely, wane over time, or stop after treatment discontinuation?

### Initial Model Risk Map

| Model domain | Why it matters | Initial risk level | What I need to verify |
|---|---|---:|---|
| OS extrapolation | Usually a major ICER driver in oncology PSMs | High | Chosen distribution, clinical validation, tail behaviour |
| PFS extrapolation | Determines time on progression-free utility and treatment costs | High | Curve crossing and subgroup fit |
| Treatment duration | Pembrolizumab has a capped treatment duration | Medium | Whether cost stopping and effect stopping are handled separately |
| Treatment effect waning | EAGs often challenge indefinite immunotherapy benefit | High | Waning start/end and effect on ICER |
| Utilities | Utility source may not match modelled population | Medium | Trial EQ-5D vs external utility values |
| Subsequent treatments | Can change OS and post-progression costs | Medium | Mix, duration, and consistency with NHS practice |

### Early View

My prior expectation is that the EAG critique will focus less on the PSM structure itself and more on the **credibility of extrapolated survival benefit**. In a NICE appraisal, the PSM is usually acceptable as a structure; the disputed part is whether the survival curves and treatment-effect assumptions produce a clinically plausible long-term trajectory.

## EAG Critique Lens

### How the EAG Reviews This Type of Case

The EAG is not only checking whether the model runs. It is asking whether the company's assumptions are:

- internally consistent with trial evidence;
- externally plausible against clinical knowledge;
- aligned with NHS treatment pathways;
- transparent enough for committee decision-making;
- conservative enough when evidence is immature.

This distinction is important for the portfolio: a health economist in consulting must be able to understand both the company logic and the EAG logic.

### Working Reflection on Issue 1

Issue 1 is my first example of how the EAG examines the alignment between trial design, trial objectives, trial results, and later interpretation of those results.

My current understanding is that the EAG is concerned about **target consistency**. If the trial was originally designed to answer one type of question, for example treatment effects within separate dMMR and pMMR cohorts, but the later economic modelling relies on a different question, for example an all-comer analysis combining those populations, then there is a risk of misalignment between the evidence generation process and the modelled decision problem.

In practical HTA terms, the concern is not simply that all-comer analysis is wrong. The concern is that a post-hoc all-comer analysis may be less reliable if it was not the original basis of the trial design and statistical analysis plan. This can create uncertainty about whether the estimated treatment effect is truly representative of the population used in the economic model.

This issue is therefore mainly an **evidence validity** issue rather than a direct cost-effectiveness parameter issue. It affects confidence in the survival inputs and subgroup assumptions that feed into the model. For this portfolio, I will treat Issue 1 as part of the EAG's broader logic: before challenging model extrapolations, the EAG first checks whether the clinical evidence being extrapolated is aligned with the decision problem.

Issue 2 extends the same evidence-validity concern to more granular subgroups. The EAG notes that some NICE-specified subgroups, such as local versus metastatic recurrence or prior surgery history, were not systematically captured in the pivotal trial. My takeaway is that this is not a main technical replication target, but it reinforces an important appraisal principle: broad modelled indications can become uncertain when the trial evidence does not fully describe clinically relevant heterogeneity.

This type of subgroup limitation is probably common in oncology appraisals because pivotal trials are rarely large enough to support robust analyses across many clinically detailed subgroups. The key point is not that every missing subgroup analysis is equally important. The EAG appears to elevate subgroup limitations when they are **decision-relevant**: when the subgroup was specified in the NICE scope, may have different prognosis or treatment effect, affects the clinical pathway, or could change the cost-effectiveness conclusion.

### Working Reflection on Issue 3

Issue 3 is a stronger candidate for a technical deep dive because it connects directly to survival modelling. The EAG's concern is that KEYNOTE-868 has relatively short follow-up, while the economic model needs to project PFS, OS, costs, and QALYs over a much longer horizon.

My current interpretation is that short follow-up is not a problem simply because "more data would always be better". It becomes decision-relevant when the observed trial period covers only the early part of the survival curves, but a material share of the modelled incremental benefit is generated in the extrapolated tail. In that setting, the model result depends heavily on assumptions about curve selection, long-term tail behaviour, treatment effect duration, and clinical plausibility.

After checking the EAG report and later factual-accuracy material in the committee papers, my more precise working interpretation is that the short-follow-up concern can be condensed into a few case-specific signals rather than a universal threshold. The median follow-up appears to be around 16.3 months, while the planned active treatment period is approximately 2.2 years. This means that the median observation time does not even fully cover the planned treatment duration. In addition, the pembrolizumab arm median OS was not reached, meaning the trial had not yet observed the time point at which 50% of patients had died.

These are not formal NICE cut-offs, but they are useful data-maturity signals. If median OS has not been reached, long-term OS benefit is still not directly observed. If median follow-up is shorter than the intended treatment duration, there is limited evidence about what happens after patients complete or stop pembrolizumab. This makes assumptions about sustained treatment effect, treatment waning, and long-term survival tails especially important.

NotebookLM review of the EAG material suggests that the EAG's concern was broader than these two signals. The EAG also appears to use several related judgement signals when assessing whether the follow-up is sufficient for long-term extrapolation:

| EAG judgement signal | Why it matters for Issue 3 | How I could show it in the portfolio |
|---|---|---|
| Few patients at risk in the KM tail | Late KM steps become unstable, so the apparent tail shape may be driven by very small numbers | Add numbers-at-risk context and avoid overinterpreting late KM steps |
| Median OS not reached | The trial has not directly observed the point at which half of patients have died | Add a horizontal 50% survival line to OS plots and mark whether it is crossed |
| Models diverge after the observed period | Several models can fit the observed KM but imply different long-term survival | Use a distribution fan plot and landmark survival table |
| Uncertain treatment-effect duration | Short follow-up gives limited evidence on whether pembrolizumab benefit persists after stopping treatment | Compare sustained benefit versus treatment-effect waning scenarios |
| External validation is weak or indirect | Evidence from melanoma or dMMR-only studies may not generalise to advanced/recurrent endometrial cancer or pMMR disease | Document external evidence limitations rather than treating them as proof of sustained benefit |
| Expert landmark estimates are speculative | Advisory-board estimates can guide plausibility but are not empirical trial evidence | Use expert landmarks as plausibility checks, not as observed validation |

This gives a more defensible interpretation than using a simple follow-up-to-horizon ratio. The EAG's concern is cumulative: immature OS, sparse KM tails, divergent extrapolation models, and uncertain post-treatment benefit all point in the same direction. Together, they make the long-term extrapolated survival benefit less secure.

For the portfolio, I should therefore avoid treating Issue 3 as a generic limitation. The useful demonstration is to show the gap between **observed evidence** and **modelled horizon**. A practical way to do this is:

- show the reconstructed KM curve and fitted parametric survival curve on the same plot;
- mark the approximate observed follow-up boundary with a vertical line;
- shade the extrapolated region beyond the observed trial data;
- compare plausible parametric distributions that fit similarly during the observed period but diverge in the long-term tail;
- report landmark survival predictions, for example 5-year, 10-year, and 20-year PFS/OS.

This also clarifies the role of uncertainty. Short follow-up can widen statistical uncertainty, but the larger HTA concern is often **structural uncertainty**: different plausible extrapolation models may produce meaningfully different long-term survival, even if their in-sample AIC/BIC or visual fit is similar. This is why an EAG may challenge an AIC-selected curve if its long-term behaviour lacks clinical credibility.

Initial technical plan for this issue:

| Demonstration | Purpose | Possible output |
|---|---|---|
| Observed vs modelled horizon | Show why follow-up is short in model terms | Table comparing follow-up length with model horizon |
| Extrapolation-zone plot | Separate observed KM evidence from modelled tail | KM + fitted curve + shaded extrapolation region |
| Distribution fan plot | Show structural uncertainty across plausible curves | Multiple parametric survival tails |
| Landmark survival table | Quantify tail differences | 5-, 10-, and 20-year survival by distribution |
| Simplified scenario impact | Link survival uncertainty to economic outputs | Directional QALY / ICER impact |

### Working Reflection on Issue 4

Issue 4 is a utility evidence limitation. The EAG notes that HRQoL/patient-reported outcomes were assessed in the pMMR cohort only, with no equivalent dMMR HRQoL data available. The company's explanation is that the smaller dMMR sample size created insufficient statistical power, so HRQoL analyses were prespecified for pMMR.

My takeaway is that this is not a main technical replication target at this stage, but it is important for the appraisal logic. In a cost-utility model, HRQoL evidence feeds into health-state utilities and therefore QALYs. If utility evidence comes only from pMMR patients but is applied more broadly, then the representativeness of utility inputs becomes uncertain, especially for subgroup-specific cost-effectiveness.

### EAG Challenge Log

| Issue | Company position | EAG concern | Why the EAG cares | Potential ICER direction |
|---|---|---|---|---|
| OS extrapolation | TODO | TODO | Long-term OS tail often drives QALY gain | TODO |
| PFS extrapolation | TODO | TODO | Affects time in PF state and treatment exposure | TODO |
| Curve crossing | TODO | TODO | OS should remain above PFS; implausible crossing signals model failure | TODO |
| Subgroup modelling | TODO | TODO | dMMR and pMMR may have different baseline risks and treatment effects | TODO |
| Treatment effect waning | TODO | TODO | Indefinite immunotherapy benefit may overstate long-term gain | TODO |
| Utility values | TODO | TODO | Utility source can materially affect incremental QALYs | TODO |
| Starting age | 65.4 years (mean age of dMMR/pMMR pooled cohort in KEYNOTE-868) | Trial population is younger than UK real-world patients (mean age 67.1 years based on Pennington 2016) | Affects background mortality and lifetime QALY potential | Modestly increases ICER (due to higher background mortality truncating long-term QALYs) |
| Resource use / Monitoring | Standard monitoring costs applied to both arms | Intervention arm requires additional routine safety monitoring (blood tests, thyroid, liver, glucose tests) | Underestimating routine monitoring costs artificially reduces intervention costs | Modestly increases ICER (due to higher treatment monitoring costs in the intervention arm) |
| Adverse events | Standard AE costing (Grade 3-5 AEs with >5% incidence; only anemia/hypertension costed) | Immunotherapy-related AEs (irAEs) are expensive and omitted from the costings | Omiting severe immunotherapy AEs understates treatment-related management costs | Minimally increases ICER (due to limited impact of one-off acute AE costs) |
| Subsequent therapy | TODO | TODO | Can confound OS and influence post-progression costs | TODO |

### Committee Preferred Assumptions

| Assumption | Company base case | EAG preferred | Committee preferred | My comment |
|---|---|---|---|---|
| dMMR OS curve | TODO | TODO | TODO | TODO |
| pMMR OS curve | TODO | TODO | TODO | TODO |
| dMMR PFS curve | TODO | TODO | TODO | TODO |
| pMMR PFS curve | TODO | TODO | TODO | TODO |
| Waning assumption | TODO | TODO | TODO | TODO |
| Utility source | TODO | TODO | TODO | TODO |
| Starting age | 65.4 years | 67.1 years | 67.1 years | Committee preferred 67.1 years to reflect real-world NHS patients and maintain consistency with TA963 |
| Resource use / Monitoring | Standard monitoring | Additional routine safety monitoring (EAG clinical expert estimates) | Additional routine safety monitoring | Committee preferred EAG's monitoring frequencies to reflect NHS clinical guidelines for patients on immunotherapy |
| Adverse events | >5% Grade 3-5 AEs (anemia & hypertension only) | Include Grade 3-5 AEs with >=2% incidence | Company base case (>5% Grade 3-5 AEs) | Committee accepted the company base case because including broader AEs had a minimal impact on the ICER |

## Technical Replication in R

This section starts with synthetic parameters. Replace `model_inputs` once the NICE committee papers have been abstracted.

### Placeholder Model Inputs

```{r model-inputs}
model_inputs <- list(
  horizon_months = 360,
  cycle_length_months = 1,
  annual_discount = 0.035,
  utility_pf = 0.76,
  utility_pd = 0.62,
  monthly_cost_pf_soc = 850,
  monthly_cost_pf_pembro = 6200,
  monthly_cost_pd = 1250,
  one_off_ae_cost_soc = 450,
  one_off_ae_cost_pembro = 900,
  max_pembro_months = 24,
  waning_start_month = 60,
  waning_end_month = 84
)

survival_params <- tibble::tribble(
  ~Arm, ~Endpoint, ~Shape, ~Scale,
  "Comparator", "PFS", 1.18, 15.5,
  "Comparator", "OS",  1.08, 43.0,
  "Pembrolizumab", "PFS", 1.06, 26.0,
  "Pembrolizumab", "OS",  1.02, 60.0
)

kable(survival_params, caption = "Synthetic survival inputs for template testing")
```

### Survival Functions

```{r survival-functions}
weibull_surv <- function(t, shape, scale) {
  exp(-((t / scale) ^ shape))
}

get_surv <- function(arm, endpoint, t) {
  pars <- survival_params |>
    filter(Arm == arm, Endpoint == endpoint)

  weibull_surv(t, shape = pars$Shape, scale = pars$Scale)
}

surv_df <- tibble(Month = 0:model_inputs$horizon_months) |>
  mutate(
    Comparator_PFS = get_surv("Comparator", "PFS", Month),
    Comparator_OS = get_surv("Comparator", "OS", Month),
    Pembro_PFS_base = get_surv("Pembrolizumab", "PFS", Month),
    Pembro_OS_base = get_surv("Pembrolizumab", "OS", Month)
  )

surv_long <- surv_df |>
  pivot_longer(-Month, names_to = "Curve", values_to = "Survival")

ggplot(surv_long, aes(Month, Survival, colour = Curve)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Base-case survival extrapolation",
    subtitle = "Synthetic curves pending NICE parameter abstraction",
    x = "Month",
    y = "Survival probability",
    colour = NULL
  ) +
  theme_minimal()
```

### Curve Crossing Diagnostic

```{r curve-crossing}
crossing_check <- surv_df |>
  transmute(
    Month,
    Comparator_OS_below_PFS = Comparator_OS < Comparator_PFS,
    Pembro_OS_below_PFS = Pembro_OS_base < Pembro_PFS_base
  ) |>
  summarise(
    Comparator_first_crossing = ifelse(any(Comparator_OS_below_PFS), min(Month[Comparator_OS_below_PFS]), NA),
    Pembro_first_crossing = ifelse(any(Pembro_OS_below_PFS), min(Month[Pembro_OS_below_PFS]), NA)
  )

kable(crossing_check, caption = "OS/PFS curve crossing diagnostic")
```

### EAG-Style Treatment Effect Waning Scenario

```{r treatment-effect-waning}
apply_waning <- function(trt_surv, comp_surv, month, start, end) {
  if (month <= start) return(trt_surv)

  progress <- min(1, (month - start) / (end - start))
  weight <- 1 - progress

  (trt_surv ^ weight) * (comp_surv ^ (1 - weight))
}

surv_df <- surv_df |>
  rowwise() |>
  mutate(
    Pembro_PFS_waning = apply_waning(
      Pembro_PFS_base, Comparator_PFS, Month,
      model_inputs$waning_start_month, model_inputs$waning_end_month
    ),
    Pembro_OS_waning = apply_waning(
      Pembro_OS_base, Comparator_OS, Month,
      model_inputs$waning_start_month, model_inputs$waning_end_month
    )
  ) |>
  ungroup()

waning_long <- surv_df |>
  select(Month, Pembro_OS_base, Pembro_OS_waning, Comparator_OS) |>
  pivot_longer(-Month, names_to = "Curve", values_to = "Survival")

ggplot(waning_long, aes(Month, Survival, colour = Curve)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "OS extrapolation under treatment-effect waning",
    subtitle = "Illustrative EAG-style scenario analysis",
    x = "Month",
    y = "Survival probability",
    colour = NULL
  ) +
  theme_minimal()
```

### Partitioned Survival Model Engine

```{r psm-engine}
make_trace <- function(df, pfs_col, os_col, label) {
  df |>
    transmute(
      Arm = label,
      Month,
      PF = pmax(.data[[pfs_col]], 0),
      PD = pmax(.data[[os_col]] - .data[[pfs_col]], 0),
      Dead = pmax(1 - .data[[os_col]], 0)
    ) |>
    mutate(
      Total = PF + PD + Dead,
      PF = PF / Total,
      PD = PD / Total,
      Dead = Dead / Total
    ) |>
    select(-Total)
}

trace_soc <- make_trace(surv_df, "Comparator_PFS", "Comparator_OS", "Comparator")
trace_pembro_base <- make_trace(surv_df, "Pembro_PFS_base", "Pembro_OS_base", "Pembrolizumab base case")
trace_pembro_waning <- make_trace(surv_df, "Pembro_PFS_waning", "Pembro_OS_waning", "Pembrolizumab waning scenario")

trace_all <- bind_rows(trace_soc, trace_pembro_base, trace_pembro_waning)

trace_all |>
  filter(Month <= 120) |>
  pivot_longer(c(PF, PD, Dead), names_to = "State", values_to = "Occupancy") |>
  ggplot(aes(Month, Occupancy, fill = State)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~Arm, ncol = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Partitioned survival state occupancy",
    subtitle = "First 10 years shown",
    x = "Month",
    y = "Cohort proportion",
    fill = NULL
  ) +
  theme_minimal()
```

### Cost-Utility Calculation

```{r cost-utility}
discount_factor <- function(month, annual_rate) {
  (1 + annual_rate) ^ (-month / 12)
}

evaluate_trace <- function(trace, scenario_name) {
  trace |>
    mutate(
      discount = discount_factor(Month, model_inputs$annual_discount),
      on_pembro_treatment = Arm != "Comparator" & Month <= model_inputs$max_pembro_months,
      qaly = (PF * model_inputs$utility_pf + PD * model_inputs$utility_pd) * discount / 12,
      cost = case_when(
        Arm == "Comparator" ~
          PF * model_inputs$monthly_cost_pf_soc + PD * model_inputs$monthly_cost_pd,
        on_pembro_treatment ~
          PF * model_inputs$monthly_cost_pf_pembro + PD * model_inputs$monthly_cost_pd,
        TRUE ~
          PF * model_inputs$monthly_cost_pf_soc + PD * model_inputs$monthly_cost_pd
      ) * discount
    ) |>
    summarise(
      Scenario = scenario_name,
      Total_QALY = sum(qaly),
      Total_Cost = sum(cost),
      .groups = "drop"
    )
}

econ_results <- bind_rows(
  evaluate_trace(trace_soc, "Comparator"),
  evaluate_trace(trace_pembro_base, "Pembrolizumab base case"),
  evaluate_trace(trace_pembro_waning, "Pembrolizumab waning scenario")
)

incremental_results <- econ_results |>
  filter(Scenario != "Comparator") |>
  mutate(
    Comparator_QALY = econ_results$Total_QALY[econ_results$Scenario == "Comparator"],
    Comparator_Cost = econ_results$Total_Cost[econ_results$Scenario == "Comparator"],
    Incremental_QALY = Total_QALY - Comparator_QALY,
    Incremental_Cost = Total_Cost - Comparator_Cost,
    ICER = Incremental_Cost / Incremental_QALY
  ) |>
  select(Scenario, Total_QALY, Total_Cost, Incremental_QALY, Incremental_Cost, ICER)

kable(
  incremental_results,
  digits = 2,
  caption = "Illustrative cost-effectiveness results using synthetic inputs"
)
```

## What I Could and Could Not Replicate

### Replicable with Public Information

- Overall PSM structure.
- Survival curve logic and long-term extrapolation diagnostics.
- Scenario analysis for treatment-effect waning.
- Directional effect of EAG-preferred assumptions.
- Transparent critique table linking assumption changes to ICER pressure.

### Not Replicable without Confidential Inputs

- Exact company base-case ICER.
- Confidential drug acquisition costs and patient access scheme effects.
- Full company model workbook logic.
- Exact PSA distributional assumptions.
- Patient-level trial data and all subgroup covariance structures.

## My Appraisal Conclusion

### Where I Agree with the EAG

TODO: Summarise the EAG concerns that are methodologically strongest.

### Where the Company Case Is Still Plausible

TODO: Identify where the company assumptions are defensible, especially if supported by trial follow-up, biological rationale, or clinical expert validation.

### What This Case Teaches for HEOR Consulting

This case is useful for consulting practice because it shows that technical modelling is only one layer of a NICE submission. The more important skill is understanding how evidence uncertainty becomes an appraisal risk. A model can be internally coherent and still be challenged if its long-term assumptions are not clinically credible or decision-relevant.

## Next Steps

1. Abstract the NICE guidance and committee discussion into the tables above.
2. Extract the EAG's preferred assumptions and scenario ICERs from committee papers.
3. Replace synthetic survival inputs with published or digitised estimates.
4. Decide whether to add a small pseudo-KM digitisation exercise.
5. Finalise the EAG critique narrative after comparing company, EAG, and committee positions.

---

*Disclaimer: This is an independent educational case study based on publicly available NICE documents. It is not an official appraisal and does not reproduce confidential company material.*
