---
title: "TA1092 EAG Appraisal Case: Pembrolizumab + Chemotherapy in Endometrial Cancer"
subtitle: "Survival Extrapolation, Subgroup Uncertainty, and the EAG Review Lens"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    number-sections: true
    code-annotations: below
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

set.seed(2026)
```

## Background

::: {.callout-note collapse="true"}
## Consulting Blueprint: Value-Driven Scope & Budget Control (Click to expand)

This case study functions as a high-fidelity HTA appraisal audit and model reconstruction portfolio, simulating the role of a Principal Health Economist or Senior Modeler in a consulting practice. It evaluates the NICE technology appraisal **TA1092** concerning pembrolizumab with carboplatin and paclitaxel for first-line advanced or recurrent endometrial cancer.

In commercial HEOR consulting, project delivery is strictly constrained by billable hours and the contract’s Scope of Work (SOW). A common challenge for modelers transitioning from academia is the tendency toward academic perfectionism—spending excessive hours pursuing non-reproducible parameters or trying to replicate confidential manufacturer calculations.

This project demonstrates a pragmatic, **value-driven scope management strategy** under a target budget of ~15 hours. The objective is not to reproduce the confidential manufacturer workbook or match the exact confidential ICERs (which are driven by undisclosed discounts). Instead, we deliberately restrict our scope to focus on **high-value technical nodes**—the key structural assumptions and parameters that EAGs challenge most and to which the cost-effectiveness decisions are highly sensitive.

### Project Scope Statement

| Domain | In Scope (High-Value Delivery) | Out of Scope (Academic perfectionism / Confidentiality barriers) |
|---|---|---|
| **Efficacy Data** | • Reconstructing pseudo-IPD from published trial subgroup curves (NEJM Fig 2A/B) via digitized KM estimates.<br>• Fitting candidate parametric distributions and Royston-Parmar splines in R. | • Full patient-level clinical trial database replication (confidential and unavailable). |
| **Model Structure** | • Building a de novo 3-state weekly PSM simulation engine in R (progression-free, progressed disease, death). | • Replicating the exact proprietary, multi-sheet manufacturer Excel workbook layout. |
| **HTA Adjustments** | • Coding technical solutions for HTA critiques: starting age corrections, treatment-effect waning (5–7 years), and curve-crossing capping (`pmin(PFS, OS)`). | • Re-estimating utilities or resource frequencies without public data access. |
| **Costs & Economics** | • Mapping public list prices, administration costs, acute AE management costs, and NHS-aligned subsequent therapies. | • Accessing or estimating confidential Drug Access Agreements (patient access scheme discounts). |
| **Uncertainty Analysis** | • Running parallel deterministic scenario analysis (Company vs. EAG base case) and parameter capping checks. | • Full 5,000-run probabilistic sensitivity analysis (PSA) requiring joint parameter covariance matrices. |
:::


### Clinical Context

**Disease area:** untreated primary advanced or recurrent endometrial cancer.

**Intervention:** Pembrolizumab with carboplatin and paclitaxel, followed by pembrolizumab maintenance (up to 14 additional cycles).

**Comparator:** Carboplatin with paclitaxel followed by placebo maintenance in the clinical trial setting (KEYNOTE-868), and carboplatin with paclitaxel alone without maintenance in the economic model setting representing standard NHS clinical practice.

**Key clinical trial:** KEYNOTE-868 / NRG-GY018.

**Main modelling issue:** the intervention combines immunotherapy with platinum-taxane chemotherapy. The health economic question is whether the additional cost of pembrolizumab is justified by improved progression-free survival (PFS), overall survival (OS), and quality-adjusted life years (QALYs), especially when long-term OS data are immature.

### Decision Problem

| Element | Working abstraction |
|---|---|
| Population | Adults with untreated primary advanced or recurrent endometrial cancer |
| Subgroups | dMMR/MSI-H and pMMR/MSS populations |
| Intervention | Pembrolizumab + carboplatin + paclitaxel, then pembrolizumab maintenance |
| Comparator | Carboplatin + paclitaxel followed by placebo maintenance (trial); Carboplatin + paclitaxel alone without maintenance (NHS economic model) |
| Outcomes | PFS, OS, adverse events, health-related quality of life, costs, QALYs |
| Model type | Three-state PSM |
| NICE decision tension | Whether extrapolated survival benefit is clinically plausible and cost-effective across molecular subgroups |

### Evidence & Appraisal Registry

To maintain a concise background narrative, the complete HTA literature registry and parameter traceability mapping are detailed in the [HTA Evidence & Reference Registry](#appendix-a-hta-parameter-traceability-reference-registry) in the Appendix.

::: {.callout-note collapse="true"}
## Modeler's Initial Risk Screening & HTA Audit Focus (Click to expand)

When evaluating a manufacturer's cost-effectiveness submission for an oncology immunotherapy combination, an experienced HTA model auditor screens for standard systemic risk nodes. In this appraisal, our initial risk-screening checklist targets four critical domains:

1. **Survival Extrapolation & Tail Plausibility (High Risk)**:
   - *Audit Target*: When clinical trial follow-up is relatively short (median < 2 years), the choice of parametric models is highly sensitive. We must verify if the long-term overall survival (OS) tails are clinically plausible or if they project unrealistic lifetime "immortal patient" fractions.
   - *Key Focus*: Subgroup sample sizes (dMMR vs. pMMR) and the risk of PFS and OS curves crossing in the long term, which violates logical consistency.

2. **Immunotherapy Treatment Duration & Efficacy Waning (High Risk)**:
   - *Audit Target*: Immunotherapies like pembrolizumab are typically capped (e.g., a 2-year stopping rule). The model must not assume that treatment benefit persists indefinitely after treatment discontinuation.
   - *Key Focus*: Implementing and testing treatment-effect waning scenarios (e.g., waning starting at 2 or 5 years and converging to control hazards) to evaluate the stability of the ICER.

3. **Subgroup Applicability & Data Maturity (Medium Risk)**:
   - *Audit Target*: The clinical and economic benefits may differ significantly between mismatch repair deficient (dMMR) and proficient (pMMR) cohorts. We need to check if separate models are statistically stable and if baseline characteristics align with the UK target population.

4. **Resource Cost & Subsequent Therapy Alignments (Medium Risk)**:
   - *Audit Target*: Subsequent treatment mixes in the comparator arm must reflect real-world NHS standard pathways rather than non-reimbursed trial settings, as differences in active subsequent therapies can distort the cost delta.
:::

## EAG Critique Lens

### How the EAG Reviews This Type of Case

The EAG is not only checking whether the model runs. It is asking whether the company's assumptions are:

- internally consistent with trial evidence;
- externally plausible against clinical knowledge;
- aligned with NHS treatment pathways;
- transparent enough for committee decision-making;
- conservative enough when evidence is immature.

This distinction is important for the portfolio: a health economist in consulting must be able to understand both the company logic and the EAG logic.

### Summary of NICE Decision Tensions

Based on the HTA dossier review, the appraisal debate centers on translating immature trial survival data into a 30-year extrapolation model. The key methodological debates include:

- **Survival Tail Plausibility**: Standard parametric models preferred by the company yielded optimistic long-term tails. The EAG preferred flexible spline models to better capture hazard dynamics.
- **Curve-Crossing Implausibility**: The EAG's spline-based placebo PFS model crossed and exceeded the active pembrolizumab PFS model in the long-term tail. This logical inconsistency required a structural capping mechanism.
- **Efficacy Waning**: The company assumed infinite treatment benefit post-discontinuation, which the EAG challenged. A compromise waning window was introduced to transition active treatment hazards to control hazards.
- **Utility Consistency**: To avoid confounding utility estimates by mixing clinical trial sources (KEYNOTE-158 and KEYNOTE-B21), the EAG advocated for exclusive use of KEYNOTE-158.

### Summary of the 10 EAG Key Issues

The EAG identified ten key clinical, methodological, and economic issues in the company's submission:

1. **Representativeness of post-hoc all-comer analysis**: Post-hoc pooling of trial cohorts may not represent the trial's original design and statistical analysis plan.
2. **Subgroup limitations**: Site of recurrence (local vs. metastatic) and debulking surgery history were not systematically collected in KEYNOTE-868.
3. **Short trial follow-up**: Immaturity of trial data limits the robustness of long-term survival effectiveness conclusions.
4. **Subgroup-restricted HRQoL**: Health-related quality of life was collected only in the pMMR cohort, leaving the dMMR cohort without trial-specific utility data.
5. **Uncertain overall survival benefit**: The company's survival models potentially overestimate long-term immunotherapy tail benefits.
6. **Understated baseline age**: The modeled baseline age (65.4 years) was younger than standard UK patients (67.1 years), affecting background mortality.
7. **Representativeness of health state utilities**: Mapped utilities from KEYNOTE-158 (second-line dMMR) may not represent the first-line all-comer population.
8. **Underestimated safety monitoring resources**: Safety monitoring costs (blood, liver, thyroid tests) for patients on active immunotherapy were omitted.
9. **Incomplete adverse event costing**: High >=5% incidence threshold omitted expensive immunotherapy-related adverse events.
10. **Misaligned subsequent treatment mix**: Subsequent therapies in the comparator arm did not align with NHS clinical guidelines.

### EAG Challenge Log

<details>
<summary><b>Click to expand: Detailed EAG Challenge Log & ICER Impact Map</b></summary>

| Issue | Company position | EAG concern | Why the EAG cares | Potential ICER direction |
|---|---|---|---|---|
| OS extrapolation | Standard parametric curves (log-logistic/gamma) based on trial data fit | Overestimated overall survival tails due to short trial follow-up; preferred flexible splines for pMMR | Long-term OS tail is the major driver of lifetime QALY gains | Significantly increases the ICER |
| PFS extrapolation | Generalised gamma (dMMR) and 2-piece generalised gamma (pMMR) | Prone to tail overestimation; preferred log-logistic/generalised gamma (dMMR) and flexible splines (pMMR) | Determines duration of progression-free utility and active treatment exposure | Increases the ICER |
| Curve crossing | Long-term crossing is clinically implausible; active treatment should always be superior | Crossing is a tail extrapolation uncertainty artifact and doesn't invalidate observed trial fit | PFS must not cross treatment arms; uncapped models are structurally flawed | Modestly reduces the ICER when placebo PFS is capped |
| Subgroup modelling | Modeled dMMR and pMMR cohorts separately using trial subgroup baseline inputs | Supported separate models, but highlighted dMMR immaturity due to small sample size | Treatment effect and background risks differ fundamentally by molecular status | Results in distinct ICERs for dMMR (cost-effective) and pMMR (marginal) |
| Treatment effect waning | No waning applied (assumed lifetime immunotherapy benefit) | Short trial follow-up cannot justify indefinite post-discontinuation benefit; preferred 5-7 year waning | Overestimates long-term survival gain in the extrapolated tail | Significantly increases the ICER |
| Utility values | Used mixed sources (KEYNOTE-158 for PD, KEYNOTE-B21 for PF health states) | Opposed mixing utility sources; preferred using KEYNOTE-158 exclusively for all states | Health state utility values directly scale cumulative QALY gains | Modestly to significantly increases the ICER |
| Starting age | 65.4 years (mean age of dMMR/pMMR pooled cohort in KEYNOTE-868) | Trial population is younger than UK real-world patients (mean age 67.1 years based on Pennington 2016) | Affects background mortality and lifetime QALY potential | Modestly increases ICER (due to higher background mortality truncating long-term QALYs) |
| Resource use / Monitoring | Standard monitoring costs applied to both arms | Intervention arm requires additional routine safety monitoring (blood tests, thyroid, liver, glucose tests) | Underestimating routine monitoring costs artificially reduces intervention costs | Modestly increases ICER (due to higher treatment monitoring costs in the intervention arm) |
| Adverse events | Standard AE costing (Grade 3-5 AEs with >5% incidence; only anemia/hypertension costed) | Immunotherapy-related AEs (irAEs) are expensive and omitted from the costings | Omiting severe immunotherapy AEs understates treatment-related management costs | Minimally increases ICER (due to limited impact of one-off acute AE costs) |
| Subsequent therapy | Company assumed a portion of the CT arm receives pembrolizumab monotherapy post-progression | Pembrolizumab monotherapy is not NHS standard 2L care for pMMR patients; should be pembrolizumab + lenvatinib | Standardizing subsequent care in the comparator arm directly influences comparator cost levels | Decreases ICER (due to subsequent therapy costs in the comparator arm narrowing the cost delta) |
</details>

### Committee Preferred Assumptions

<details>
<summary><b>Click to expand: Company vs. EAG vs. Committee Preferred Assumptions Table</b></summary>

| Assumption | Company base case | EAG preferred | Committee preferred | My comment |
|---|---|---|---|---|
| dMMR OS curve | Log-logistic (pembro) / Exponential (placebo) | Log-logistic (pembro) / Exponential (placebo) | Log-logistic (pembro) / Exponential (placebo) | Consensus reached. Both parties agreed these represented the best fit and clinical plausibility for the dMMR cohort. |
| pMMR OS curve | Log-logistic (pembro) / Gamma (placebo) | 1-knot normal spline (pembro) / 1-knot hazard spline (placebo) | 1-knot normal spline (pembro) / 1-knot hazard spline (placebo) | Committee preferred the EAG's spline models to capture the changing hazard observed in the trial. |
| dMMR PFS curve | Generalised gamma (pembro) / 2-piece gamma (placebo, 27-wk cut) | Log-logistic (pembro) / Generalised gamma (placebo) | Log-logistic (pembro) / Generalised gamma (placebo) | Committee adopted the EAG's preferred curves due to improved fit and tail plausibility. |
| pMMR PFS curve | 2-piece generalised gamma (pembro, 37-wk cut) / 1-knot odds spline (placebo) | 1-knot hazard spline (pembro) / 2-knot hazard spline (placebo) | 1-knot hazard spline (pembro) / 2-knot hazard spline (placebo, capped) | Sided with the company's implausibility argument: mandated capping placebo PFS at pembro PFS in the tail. |
| Waning assumption | No waning (lifetime treatment effect) | Gradual waning 5 to 7 years (all patients) | Gradual waning 5 to 7 years (non-complete responders only) | Plausible compromise. Waning applied between 5-7 years only for patients without complete response. |
| Utility source | Mixed (KEYNOTE-158 for PD, KEYNOTE-B21 for PF) | KEYNOTE-158 (all states and subgroups) | KEYNOTE-158 (all states and subgroups) | Committee preferred keeping a single consistent trial source to avoid mixing different patient populations. |
| Starting age | 65.4 years | 67.1 years | 67.1 years | Committee preferred 67.1 years to reflect real-world NHS patients and maintain consistency with TA963 |
| Resource use / Monitoring | Standard monitoring | Additional routine safety monitoring (EAG clinical expert estimates) | Additional routine safety monitoring | Committee preferred EAG's monitoring frequencies to reflect NHS clinical guidelines for patients on immunotherapy |
| Adverse events | >5% Grade 3-5 AEs (anemia & hypertension only) | Include Grade 3-5 AEs with >=2% incidence | Company base case (>5% Grade 3-5 AEs) | Committee accepted the company base case because including broader AEs had a minimal impact on the ICER |
| Subsequent therapy | Pembrolizumab monotherapy allowed in CT arm | Reallocate pembrolizumab monotherapy to pembrolizumab + lenvatinib | Reallocated subsequent therapy for pMMR patients | Committee preferred aligning subsequent therapies with NHS clinical guidelines for pMMR patients, which reduced the cost delta and decreased the ICER |
</details>

## Workflow Setup & Exploratory Analysis

::: {.column-margin}
**Time Log: 2.5 hours**
- Digitised sub-group curves from NEJM Figure 2A/B.
- Built and tidied the long-format IPD dataset for R processing.
:::


In this chapter, we load the required packages, set up local file paths, import the reconstructed individual patient-level data (pseudo-IPD), and perform exploratory curve inspections.

### Load Packages & Settings

This block loads the R packages used in this workflow.

```{r load-packages}
library(survival)
library(flexsurv)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
library(knitr)

tibble(
  package = c("survival", "flexsurv", "dplyr", "tidyr", "purrr", "readr", "ggplot2", "knitr"),
  role = c(
    "Create survival objects and Kaplan-Meier estimates",
    "Fit parametric survival models",
    "Data manipulation",
    "Unnest list-columns",
    "Loop over curves and distributions",
    "Read and write CSV files",
    "Plot KM curves and extrapolations",
    "Display readable tables"
  )
) |>
  kable()
```


### Define File Paths

The main input is the combined long-format pseudo-IPD file.

```{r define-paths}
base_dir <- normalizePath(".", mustWork = TRUE)
input_path <- file.path(base_dir, "survival_inputs", "survival_ipd_long.csv")
output_dir <- file.path(base_dir, "survival_models")
plot_dir <- file.path(output_dir, "plots")

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

tibble(
  object = c("base_dir", "input_path", "output_dir", "plot_dir"),
  path = c(base_dir, input_path, output_dir, plot_dir),
  exists = c(dir.exists(base_dir), file.exists(input_path), dir.exists(output_dir), dir.exists(plot_dir))
) |>
  kable()
```


### Read & Transform Reconstructed Input IPD

The dataset is reconstructed pseudo-IPD. Each row is one pseudo-patient.

`status = 1` means the event happened. For PFS this means progression/death; for OS this means death.

```{r read-input-data}
ipd <- read_csv(input_path, show_col_types = FALSE)

ipd |>
  head(3) |>
  knitr::kable()
```

```{r transform-input-data}
ipd <- ipd |>
  mutate(
    subgroup = factor(subgroup, levels = c("dMMR", "pMMR")),
    endpoint = factor(endpoint, levels = c("PFS", "OS")),
    treatment = factor(treatment, levels = c("placebo_chemo", "pembro_chemo")),
    status = as.integer(status),
    time_months = as.numeric(time_months),
    curve_id = paste(subgroup, endpoint, treatment, sep = "_")
  )

ipd |>
  count(subgroup, endpoint, treatment, status_label) |>
  arrange(subgroup, endpoint, treatment, status_label) |>
  kable()
```


### Exploratory Curve Inspection & KM Plotting

#### Inspecting pMMR PFS Pembrolizumab Cohort

Before fitting every curve, inspect one explicitly named curve: **pMMR PFS, pembro arm**.

```{r inspect-pmmr-pfs-pembro-ipd}
pmmr_pfs_pembro_ipd <- ipd |>
  filter(subgroup == "pMMR", endpoint == "PFS", treatment == "pembro_chemo")

pmmr_pfs_pembro_ipd |>
  head(3) |>
  knitr::kable()

tibble(
  curve = "pMMR PFS pembrolizumab + chemotherapy",
  n = nrow(pmmr_pfs_pembro_ipd),
  events = sum(pmmr_pfs_pembro_ipd$status),
  censored = sum(pmmr_pfs_pembro_ipd$status == 0),
  max_follow_up_months = max(pmmr_pfs_pembro_ipd$time_months)
) |>
  kable()
```


#### Creating the Survival Object

`Surv(time, status)` is the standard R representation of right-censored survival data.

```{r create-surv-object}
pmmr_pfs_pembro_surv <- Surv(pmmr_pfs_pembro_ipd$time_months, pmmr_pfs_pembro_ipd$status)

tibble(
  raw_time = pmmr_pfs_pembro_ipd$time_months[1:5],
  raw_status = pmmr_pfs_pembro_ipd$status[1:5],
  surv_object = as.character(pmmr_pfs_pembro_surv[1:5])
) |>
  kable()
```


#### Reconstructed Kaplan-Meier Curve

This is the Kaplan-Meier curve reconstructed from pseudo-IPD.

```{r plot-pmmr-pfs-pembro-km}
pmmr_pfs_pembro_km <- survfit(Surv(time_months, status) ~ 1, data = pmmr_pfs_pembro_ipd)

pmmr_pfs_pembro_km_df <- tibble(
  time_months = c(0, pmmr_pfs_pembro_km$time),
  survival = c(1, pmmr_pfs_pembro_km$surv)
)

ggplot(pmmr_pfs_pembro_km_df, aes(time_months, survival)) +
  geom_step(linewidth = 0.8, colour = "#008B8B") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Reconstructed KM: pMMR PFS, pembro arm",
    x = "Months",
    y = "Survival probability"
  ) +
  theme_minimal()
```


## Parametric Survival Fitting & Diagnostics

::: {.column-margin}
**Time Log: 5.0 hours (Core Extrapolation)**
- Developed the `flexsurv` model fitting wrappers and batch parameters extraction.
- Plotted Extrapolation-Zones and the Distribution Fan Plot to evaluate long-run tail divergence.
:::


This chapter outlines the standard parametric survival fitting workflow. We fit seven candidate distributions (exponential, Weibull, Gompertz, log-normal, log-logistic, gamma, and generalized gamma) to each curve, assess fit summaries via information criteria (AIC and BIC), and extract the fitted parameters.

### Model Setup & Safe Fitting Wrapper

#### Candidate Parametric Distributions

These are common survival distributions used in oncology HTA submissions and EAG critiques.

```{r candidate-distributions}
candidate_dists <- c(
  "exponential" = "exp",
  "weibull" = "weibull",
  "gompertz" = "gompertz",
  "lognormal" = "lnorm",
  "loglogistic" = "llogis",
  "gamma" = "gamma",
  "generalized_gamma" = "gengamma"
)

tibble(
  display_name = names(candidate_dists),
  flexsurv_name = unname(candidate_dists)
) |>
  kable()
```


#### Weibull Model Fitting Demonstration

This block fits a Weibull model to the pMMR PFS pembro curve.

```{r fit-one-weibull}
pmmr_pfs_pembro_weibull <- flexsurvreg(
  Surv(time_months, status) ~ 1,
  data = pmmr_pfs_pembro_ipd,
  dist = "weibull"
)

as.data.frame(pmmr_pfs_pembro_weibull$res) |>
  tibble::rownames_to_column("parameter") |>
  as_tibble() |>
  select(parameter, estimate = est, lower_95 = `L95%`, upper_95 = `U95%`, se) |>
  kable(digits = 4)
```


#### Safe Fitting Wrapper Function

Some distributions can fail for sparse or immature OS data. This function catches failures and returns a row explaining what happened.

```{r define-fitting-function}
`%||%` <- function(x, y) if (is.null(x)) y else x

fit_one_distribution <- function(dat, dist_label, dist_name) {
  tryCatch(
    {
      fit <- flexsurvreg(Surv(time_months, status) ~ 1, data = dat, dist = dist_name)
      n_par <- nrow(fit$res)
      fit_aic <- as.numeric(fit$AIC %||% NA_real_)
      fit_loglik <- as.numeric(fit$loglik %||% NA_real_)

      tibble(
        dist_label = dist_label,
        dist_name = dist_name,
        fit = list(fit),
        converged = TRUE,
        error = NA_character_,
        aic = fit_aic,
        bic = -2 * fit_loglik + log(nrow(dat)) * n_par,
        loglik = fit_loglik,
        n = nrow(dat),
        events = sum(dat$status)
      )
    },
    error = function(e) {
      tibble(
        dist_label = dist_label,
        dist_name = dist_name,
        fit = list(NULL),
        converged = FALSE,
        error = conditionMessage(e),
        aic = NA_real_,
        bic = NA_real_,
        loglik = NA_real_,
        n = nrow(dat),
        events = sum(dat$status)
      )
    }
  )
}

tibble(function_created = "fit_one_distribution()", purpose = "Fit one distribution and return diagnostics") |>
  kable()
```


#### Details of the Wrapper Logic

This function is a small wrapper around `flexsurvreg()`. The reason to write a wrapper is that we want to fit many distributions across many curves. Some fits may fail, especially for immature OS curves with few events. If one fit fails, we do not want the whole document to stop.

The function has three inputs:

```r
fit_one_distribution(dat, dist_label, dist_name)
```

- `dat`: the pseudo-IPD for one curve, for example `pmmr_pfs_pembro_ipd`.
- `dist_label`: the readable name used in output tables, for example `"weibull"`.
- `dist_name`: the name required by `flexsurvreg()`, for example `"weibull"` or `"llogis"`.

The helper:

```r
`%||%` <- function(x, y) if (is.null(x)) y else x
```

means: if `x` is missing or `NULL`, use `y` instead. Here it prevents the code from breaking if a fitted object does not contain `AIC` or `loglik`.

The core model fit is:

```r
fit <- flexsurvreg(Surv(time_months, status) ~ 1, data = dat, dist = dist_name)
```

This means:

- use `time_months` and `status` from `dat`;
- fit one survival curve, because the right-hand side is `~ 1`;
- use the parametric distribution named in `dist_name`.

The object `fit` is a fitted survival model. It contains parameter estimates, log-likelihood, AIC, and other information. We store it inside a tibble using:

```r
fit = list(fit)
```

This creates a **list-column**. Without `list()`, a complex model object cannot fit cleanly inside one table cell.

The successful-output part returns one row:

```r
tibble(
  dist_label = dist_label,
  dist_name = dist_name,
  fit = list(fit),
  converged = TRUE,
  error = NA_character_,
  aic = fit_aic,
  bic = -2 * fit_loglik + log(nrow(dat)) * n_par,
  loglik = fit_loglik,
  n = nrow(dat),
  events = sum(dat$status)
)
```

Key fields:

- `converged = TRUE`: the model fitted successfully.
- `error = NA_character_`: there was no error message.
- `aic`: model fit criterion reported by `flexsurv`.
- `bic`: calculated manually using `-2 * logLik + log(n) * number_of_parameters`.
- `n`: number of pseudo-patients in this curve.
- `events`: number of observed events.

The `tryCatch()` part is the safety mechanism:

```r
tryCatch(
  { successful fitting code },
  error = function(e) { failure output }
)
```

If `flexsurvreg()` fails, the function still returns a row. That row has:

- `fit = list(NULL)`
- `converged = FALSE`
- `error = conditionMessage(e)`
- `aic`, `bic`, `loglik` as `NA`

This is useful for modelling workflow QA. Instead of silently losing failed fits, the output table records exactly which distribution failed and why.

In HEOR work, this matters because sparse OS data can make complex models unstable. A reviewer or EAG would want to know not only which model had the best AIC, but also whether some candidate models failed, behaved oddly, or were unsuitable for extrapolation.


### Fitting Batch Curves & Extracting Parameters

#### Comparing Candidate Fits for One Curve

For the pMMR PFS pembro curve, fit all candidate distributions and rank by AIC.

```{r compare-pmmr-pfs-pembro-fits}
pmmr_pfs_pembro_fit_comparison <- map2_dfr(
  names(candidate_dists),
  candidate_dists,
  ~fit_one_distribution(pmmr_pfs_pembro_ipd, .x, .y)
)

pmmr_pfs_pembro_fit_comparison |>
  select(dist_label, converged, n, events, loglik, aic, bic, error) |>
  arrange(aic) |>
  kable(digits = 3)
```


#### Fitting Every Curve

Now repeat the same process for all subgroup-endpoint-treatment curves.

```{r fit-all-curves}
curve_groups <- ipd |>
  group_by(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id) |>
  group_split()

fit_tbl <- map_dfr(curve_groups, function(dat) {
  meta <- dat |>
    slice(1) |>
    select(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id)

  res <- map2_dfr(
    names(candidate_dists),
    candidate_dists,
    ~fit_one_distribution(dat, .x, .y)
  )

  bind_cols(res, meta[rep(1, nrow(res)), ])
})

fit_tbl |>
  count(subgroup, endpoint, treatment, converged, name = "number_of_candidate_distributions") |>
  kable()
```


#### Creating the Fit Summary Table

AIC is useful for first-pass comparison, but not sufficient for final HTA model choice.

```{r fit-summary}
fit_summary <- fit_tbl |>
  select(
    subgroup, endpoint, treatment, arm_role, treatment_label, curve_id,
    dist_label, dist_name, converged, error, n, events, loglik, aic, bic
  ) |>
  arrange(subgroup, endpoint, treatment, aic)

top3_fit_summary <- fit_summary |>
  filter(converged) |>
  group_by(subgroup, endpoint, treatment) |>
  arrange(aic, .by_group = TRUE) |>
  slice_head(n = 3) |>
  ungroup() |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, bic)

top3_fit_summary |>
  kable(digits = 2)
```


#### Extract Model Parameters

The object `parameter_tbl` stores parameter estimates from **every converged model** across all subgroups, endpoints, treatment arms, and distributions.

The table displayed below is only an example: it filters `parameter_tbl` to show **pMMR PFS, pembro arm** parameters.

```{r extract-parameters}
extract_parameters <- function(fit) {
  if (is.null(fit)) {
    return(tibble(parameter = character(), estimate = numeric(), se = numeric()))
  }

  as.data.frame(fit$res) |>
    tibble::rownames_to_column("parameter") |>
    as_tibble() |>
    transmute(parameter, estimate = est, se = se)
}

parameter_tbl <- fit_tbl |>
  filter(converged) |>
  mutate(parameters = map(fit, extract_parameters)) |>
  select(subgroup, endpoint, treatment, curve_id, dist_label, parameters) |>
  unnest(parameters)

parameter_tbl |>
  filter(subgroup == "pMMR", endpoint == "PFS", treatment == "pembro_chemo") |>
  arrange(dist_label, parameter) |>
  transmute(
    distribution = dist_label,
    parameter,
    estimate = round(estimate, 4),
    std_error = round(se, 4)
  ) |>
  kable(caption = "Example display: parameter estimates for pMMR PFS, pembro arm")
```


### Model Ranking & First-Pass AIC Selection

This is a mechanical first-pass choice. Later we will critique the tails.

```{r select-best-fit}
best_fits <- fit_tbl |>
  filter(converged, !is.na(aic)) |>
  group_by(subgroup, endpoint, treatment) |>
  slice_min(aic, n = 1, with_ties = FALSE) |>
  ungroup()

best_fits |>
  select(subgroup, endpoint, treatment, best_dist = dist_label, n, events, aic, bic) |>
  arrange(subgroup, endpoint, treatment) |>
  kable(digits = 2)
```


## Extrapolation Zone & Data Maturity Critique (Issue 3)

This chapter demonstrates how the short follow-up in clinical trials can create substantial uncertainty during long-term overall survival and progression-free survival extrapolations. We present the EAG's Issue 3 critique framework, visualize the extrapolation zones, and evaluate parametric tail divergence at key clinical landmarks.

### Predicting Survival Over 30-Year Horizon

#### Extrapolating Selected Distributions

The extrapolation horizon is 360 months, or 30 years.

The object `extrapolated` still stores every best-fit curve over the full monthly time grid. To keep the document readable, the displayed output below is only a small sanity check for **pMMR PFS, pembro arm**.

```{r extrapolate-best-fit}
time_grid <- tibble(time_months = seq(0, 360, by = 1))

predict_survival <- function(fit, times) {
  s <- summary(fit, t = times, type = "survival")

  if (is.data.frame(s)) {
    return(as.numeric(s$est))
  }

  if (is.list(s) && length(s) == 1 && is.data.frame(s[[1]])) {
    return(as.numeric(s[[1]]$est))
  }

  map_dbl(s, ~.x$est)
}

extrapolated <- best_fits |>
  mutate(
    pred = map(
      fit,
      ~tibble(
        time_months = time_grid$time_months,
        survival = predict_survival(.x, time_grid$time_months)
      )
    )
  ) |>
  select(
    subgroup, endpoint, treatment, arm_role, treatment_label, curve_id,
    best_dist = dist_label, aic, bic, pred
  ) |>
  unnest(cols = c(pred))

extrapolated |>
  filter(
    subgroup == "pMMR",
    endpoint == "PFS",
    treatment == "pembro_chemo",
    time_months %in% c(12, 24, 60, 120)
  ) |>
  arrange(subgroup, endpoint, treatment, time_months) |>
  select(subgroup, endpoint, treatment, best_dist, time_months, survival) |>
  kable(digits = 4, caption = "Sanity-check display: pMMR PFS pembro extrapolated survival")
```


#### Visualizing Long-Run Extrapolations

Tables are not the best way to inspect extrapolated survival. For HTA work, the curve shape is more informative. Here we show two examples:

- pMMR PFS, because it is technically important for the partitioned survival model.
- dMMR OS, because it is immature and therefore useful for thinking about EAG-style tail critique.

```{r plot-selected-long-run-extrapolations}
selected_long_run <- extrapolated |>
  filter(
    (subgroup == "pMMR" & endpoint == "PFS") |
      (subgroup == "dMMR" & endpoint == "OS"),
    time_months <= 120
  ) |>
  mutate(
    curve = paste(subgroup, endpoint),
    treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
    line_label = paste0(treatment_short, " (", best_dist, ")")
  )

ggplot(selected_long_run, aes(time_months, survival, colour = line_label)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~curve) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 24)) +
  labs(
    title = "Selected AIC-best long-run extrapolations",
    subtitle = "Legend shows treatment arm and AIC-selected distribution",
    x = "Months",
    y = "Survival probability",
    colour = "Arm (distribution)"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.width = unit(1.2, "lines")
  )
```


#### Comparing Reconstructed KM with Parametric Best Fits

This plot is where HTA judgement begins. In this specific pMMR PFS example, the AIC-selected curves do not show an obvious visual problem. The point of the plot is more modest: it lets us check whether the fitted curve tracks the reconstructed KM reasonably before we start interpreting the long-term extrapolation.

The broader EAG-style warning is that a curve can have the best AIC and still become implausible later. Examples would include:

- a PFS curve that remains high for decades despite advanced cancer;
- an OS curve that falls below PFS, which is structurally impossible in a partitioned survival model;
- a treatment curve with an indefinitely widening benefit when clinical experts expect treatment effect waning;
- a curve tail that implies long-term survival far above age- and disease-plausible expectations.

Those problems are not necessarily visible in this pMMR PFS plot; they are the kinds of issues we will check in later model critique.

```{r plot-pmmr-pfs-best-fit}
km_points <- ipd |>
  group_by(subgroup, endpoint, treatment, arm_role, treatment_label, curve_id) |>
  group_modify(~{
    km <- survfit(Surv(time_months, status) ~ 1, data = .x)
    tibble(time_months = c(0, km$time), survival = c(1, km$surv))
  }) |>
  ungroup()

plot_data <- extrapolated |>
  filter(subgroup == "pMMR", endpoint == "PFS", time_months <= 120) |>
  mutate(
    source = "Parametric best fit",
    treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
    curve_label = paste0(treatment_short, " (", best_dist, ")")
  ) |>
  select(subgroup, endpoint, treatment, treatment_short, curve_label, time_months, survival, source) |>
  bind_rows(
    km_points |>
      filter(subgroup == "pMMR", endpoint == "PFS") |>
      mutate(
        source = "Reconstructed KM",
        treatment_short = if_else(treatment == "pembro_chemo", "Pembro", "Placebo"),
        curve_label = treatment_short
      ) |>
      select(subgroup, endpoint, treatment, treatment_short, curve_label, time_months, survival, source)
  )

ggplot(plot_data, aes(time_months, survival, colour = curve_label)) +
  geom_step(
    data = filter(plot_data, source == "Reconstructed KM"),
    linewidth = 0.85,
    alpha = 0.55
  ) +
  geom_line(
    data = filter(plot_data, source == "Parametric best fit"),
    linewidth = 1.15
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 12)) +
  labs(
    title = "pMMR PFS: reconstructed KM vs AIC-selected extrapolation",
    subtitle = "Step curves are reconstructed KM; solid smooth curves are parametric best fits",
    x = "Months",
    y = "Survival probability",
    colour = "Curve"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.width = unit(1.2, "lines")
  )
```


### EAG Issue 3 Framework & Extrapolation-Zone Plotting

#### Introduction to Issue 3 Demonstration

The EAG's Issue 3 is not just saying that longer follow-up would be preferable. The practical concern is that KEYNOTE-868 provides relatively early survival evidence, while the economic model requires long-term PFS and OS predictions.

In the EAG report, several values are redacted in the Issue 3 table. Later consultation comments in the same committee papers refer to the median follow-up as **16.3 months**. I use that value below as a working interpretation, while noting that the value is redacted in the Issue 3 table itself.

This section uses **pMMR OS, pembro arm** as the worked example. OS is usually one of the most important drivers of life years and QALYs in oncology models, and short follow-up is especially important when OS benefit is immature.


#### EAG Logic Behind Issue 3

Before plotting anything, it is useful to translate the EAG's concern into modelling terms.

```{r issue3-eag-logic-table}
issue3_eag_logic <- tibble::tribble(
  ~`EAG point`, ~`Meaning for modelling`,
  "Median follow-up was about 16.3 months", "Most direct trial evidence covers early survival only",
  "Survival estimates for 2 years or more rely on extrapolation", "Beyond roughly 24 months, estimates are increasingly model-driven",
  "PFS and OS estimates extend out to 20 years", "The model must predict far beyond observed follow-up",
  "Median OS was not reached in the pembrolizumab arm", "The OS benefit is immature and long-term benefit is uncertain",
  "Active treatment lasts up to about 2.2 years", "Follow-up is short relative to the full treatment and post-treatment period"
)

issue3_eag_logic |>
  kable(
    caption = "Issue 3 translated into modelling questions",
    align = "l"
  )
```

The important point is that the EAG did not apply a simple universal rule such as "follow-up below X months is unacceptable". The concern comes from the combination of short follow-up, immature OS, and a model that needs to estimate survival far beyond the observed trial period.

Two judgement points are especially important here.

First, **median survival** is a maturity signal. If median OS has not been reached in the intervention arm, then fewer than half of patients in that arm have died by the data cut. That is clinically encouraging, but it also means the long-term OS curve is not well observed. The model must infer the later part of OS from limited event data.

Second, **median follow-up should be interpreted against treatment duration**. Pembrolizumab treatment in this regimen can last for around 2.2 years, but the median follow-up referred to in the committee papers is around 16.3 months. This means the median patient had not yet been followed for as long as the maximum planned active treatment period. From an EAG perspective, that creates uncertainty not only about survival during treatment, but also about what happens after treatment stops.

```{r issue3-maturity-checks}
issue3_maturity_checks <- tibble::tribble(
  ~`Maturity check`, ~`What it asks`, ~`Why it matters for Issue 3`,
  "Median OS reached?", "Has at least half the arm experienced the OS event?", "If median OS is not reached, long-term OS benefit is inferred rather than directly observed.",
  "Median follow-up vs active treatment duration", "Has the median patient been observed through the full planned treatment period and beyond?", "If median follow-up is shorter than maximum treatment duration, there is limited evidence on post-treatment survival and treatment-effect durability.",
  "2-year boundary", "Are estimates beyond roughly 24 months still supported by enough observed data?", "The EAG considered 2 years+ estimates highly susceptible to extrapolation error.",
  "20-year projection", "How far beyond trial follow-up does the economic model project?", "The longer the projection, the more long-term QALYs depend on model structure and tail assumptions."
)

issue3_maturity_checks |>
  kable(
    caption = "Issue 3 maturity checks used to interpret short follow-up",
    align = "l"
  )
```


#### Observed Evidence Versus the 20-Year Projection

The first technical check is descriptive: how much reconstructed follow-up do we have, and how does it compare with the 20-year projection horizon discussed by the EAG?

```{r issue3-observed-vs-modelled-horizon}
issue3_projection_horizon_months <- 240
issue3_median_follow_up_months <- 16.3
issue3_eag_two_year_boundary <- 24
issue3_active_treatment_months <- (30 + 84) * 7 / 30.4375

issue3_horizon_summary <- ipd |>
  group_by(subgroup, endpoint, treatment, treatment_label) |>
  summarise(
    n = n(),
    events = sum(status),
    censored = sum(status == 0),
    max_reconstructed_follow_up_months = max(time_months),
    model_horizon_months = issue3_projection_horizon_months,
    extrapolation_ratio = model_horizon_months / max_reconstructed_follow_up_months,
    .groups = "drop"
  ) |>
  arrange(endpoint, subgroup, treatment)

issue3_horizon_summary |>
  transmute(
    curve = paste(
      subgroup,
      endpoint,
      if_else(treatment == "pembro_chemo", "Pembro", "Placebo")
    ),
    events = paste0(events, "/", n),
    `max FU, mo` = round(max_reconstructed_follow_up_months, 1),
    `20-year horizon, mo` = model_horizon_months,
    `observed share of 20y` = paste0(round(100 * max_reconstructed_follow_up_months / model_horizon_months, 1), "%"),
    `20y / max FU` = paste0(round(extrapolation_ratio, 1), "x")
  ) |>
  kable(
    caption = "Observed reconstructed follow-up versus 20-year projection horizon",
    align = "l"
  )
```

This table is not a threshold test. It only shows scale. The more important question is not whether the ratio is "too high" in isolation, but whether the economic conclusion depends on the extrapolated part of the curve.

For this portfolio, the table sets up the next two checks:

- where the observed evidence ends and the modelled tail begins;
- whether different plausible parametric curves produce different long-term survival.


#### Extrapolation-Zone Plot

The next plot turns Issue 3 into a visual modelling problem. It marks:

- **16.3 months**: median follow-up referred to in the committee papers;
- **24 months**: the EAG's practical boundary for "2 years+" estimates;
- **26.2 months**: approximate maximum active treatment duration based on 30 weeks plus 84 weeks;
- **24 months onward**: the region where estimates are increasingly driven by extrapolation.

```{r issue3-extrapolation-zone-plot}
issue3_curve_id <- "pMMR_OS_pembro_chemo"

issue3_km <- km_points |>
  filter(curve_id == issue3_curve_id)

issue3_best_fit <- extrapolated |>
  filter(curve_id == issue3_curve_id, time_months <= 240)

issue3_observed_boundary <- max(
  ipd |>
    filter(curve_id == issue3_curve_id) |>
    pull(time_months)
)

issue3_best_dist <- issue3_best_fit |>
  distinct(best_dist) |>
  pull(best_dist)

ggplot() +
  annotate(
    "rect",
    xmin = issue3_eag_two_year_boundary,
    xmax = 240,
    ymin = 0,
    ymax = 1,
    fill = "#F2C94C",
    alpha = 0.18
  ) +
  geom_step(
    data = issue3_km,
    aes(time_months, survival),
    colour = "#008B8B",
    linewidth = 0.9,
    alpha = 0.8
  ) +
  geom_line(
    data = issue3_best_fit,
    aes(time_months, survival),
    colour = "#B22222",
    linewidth = 1.1
  ) +
  geom_vline(
    xintercept = issue3_median_follow_up_months,
    linetype = "dashed",
    colour = "#2F80ED"
  ) +
  geom_vline(
    xintercept = issue3_eag_two_year_boundary,
    linetype = "dashed",
    colour = "#B22222"
  ) +
  geom_vline(
    xintercept = issue3_active_treatment_months,
    linetype = "dotted",
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = issue3_median_follow_up_months + 2,
    y = 0.98,
    label = "Median FU\n16.3 mo",
    hjust = 0,
    size = 3.0,
    colour = "#2F80ED"
  ) +
  annotate(
    "text",
    x = issue3_eag_two_year_boundary + 2,
    y = 0.86,
    label = "2 years+\nmodel-driven",
    hjust = 0,
    size = 3.0,
    colour = "#B22222"
  ) +
  annotate(
    "text",
    x = issue3_active_treatment_months + 2,
    y = 0.74,
    label = "Approx max\nactive treatment",
    hjust = 0,
    size = 3.0,
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = 85,
    y = 0.95,
    label = "EAG concern: long-term extrapolated region",
    hjust = 0,
    size = 3.5,
    colour = "#7A5A00"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 240), breaks = seq(0, 240, 24)) +
  labs(
    title = "Issue 3 demonstration: observed pMMR OS evidence versus extrapolated model tail",
    subtitle = paste0("AIC-selected distribution: ", issue3_best_dist),
    x = "Months",
    y = "Overall survival probability",
    caption = "Teal step curve = reconstructed KM; red smooth curve = AIC-selected parametric extrapolation."
  ) +
  theme_minimal()
```

The point of this plot is not that the red curve is necessarily wrong. The point is that the area most relevant to 5-, 10-, and 20-year survival is not directly observed in the trial. It is a modelled continuation from early evidence.


### Parametric Tail Divergence & Landmark Predictions

#### Distribution Fan Plot

A second EAG-style check is to ask whether several distributions fit the observed period reasonably but imply different long-term tails.

```{r issue3-distribution-fan-plot}
issue3_all_distribution_predictions <- fit_tbl |>
  filter(curve_id == issue3_curve_id, converged) |>
  mutate(
    pred = map(
      fit,
      ~tibble(
        time_months = seq(0, 240, by = 1),
        survival = predict_survival(.x, seq(0, 240, by = 1))
      )
    )
  ) |>
  select(dist_label, aic, bic, pred) |>
  unnest(pred)

ggplot() +
  annotate(
    "rect",
    xmin = issue3_eag_two_year_boundary,
    xmax = 240,
    ymin = 0,
    ymax = 1,
    fill = "#F2C94C",
    alpha = 0.14
  ) +
  geom_step(
    data = issue3_km,
    aes(time_months, survival),
    colour = "black",
    linewidth = 0.9,
    alpha = 0.55
  ) +
  geom_line(
    data = issue3_all_distribution_predictions,
    aes(time_months, survival, colour = dist_label),
    linewidth = 0.9
  ) +
  geom_vline(
    xintercept = issue3_eag_two_year_boundary,
    linetype = "dashed",
    colour = "#B22222"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 240), breaks = seq(0, 240, 24)) +
  labs(
    title = "Issue 3 demonstration: plausible distributions can diverge after follow-up ends",
    subtitle = "The 2-year boundary is marked because the EAG considered 2 years+ estimates highly extrapolation-dependent",
    x = "Months",
    y = "Overall survival probability",
    colour = "Distribution"
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
```

This is the visual reason why short follow-up matters. The observed KM can look compatible with several parametric shapes, but the long-term tail can differ materially once the model moves beyond the trial data.


#### Landmark Survival Table

The same point can be expressed numerically. This table compares predicted survival at landmark time points that are far beyond the observed follow-up.

```{r issue3-landmark-survival-table}
issue3_landmarks <- c(24, 36, 60, 120, 240)

issue3_landmark_table <- issue3_all_distribution_predictions |>
  filter(time_months %in% issue3_landmarks) |>
  mutate(
    survival_percent = round(100 * survival, 1),
    aic = round(aic, 1)
  ) |>
  select(distribution = dist_label, aic, time_months, survival_percent) |>
  tidyr::pivot_wider(
    names_from = time_months,
    values_from = survival_percent,
    names_prefix = "month_"
  ) |>
  arrange(aic)

issue3_landmark_table |>
  kable(
    caption = "pMMR OS pembro arm: landmark survival predictions by distribution (%)",
    align = "l"
  )
```

If the 10-year or 20-year predictions differ meaningfully across distributions, then the model has structural uncertainty. This is not solved by choosing the lowest AIC alone, because AIC mainly assesses fit to the available observed data. It does not prove that the extrapolated tail is clinically plausible.


#### Appraisal Interpretation

This section turns the EAG's Issue 3 into a reproducible modelling demonstration:

- the EAG's concern is translated into concrete modelling checks;
- the observed follow-up, 2-year boundary, and 20-year horizon are shown on the same scale;
- alternative distributions show the range of plausible long-term tails;
- landmark predictions quantify how much the tails differ;
- the interpretation links directly to NICE/EAG judgement about survival extrapolation uncertainty.

The key portfolio message is: short follow-up matters when the cost-effectiveness result depends on survival benefits generated beyond the observed trial period. The issue is not that extrapolation is avoidable; oncology economic models often require extrapolation. The EAG concern is that, with immature OS and limited follow-up, the long-term survival benefit may be driven more by modelling assumptions than by observed trial evidence.


## Cost-Effectiveness under AIC-Selected Standard Parametric Models

::: {.column-margin}
**Time Log: 2.0 hours (Base Engine)**
- Coded the weekly cycle Partitioned Survival Model (PSM) engine over a 35-year horizon.
- Mapped health-state utility values and list treatment cost parameters.
:::


In this chapter, we implement a three-state Partitioned Survival Model (PSM) using a weekly cycle and a 35-year time horizon. This section focuses on health-state occupancy, discounted costs, and QALY outcomes under standard parametric curves selected purely by the best AIC fit.

### Economic Inputs & PSM Engine Setup

#### Overview of the Starter Cost-Effectiveness Module

The previous sections focus on survival modelling. To connect Issue 5 to the economic model, we need to show how PFS and OS extrapolations feed into cost-effectiveness.

This is not yet a NICE base-case replication. It is a transparent, simplified partitioned survival model (PSM) that uses extracted public parameters where I can currently verify them, and clearly labels remaining illustrative or unverified inputs. The purpose is to move from a pure survival-fitting exercise toward a mock cost-effectiveness case.

- PFS and OS curves determine health-state occupancy.
- Health-state occupancy determines QALYs and costs.
- OS tail assumptions can materially change incremental QALYs and the ICER.


#### Economic Parameter Table

The table below separates the inputs into three groups:

- **Extracted:** taken from the NotebookLM extraction output and plausible to use in this mock case.
- **Needs source verification:** extracted or implied, but should be checked against the original PDF before being treated as final.
- **Illustrative:** used only to keep the model runnable until a verified parameter is abstracted.

```{r starter-ce-inputs}
ce_inputs <- tibble::tribble(
  ~Input, ~Value, ~Status, ~Note,
  "Time horizon", "35 years", "Extracted", "Lifetime horizon in the economic model",
  "Cycle length", "1 week", "Extracted", "Consistent with weekly cost and utility calculations",
  "Annual discount rate", "3.5%", "Extracted", "Costs and health effects",
  "Perspective", "UK NHS and Personal Social Services", "Extracted", "Not directly used in this simplified calculation",
  "Model structure", "3-state PSM", "Extracted", "Progression-free, progressed disease, death",
  "Pembro treatment duration", "20 cycles; 6 Q3W combination + 14 Q6W maintenance", "Extracted", "Simplified as 18 weeks + 84 weeks of active pembro cost",
  "Pembro stopping rule", "24 months / 2 years", "Extracted", "Approximated by the 102-week active dosing schedule in this mock model",
  "Chemotherapy duration", "6 cycles Q3W", "Extracted", "Applied as 18 weeks of active chemotherapy cost in this mock model",
  "Pembrolizumab list price", "&#163;2,630 per 100 mg vial", "Extracted", "List price; confidential commercial arrangement not available",
  "Pembrolizumab combination cost", "&#163;5,260 pembro drug + &#163;106 chemotherapy drug + &#163;277 admin per Q3W cycle", "Extracted and inferred", "Before confidential discount; chemotherapy drug cost is added using the comparator RDI-adjusted value",
  "Pembrolizumab maintenance cost", "&#163;10,520 drug + &#163;217 admin per Q6W cycle", "Extracted", "Before confidential discount; smoothed to weekly cost in this mock model",
  "Comparator chemotherapy cost", "&#163;106 drug + &#163;277 admin per Q3W cycle", "Extracted", "RDI-adjusted drug cost plus complex chemotherapy administration",
  "Terminal care cost", "&#163;8,829.07 one-off at death", "Extracted", "Added in this mock model through weekly death increments",
  "PF utility", "0.814", "Extracted", "Updated base-case PF utility in the NotebookLM extraction",
  "PD utility", "0.720", "Extracted", "Updated base-case PD utility in the NotebookLM extraction",
  "PF resource-use cost", "&#163;73.59 on-treatment Q3W; &#163;44.15 pembro Q6W maintenance (weeks 19-102); &#163;27.59 off-treatment year 1-2; &#163;13.79 off-treatment year 3+", "Corrected from extracted inputs", "NICE updated model frequencies: 0.33 (Q3W on-treatment), 0.17 (Q6W pembro maintenance), 0.08 (off-treatment year 1-2), 0.04 (off-treatment year 3+) for CT scan/outpatient/blood test",
  "PD resource-use cost", "&#163;26.67 per week", "Inferred from extracted inputs", "Computed from extracted unit costs and weekly resource frequencies",
  "AE cost", "One-off Grade 3+ anaemia and hypertension cost", "Inferred from extracted inputs", "Duration for AE QALY decrement remains redacted, so only one-off AE costs are included here",
  "Confidential pembrolizumab discount", "Scenario parameter", "Unavailable", "Handled through pembro_discount sensitivity rather than assumed in the base case"
)

ce_inputs |>
  kable(caption = "Economic inputs used in the mock PSM", align = "l", escape = FALSE)
```

```{r starter-ce-parameters}
ce_params <- list(
  horizon_weeks = 35 * 52,
  cycle_length_years = 1 / 52,
  annual_discount_rate = 0.035,
  utility_pf = 0.814,
  utility_pd = 0.720,
  weekly_cost_pf_initial = 0.08 * 160.83 + 0.33 * 179 + 0.33 * 5,
  weekly_cost_pf_pembro_maint = 0.08 * 160.83 + 0.17 * 179 + 0.17 * 5,
  weekly_cost_pf_offtreatment = 0.08 * 160.83 + 0.08 * 179 + 0.08 * 5,
  weekly_cost_pf_offtreatment_yr3 = 0.04 * 160.83 + 0.04 * 179 + 0.04 * 5,
  weekly_cost_pd = 0.04 * 160.83 + 0.11 * 179 + 0.11 * 5,
  weekly_cost_pembro_combination = (5260 + 106 + 277) / 3,
  weekly_cost_pembro_maintenance = (10520 + 217) / 6,
  weekly_cost_comparator_chemo = (106 + 277) / 3,
  pembro_discount = 0,
  pembro_combination_weeks = 18,
  pembro_maintenance_weeks = 84,
  pembro_treatment_cap_weeks = 18 + 84,
  comparator_treatment_cap_weeks = 18,
  terminal_care_cost = 8829.07,
  one_off_ae_cost_pembro = 0.169 * 565.40 + 0.056 * 735.07,
  one_off_ae_cost_comparator = 0.116 * 565.40 + 0.052 * 735.07
)

param_display <- tibble(
  Parameter = c(
    "horizon_weeks",
    "cycle_length_years",
    "annual_discount_rate",
    "utility_pf",
    "utility_pd",
    "weekly_cost_pf_initial",
    "weekly_cost_pf_pembro_maint",
    "weekly_cost_pf_offtreatment",
    "weekly_cost_pf_offtreatment_yr3",
    "weekly_cost_pd",
    "weekly_cost_pembro_combination",
    "weekly_cost_pembro_maintenance",
    "weekly_cost_comparator_chemo",
    "pembro_discount",
    "pembro_combination_weeks",
    "pembro_maintenance_weeks",
    "pembro_treatment_cap_weeks",
    "comparator_treatment_cap_weeks",
    "terminal_care_cost",
    "one_off_ae_cost_pembro",
    "one_off_ae_cost_comparator"
  ),
  Value = c(
    "1,820 weeks",
    "1/52 year per cycle",
    "3.5% per year",
    "0.814",
    "0.720",
    scales::dollar(ce_params$weekly_cost_pf_initial, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_pembro_maint, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment_yr3, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pd, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_combination, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_maintenance, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_comparator_chemo, prefix = "&#163;"),
    "0% in base case",
    "18 weeks",
    "84 weeks",
    "102 weeks",
    "18 weeks",
    "&#163;8,829.07",
    scales::dollar(ce_params$one_off_ae_cost_pembro, prefix = "&#163;"),
    scales::dollar(ce_params$one_off_ae_cost_comparator, prefix = "&#163;")
  )
)

tibble(
  Parameter_1 = c(
    "horizon_weeks", "cycle_length_years", "annual_discount_rate",
    "utility_pf", "utility_pd", "weekly_cost_pf_initial",
    "weekly_cost_pf_pembro_maint", "weekly_cost_pf_offtreatment",
    "weekly_cost_pf_offtreatment_yr3", "weekly_cost_pd",
    "weekly_cost_pembro_combination"
  ),
  Value_1 = c(
    "1,820 weeks", "1/52 year per cycle", "3.5% per year",
    "0.814", "0.720",
    scales::dollar(ce_params$weekly_cost_pf_initial,        prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_pembro_maint,   prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment,   prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pf_offtreatment_yr3, prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pd,                prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_pembro_combination, prefix = "&#163;")
  ),
  Parameter_2 = c(
    "weekly_cost_pembro_maintenance", "weekly_cost_comparator_chemo",
    "pembro_discount", "pembro_combination_weeks", "pembro_maintenance_weeks",
    "pembro_treatment_cap_weeks", "comparator_treatment_cap_weeks",
    "terminal_care_cost", "one_off_ae_cost_pembro", "one_off_ae_cost_comparator", ""
  ),
  Value_2 = c(
    scales::dollar(ce_params$weekly_cost_pembro_maintenance,  prefix = "&#163;"),
    scales::dollar(ce_params$weekly_cost_comparator_chemo,    prefix = "&#163;"),
    "0% in base case", "18 weeks", "84 weeks", "102 weeks", "18 weeks",
    "&#163;8,829.07",
    scales::dollar(ce_params$one_off_ae_cost_pembro,          prefix = "&#163;"),
    scales::dollar(ce_params$one_off_ae_cost_comparator,      prefix = "&#163;"),
    ""
  )
) |>
  kable(
    col.names = c("Parameter", "Value", "Parameter", "Value"),
    caption   = "Parameter object used by the simplified PSM",
    align     = "l",
    escape    = FALSE
  )

ce_time_grid <- tibble(
  time_weeks  = 0:ce_params$horizon_weeks,
  time_months = time_weeks * 12 / 52
)

discount_factor <- function(week, annual_rate) {
  1 / ((1 + annual_rate) ^ (week / 52))
}
```


### pMMR Cohort Analysis

```{r pmmr-aic-state-occupancy}
psm_pmmr_aic <- best_fits |>
  filter(subgroup == "pMMR", endpoint %in% c("PFS", "OS")) |>
  mutate(
    arm  = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
    pred = map(
      fit,
      ~tibble(
        time_weeks  = ce_time_grid$time_weeks,
        time_months = ce_time_grid$time_months,
        survival    = predict_survival(.x, ce_time_grid$time_months)
      )
    )
  ) |>
  select(arm, endpoint, pred) |>
  unnest(pred) |>
  tidyr::pivot_wider(names_from = endpoint, values_from = survival) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

# AIC-selected distributions for pMMR
best_fits |>
  filter(subgroup == "pMMR", endpoint %in% c("PFS", "OS")) |>
  select(subgroup, endpoint, treatment, best_dist = dist_label, aic) |>
  arrange(endpoint, treatment) |>
  kable(digits = 2, caption = "pMMR: AIC-selected distributions", align = "l")
```

```{r pmmr-aic-ce-results}
calc_incremental <- function(psm_data) {
  psm_data |>
    group_by(arm) |>
    arrange(time_weeks, .by_group = TRUE) |>
    mutate(
      discount         = discount_factor(time_weeks, ce_params$annual_discount_rate),
      death_increment  = pmax(Dead - lag(Dead, default = 0), 0),
      pf_resource_cost = case_when(
        time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_initial,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_pembro_maint,
        time_weeks <= 104 ~
          ce_params$weekly_cost_pf_offtreatment,
        TRUE ~
          ce_params$weekly_cost_pf_offtreatment_yr3
      ),
      qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
        ce_params$cycle_length_years * discount,
      disease_management_cost = (
        PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
      ) * discount,
      treatment_cost = case_when(
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
          PF * ce_params$weekly_cost_pembro_combination *
            (1 - ce_params$pembro_discount) * discount,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_pembro_maintenance *
            (1 - ce_params$pembro_discount) * discount,
        arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_comparator_chemo * discount,
        TRUE ~ 0
      ),
      ae_cost = case_when(
        arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
        arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
        TRUE ~ 0
      ),
      terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
      total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
    ) |>
    ungroup() |>
    group_by(arm) |>
    summarise(qalys = sum(qaly), costs = sum(total_cost), .groups = "drop") |>
    summarise(
      incremental_qalys = qalys[arm == "Pembro + chemo"] - qalys[arm == "Placebo + chemo"],
      incremental_costs = costs[arm == "Pembro + chemo"] - costs[arm == "Placebo + chemo"],
      icer = incremental_costs / incremental_qalys
    )
}

ce_incremental_pmmr_aic <- calc_incremental(psm_pmmr_aic)

ce_incremental_pmmr_aic |>
  mutate(across(c(incremental_qalys), ~round(.x, 3)),
         across(c(incremental_costs, icer), ~round(.x, 0))) |>
  transmute(`Incremental QALYs` = incremental_qalys,
            `Incremental costs (&#163;)` = incremental_costs,
            `ICER (&#163;/QALY)` = icer) |>
  kable(caption = "pMMR (AIC-selected): incremental cost-effectiveness", align = "l", escape = FALSE)
```


### dMMR Cohort Analysis

```{r dmmr-aic-state-occupancy}
psm_dmmr_aic <- best_fits |>
  filter(subgroup == "dMMR", endpoint %in% c("PFS", "OS")) |>
  mutate(
    arm  = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
    pred = map(
      fit,
      ~tibble(
        time_weeks  = ce_time_grid$time_weeks,
        time_months = ce_time_grid$time_months,
        survival    = predict_survival(.x, ce_time_grid$time_months)
      )
    )
  ) |>
  select(arm, endpoint, pred) |>
  unnest(pred) |>
  tidyr::pivot_wider(names_from = endpoint, values_from = survival) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

best_fits |>
  filter(subgroup == "dMMR", endpoint %in% c("PFS", "OS")) |>
  select(subgroup, endpoint, treatment, best_dist = dist_label, aic) |>
  arrange(endpoint, treatment) |>
  kable(digits = 2, caption = "dMMR: AIC-selected distributions", align = "l")
```

```{r dmmr-aic-ce-results}
ce_incremental_dmmr_aic <- calc_incremental(psm_dmmr_aic)

ce_incremental_dmmr_aic |>
  mutate(across(c(incremental_qalys), ~round(.x, 3)),
         across(c(incremental_costs, icer), ~round(.x, 0))) |>
  transmute(`Incremental QALYs` = incremental_qalys,
            `Incremental costs (&#163;)` = incremental_costs,
            `ICER (&#163;/QALY)` = icer) |>
  kable(caption = "dMMR (AIC-selected): incremental cost-effectiveness", align = "l", escape = FALSE)
```


### Subgroup Synthesis & Comparison

Below, we compare the cost-effectiveness results of the pMMR and dMMR subgroups under the AIC-selected standard parametric distributions.

```{r pmmr-dmmr-comparison-aic}
bind_rows(
  ce_incremental_pmmr_aic |> mutate(subgroup = "pMMR", model = "AIC-selected standard parametric"),
  ce_incremental_dmmr_aic |> mutate(subgroup = "dMMR", model = "AIC-selected standard parametric")
) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Subgroup = subgroup,
    `Curve selection` = model,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs (£)` = incremental_costs,
    `ICER (£/QALY)` = icer
  ) |>
  kable(
    caption = "pMMR (AIC-selected) vs dMMR (AIC-selected) cost-effectiveness comparison",
    align = "l",
    escape = FALSE
  )
```

## Cost-Effectiveness under EAG-Preferred Survival Models

::: {.column-margin}
**Time Log: 3.5 hours (EAG Scenarios)**
- Programmed Royston-Parmar Splines and the `pmin(PFS, OS)` crossing diagnostic capping rules.
- Integrated the 5-7 year immunotherapy treatment efficacy waning function.
- Reallocated subsequent treatment market shares to reflect UK NHS guidelines.
:::


In this chapter, we evaluate the cost-effectiveness of pembrolizumab plus chemotherapy across subgroups using the EAG-preferred curve selections. For the pMMR cohort, the EAG preferred Royston-Parmar flexible spline models, whereas for the dMMR cohort, standard parametric models were retained but alternative distributions were selected.

### Spline and Standard Model Fitting Setup

#### EAG Spline Model Fitting (pMMR)

The EAG preferred spline models for pMMR (Table 2) are Royston-Parmar flexible parametric models fitted with `flexsurvspline()`. Unlike standard parametric distributions, spline models can accommodate non-monotonic hazard functions and are more flexible in the observed period.

```{r pmmr-spline-function}
fit_one_spline <- function(dat, k, scale, label) {
  tryCatch({
    fit     <- flexsurvspline(Surv(time_months, status) ~ 1, data = dat, k = k, scale = scale)
    fit_aic <- as.numeric(AIC(fit) %||% NA_real_)
    fit_ll  <- as.numeric(fit$loglik %||% NA_real_)
    tibble(
      dist_label = label,
      fit        = list(fit),
      converged  = TRUE,
      error      = NA_character_,
      aic        = fit_aic,
      loglik     = fit_ll,
      n          = nrow(dat),
      events     = sum(dat$status)
    )
  }, error = function(e) {
    tibble(
      dist_label = label,
      fit        = list(NULL),
      converged  = FALSE,
      error      = conditionMessage(e),
      aic        = NA_real_,
      loglik     = NA_real_,
      n          = nrow(dat),
      events     = sum(dat$status)
    )
  })
}

tibble(
  `Function` = "fit_one_spline(dat, k, scale, label)",
  `Backend`  = "flexsurvspline()",
  `Purpose`  = "Fit Royston-Parmar spline model for one curve"
) |> kable()
```

```{r pmmr-spline-specs}
eag_pmmr_spline_specs <- tibble::tribble(
  ~subgroup, ~endpoint, ~treatment,      ~k, ~scale,   ~dist_label,
  "pMMR",    "PFS",     "placebo_chemo", 2,  "hazard", "2-knot hazards",
  "pMMR",    "OS",      "placebo_chemo", 1,  "hazard", "1-knot hazards",
  "pMMR",    "PFS",     "pembro_chemo",  1,  "hazard", "1-knot hazards",
  "pMMR",    "OS",      "pembro_chemo",  1,  "normal", "1-knot normal"
)

eag_pmmr_spline_specs |>
  kable(caption = "EAG-preferred spline specifications for pMMR (Table 2)", align = "l")
```

```{r pmmr-spline-fits}
spline_fit_rows <- vector("list", nrow(eag_pmmr_spline_specs))
for (i in seq_len(nrow(eag_pmmr_spline_specs))) {
  spec_i <- eag_pmmr_spline_specs[i, ]
  dat_i  <- ipd |> filter(
    as.character(subgroup)  == spec_i$subgroup,
    as.character(endpoint)  == spec_i$endpoint,
    as.character(treatment) == spec_i$treatment
  )
  res_i <- fit_one_spline(dat_i, spec_i$k, spec_i$scale, spec_i$dist_label)
  spline_fit_rows[[i]] <- bind_cols(
    tibble(subgroup = spec_i$subgroup,
           endpoint = spec_i$endpoint,
           treatment = spec_i$treatment),
    res_i
  )
}
eag_pmmr_spline_fits <- bind_rows(spline_fit_rows)

eag_pmmr_spline_fits |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, converged, error) |>
  kable(digits = 2, caption = "EAG-preferred spline fit summary for pMMR", align = "l")

```


#### EAG Model Selection (dMMR)

```{r dmmr-eag-model-selection}
eag_dmmr_selections <- tibble::tribble(
  ~subgroup, ~endpoint, ~treatment,      ~dist_label,
  "dMMR",    "PFS",     "placebo_chemo", "generalized_gamma",
  "dMMR",    "OS",      "placebo_chemo", "exponential",
  "dMMR",    "PFS",     "pembro_chemo",  "loglogistic",
  "dMMR",    "OS",      "pembro_chemo",  "loglogistic"
)

eag_dmmr_fits <- fit_tbl |>
  inner_join(eag_dmmr_selections, by = c("subgroup", "endpoint", "treatment", "dist_label")) |>
  filter(converged)

eag_dmmr_fits |>
  select(subgroup, endpoint, treatment, dist_label, n, events, aic, bic) |>
  arrange(endpoint, treatment) |>
  kable(digits = 2, caption = "EAG-preferred distributions for dMMR (Table 2)", align = "l")
```


### pMMR Cohort Analysis

#### State Occupancy with EAG Spline Models

```{r pmmr-spline-state-occupancy}
get_spline_surv <- function(fit_obj, times) {
  # Unwrap list(model_obj) wrapper only if not already a flexsurv model
  if (!inherits(fit_obj, "flexsurvreg")) fit_obj <- fit_obj[[1]]
  s <- summary(fit_obj, t = times, type = "survival", ci = FALSE)
  if (is.data.frame(s))                    return(as.numeric(s$est))
  if (is.list(s) && is.data.frame(s[[1]])) return(as.numeric(s[[1]]$est))
  if (is.list(s) && is.numeric(s[[1]]))    return(as.numeric(s[[1]]))
  stop(paste("Cannot extract survival: class(s)=", class(s), "class(s[[1]])=", class(s[[1]])))
}

pmmr_sp_curves <- eag_pmmr_spline_fits |>
  filter(converged) |>
  mutate(arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo")) |>
  select(arm, endpoint, fit)

surv_rows <- vector("list", nrow(pmmr_sp_curves))
for (i in seq_len(nrow(pmmr_sp_curves))) {
  surv_rows[[i]] <- tibble(
    arm         = pmmr_sp_curves$arm[[i]],
    endpoint    = pmmr_sp_curves$endpoint[[i]],
    time_weeks  = ce_time_grid$time_weeks,
    time_months = ce_time_grid$time_months,
    surv        = get_spline_surv(pmmr_sp_curves$fit[[i]], ce_time_grid$time_months)
  )
}

psm_pmmr_spline <- bind_rows(surv_rows) |>
  tidyr::pivot_wider(
    id_cols     = c(arm, time_weeks, time_months),
    names_from  = endpoint,
    values_from = surv
  ) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

psm_pmmr_spline |>
  select(arm, time_weeks, PF, PD, Dead) |>
  pivot_longer(cols = c(PF, PD, Dead), names_to = "state", values_to = "proportion") |>
  mutate(time_years = time_weeks / 52) |>
  ggplot(aes(time_years, proportion, fill = state)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~arm) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 35), breaks = seq(0, 35, 5)) +
  labs(
    title    = "pMMR partitioned survival model: EAG-preferred spline models",
    subtitle = "CT PFS: 2-knot hazards | CT OS: 1-knot hazards | Pembro PFS: 1-knot hazards | Pembro OS: 1-knot normal",
    x = "Years", y = "Proportion of cohort", fill = "Health state"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


#### Discounted Costs and QALYs (EAG Spline Models)

```{r pmmr-spline-ce-results}
ce_trace_pmmr_spline <- psm_pmmr_spline |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount         = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment  = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_results_pmmr_spline <- ce_trace_pmmr_spline |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  )

ce_incremental_pmmr_spline <- ce_results_pmmr_spline |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

ce_results_pmmr_spline |>
  mutate(
    discounted_qalys = round(discounted_qalys, 3),
    discounted_costs = round(discounted_costs, 0)
  ) |>
  transmute(Arm = arm,
            `Discounted QALYs` = discounted_qalys,
            `Discounted costs (&#163;)` = discounted_costs) |>
  kable(caption = "pMMR (EAG spline): discounted results by arm", align = "l", escape = FALSE)

ce_incremental_pmmr_spline |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer              = round(icer, 0)
  ) |>
  transmute(`Incremental QALYs`     = incremental_qalys,
            `Incremental costs (&#163;)` = incremental_costs,
            `ICER (&#163;/QALY)`        = icer) |>
  kable(caption = "pMMR (EAG spline): incremental cost-effectiveness", align = "l", escape = FALSE)
```


### dMMR Cohort Analysis

#### State Occupancy with EAG-Preferred Models

```{r dmmr-state-occupancy}
psm_dmmr <- eag_dmmr_fits |>
  mutate(
    arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
    pred = map(
      fit,
      ~tibble(
        time_weeks = ce_time_grid$time_weeks,
        time_months = ce_time_grid$time_months,
        survival = predict_survival(.x, ce_time_grid$time_months)
      )
    )
  ) |>
  select(arm, endpoint, pred) |>
  unnest(pred) |>
  tidyr::pivot_wider(names_from = endpoint, values_from = survival) |>
  mutate(
    PFS  = pmin(PFS, OS),
    PF   = PFS,
    PD   = pmax(OS - PFS, 0),
    Dead = pmax(1 - OS, 0)
  )

psm_dmmr |>
  select(arm, time_weeks, PF, PD, Dead) |>
  pivot_longer(cols = c(PF, PD, Dead), names_to = "state", values_to = "proportion") |>
  mutate(time_years = time_weeks / 52) |>
  ggplot(aes(time_years, proportion, fill = state)) +
  geom_area(alpha = 0.85) +
  facet_wrap(~arm) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 35), breaks = seq(0, 35, 5)) +
  labs(
    title = "dMMR partitioned survival model (EAG-preferred distributions)",
    subtitle = "EAG: generalised gamma / exponential / log-logistic / log-logistic",
    x = "Years",
    y = "Proportion of cohort",
    fill = "Health state"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


#### Discounted Costs and QALYs

```{r dmmr-ce-results}
ce_trace_dmmr <- psm_dmmr |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount        = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_results_dmmr <- ce_trace_dmmr |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  )

ce_incremental_dmmr <- ce_results_dmmr |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

ce_results_dmmr |>
  mutate(
    discounted_qalys = round(discounted_qalys, 3),
    discounted_costs = round(discounted_costs, 0)
  ) |>
  transmute(
    Arm = arm,
    `Discounted QALYs` = discounted_qalys,
    `Discounted costs (&#163;)` = discounted_costs
  ) |>
  kable(caption = "dMMR: discounted results by arm (EAG-preferred models)", align = "l", escape = FALSE)

ce_incremental_dmmr |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs (&#163;)` = incremental_costs,
    `ICER (&#163;/QALY)` = icer
  ) |>
  kable(caption = "dMMR: incremental cost-effectiveness (EAG-preferred models)", align = "l", escape = FALSE)
```


### Subgroup Synthesis & Comparison

```{r pmmr-dmmr-comparison}
bind_rows(
  ce_incremental_pmmr_spline |> mutate(subgroup = "pMMR", model = "EAG-preferred spline (Table 2)"),
  ce_incremental_dmmr        |> mutate(subgroup = "dMMR", model = "EAG-preferred standard (Table 2)")
) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Subgroup = subgroup,
    `Curve selection` = model,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs (&#163;)` = incremental_costs,
    `ICER (&#163;/QALY)` = icer
  ) |>
  kable(
    caption = "pMMR (EAG spline) vs dMMR (EAG standard) cost-effectiveness comparison",
    align = "l",
    escape = FALSE
  )
```

Interpretation notes:

- pMMR uses EAG-preferred spline models (Table 2); dMMR uses EAG-preferred standard parametric models (Table 2).
- Both subgroups apply pMMR-derived utility values (0.814 / 0.720) because dMMR HRQoL data were not collected in KEYNOTE-868.
- Pembrolizumab confidential discount is set to 0% in both calculations; the real ICER under PAS pricing would be lower.


## Model Selection Critique: AIC-Selected vs. EAG-Preferred Models

::: {.column-margin}
**Time Log: 2.0 hours (Audit Synthesis)**
- Synthesized the final 2x2 ICER comparison tables.
- Conducted methodological reflections on the subsequent therapy "cost-efficacy disconnect" and crossover correction.
:::


This chapter summarizes the cost-effectiveness estimates across the different survival modeling approaches. We present a 2×2 comparison table, explore sensitivity analyses around confidential discounts and overall survival tail assumptions, and discuss key HTA consulting takeaways.

### 2×2 Model Selection Synthesis

```{r model-selection-comparison}
bind_rows(
  ce_incremental_pmmr_aic         |> mutate(subgroup = "pMMR", model = "AIC-selected (standard parametric)"),
  ce_incremental_pmmr_spline      |> mutate(subgroup = "pMMR", model = "EAG-preferred (spline, Table 2)"),
  ce_incremental_dmmr_aic         |> mutate(subgroup = "dMMR", model = "AIC-selected (standard parametric)"),
  ce_incremental_dmmr             |> mutate(subgroup = "dMMR", model = "EAG-preferred (standard, Table 2)")
) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer              = round(icer, 0)
  ) |>
  transmute(
    Subgroup          = subgroup,
    `Model selection` = model,
    `Incremental QALYs`      = incremental_qalys,
    `Incremental costs (&#163;)` = incremental_costs,
    `ICER (&#163;/QALY)`         = icer
  ) |>
  kable(
    caption = "2×2 comparison: AIC-selected vs EAG-preferred models, by subgroup",
    align   = "l",
    escape  = FALSE
  )
```

The key question this table answers: how sensitive is the ICER to the choice of survival model? If AIC-selected and EAG-preferred models produce similar ICERs, the cost-effectiveness conclusion is robust to curve selection. If they diverge substantially, model selection becomes a material source of decision uncertainty.



### Visual Extrapolation Comparison: AIC vs. EAG

To visually understand why the cost-effectiveness results differ substantially, we plot the reconstructed KM curves alongside the fitted and extrapolated curves for both the AIC-selected standard parametric models and the EAG-preferred spline models. The plots show the first 10 years (120 months) of projection.

```{r visual-extrapolation-comparison}
#| fig-width: 10
#| fig-height: 6
library(ggplot2)
library(dplyr)
library(tidyr)

# Combine PFS Plot Data
plot_data_pfs <- bind_rows(
  km_points |>
    filter(subgroup == "pMMR", endpoint == "PFS") |>
    mutate(
      arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
      source = "Reconstructed KM",
      model = "KM"
    ) |>
    select(arm, time_months, survival, source, model),
  psm_pmmr_aic |>
    mutate(
      survival = PFS,
      source = "AIC-selected standard parametric",
      model = "AIC"
    ) |>
    select(arm, time_months, survival, source, model),
  psm_pmmr_spline |>
    mutate(
      survival = PFS,
      source = "EAG-preferred spline model",
      model = "EAG"
    ) |>
    select(arm, time_months, survival, source, model)
)

# Combine OS Plot Data
plot_data_os <- bind_rows(
  km_points |>
    filter(subgroup == "pMMR", endpoint == "OS") |>
    mutate(
      arm = if_else(treatment == "pembro_chemo", "Pembro + chemo", "Placebo + chemo"),
      source = "Reconstructed KM",
      model = "KM"
    ) |>
    select(arm, time_months, survival, source, model),
  psm_pmmr_aic |>
    mutate(
      survival = PF + PD,
      source = "AIC-selected standard parametric",
      model = "AIC"
    ) |>
    select(arm, time_months, survival, source, model),
  psm_pmmr_spline |>
    mutate(
      survival = PF + PD,
      source = "EAG-preferred spline model",
      model = "EAG"
    ) |>
    select(arm, time_months, survival, source, model)
)

# Colors and Linetypes
model_cols <- c(
  "Reconstructed KM" = "#2D3748", 
  "AIC-selected standard parametric" = "#E53E3E", 
  "EAG-preferred spline model" = "#3182CE"
)
model_lt <- c(
  "Reconstructed KM" = "solid", 
  "AIC-selected standard parametric" = "dashed", 
  "EAG-preferred spline model" = "solid"
)

# PFS Plot
p_pfs <- ggplot(plot_data_pfs |> filter(time_months <= 120), aes(time_months, survival, colour = source, linetype = source)) +
  geom_step(data = ~filter(.x, model == "KM"), linewidth = 0.8, alpha = 0.6) +
  geom_line(data = ~filter(.x, model != "KM"), linewidth = 1.0) +
  facet_wrap(~arm, ncol = 2) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 12)) +
  scale_colour_manual(values = model_cols) +
  scale_linetype_manual(values = model_lt) +
  labs(
    title = "pMMR PFS: Extrapolation Comparison (AIC vs. EAG)",
    subtitle = "Reconstructed KM vs. AIC standard parametric (dashed, red) vs. EAG spline (solid, blue) up to 10 years",
    x = "Months",
    y = "PFS Probability",
    colour = "Model Selection",
    linetype = "Model Selection"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10)
  )

print(p_pfs)

# OS Plot
p_os <- ggplot(plot_data_os |> filter(time_months <= 120), aes(time_months, survival, colour = source, linetype = source)) +
  geom_step(data = ~filter(.x, model == "KM"), linewidth = 0.8, alpha = 0.6) +
  geom_line(data = ~filter(.x, model != "KM"), linewidth = 1.0) +
  facet_wrap(~arm, ncol = 2) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(limits = c(0, 120), breaks = seq(0, 120, 12)) +
  scale_colour_manual(values = model_cols) +
  scale_linetype_manual(values = model_lt) +
  labs(
    title = "pMMR OS: Extrapolation Comparison (AIC vs. EAG)",
    subtitle = "Reconstructed KM vs. AIC standard parametric (dashed, red) vs. EAG spline (solid, blue) up to 10 years",
    x = "Months",
    y = "OS Probability",
    colour = "Model Selection",
    linetype = "Model Selection"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10)
  )

print(p_os)
```


### Sensitivity Analysis: Confidential Discounts & OS Tail Reduction

We now implement sensitivity analyses for the confidential discount of pembrolizumab and the OS tail reduction assumption.

#### Confidential Discount Sensitivity

The pembrolizumab commercial arrangement is confidential. Instead of guessing it, the mock model treats the discount as a scenario parameter. This makes the limitation explicit and lets us ask a practical question: how sensitive is the ICER to the assumed pembrolizumab acquisition discount?

```{r starter-ce-discount-sensitivity}
#| eval: false
calculate_incremental_ce <- function(trace) {
  trace |>
    group_by(arm) |>
    summarise(
      discounted_qalys = sum(qaly),
      discounted_costs = sum(total_cost),
      .groups = "drop"
    ) |>
    summarise(
      incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
        discounted_qalys[arm == "Placebo + chemo"],
      incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
        discounted_costs[arm == "Placebo + chemo"],
      icer = incremental_costs / incremental_qalys
    )
}

run_discount_sensitivity <- function(discount_value) {
  psm_survival |>
    group_by(arm) |>
    arrange(time_weeks, .by_group = TRUE) |>
    mutate(
      discount = discount_factor(time_weeks, ce_params$annual_discount_rate),
      death_increment = pmax(Dead - lag(Dead, default = 0), 0),
      pf_resource_cost = case_when(
        time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_initial,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_pembro_maint,
        time_weeks <= 104 ~
          ce_params$weekly_cost_pf_offtreatment,
        TRUE ~
          ce_params$weekly_cost_pf_offtreatment_yr3
      ),
      qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
        ce_params$cycle_length_years * discount,
      disease_management_cost = (
        PF * pf_resource_cost +
          PD * ce_params$weekly_cost_pd
      ) * discount,
      treatment_cost = case_when(
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
          PF * ce_params$weekly_cost_pembro_combination *
            (1 - discount_value) * discount,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_pembro_maintenance *
            (1 - discount_value) * discount,
        arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_comparator_chemo * discount,
        TRUE ~ 0
      ),
      ae_cost = case_when(
        arm == "Pembro + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
        arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
        TRUE ~ 0
      ),
      terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
      total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
    ) |>
    ungroup() |>
    calculate_incremental_ce() |>
    mutate(pembro_discount = discount_value)
}

map_dfr(c(0, 0.25, 0.50, 0.75), run_discount_sensitivity) |>
  mutate(
    pembro_discount = scales::percent(pembro_discount, accuracy = 1),
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    `Pembro discount` = pembro_discount,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs` = incremental_costs,
    ICER = icer
  ) |>
  kable(caption = "Illustrative confidential-discount sensitivity", align = "l")
```


#### OS Tail Sensitivity

Issue 5 is about the uncertain degree of OS benefit. A simple way to demonstrate this is to reduce the long-term OS separation between arms after the EAG's 2-year boundary.

This scenario is not an EAG preferred model. It is a teaching sensitivity: after 104 weeks, the pembro OS benefit is gradually reduced so that only 50% of the modelled OS separation remains by 10 years.

```{r starter-ce-os-tail-sensitivity}
#| eval: false
apply_os_tail_reduction <- function(dat, start_week = 104, end_week = 520, final_fraction = 0.5) {
  dat |>
    select(arm, time_weeks, time_months, PFS, OS) |>
    tidyr::pivot_wider(names_from = arm, values_from = c(PFS, OS)) |>
    mutate(
      reduction_fraction = case_when(
        time_weeks <= start_week ~ 1,
        time_weeks >= end_week ~ final_fraction,
        TRUE ~ 1 - (1 - final_fraction) * (time_weeks - start_week) / (end_week - start_week)
      ),
      `OS_Pembro + chemo` = `OS_Placebo + chemo` +
        (`OS_Pembro + chemo` - `OS_Placebo + chemo`) * reduction_fraction
    ) |>
    select(time_weeks, time_months, starts_with("PFS_"), starts_with("OS_")) |>
    pivot_longer(
      cols = -c(time_weeks, time_months),
      names_to = c("endpoint", "arm"),
      names_sep = "_",
      values_to = "survival"
    ) |>
    pivot_wider(names_from = endpoint, values_from = survival) |>
    mutate(
      PF = pmin(PFS, OS),
      PD = pmax(OS - PFS, 0),
      Dead = pmax(1 - OS, 0)
    )
}

psm_survival_os_sensitivity <- apply_os_tail_reduction(psm_survival)

ce_trace_os_sensitivity <- psm_survival_os_sensitivity |>
  group_by(arm) |>
  arrange(time_weeks, .by_group = TRUE) |>
  mutate(
    discount = discount_factor(time_weeks, ce_params$annual_discount_rate),
    death_increment = pmax(Dead - lag(Dead, default = 0), 0),
    pf_resource_cost = case_when(
      time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_initial,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        ce_params$weekly_cost_pf_pembro_maint,
      time_weeks <= 104 ~
        ce_params$weekly_cost_pf_offtreatment,
      TRUE ~
        ce_params$weekly_cost_pf_offtreatment_yr3
    ),
    qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
      ce_params$cycle_length_years * discount,
    disease_management_cost = (
      PF * pf_resource_cost +
        PD * ce_params$weekly_cost_pd
    ) * discount,
    treatment_cost = case_when(
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
        PF * ce_params$weekly_cost_pembro_combination *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_pembro_maintenance *
          (1 - ce_params$pembro_discount) * discount,
      arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
        PF * ce_params$weekly_cost_comparator_chemo * discount,
      TRUE ~ 0
    ),
    ae_cost = case_when(
      arm == "Pembro + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
      arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
      TRUE ~ 0
    ),
    terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
    total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
  ) |>
  ungroup()

ce_scenario_results <- ce_trace_os_sensitivity |>
  group_by(arm) |>
  summarise(
    discounted_qalys = sum(qaly),
    discounted_costs = sum(total_cost),
    .groups = "drop"
  ) |>
  summarise(
    incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
      discounted_qalys[arm == "Placebo + chemo"],
    incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
      discounted_costs[arm == "Placebo + chemo"],
    icer = incremental_costs / incremental_qalys
  )

bind_rows(
  ce_incremental |> mutate(scenario = "AIC-selected OS tail"),
  ce_scenario_results |> mutate(scenario = "Reduced long-term OS separation")
) |>
  select(scenario, incremental_qalys, incremental_costs, icer) |>
  mutate(
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Scenario = scenario,
    `Incremental QALYs` = incremental_qalys,
    `Incremental costs` = incremental_costs,
    ICER = icer
  ) |>
  kable(caption = "Illustrative sensitivity to long-term OS benefit", align = "l")
```

This is the bridge to Issue 5. If reducing the long-term OS separation reduces incremental QALYs and increases the ICER, then the OS tail is decision-relevant. That is why the EAG focuses on whether the company's long-term OS benefit is clinically plausible.


### Clinical Logic Validation and Capping under Parameter Uncertainty

While the EAG-preferred spline models for the pMMR cohort (Table 2) offer superior statistical fit during the trial period, they introduce a clinical logic violation: the long-term extrapolated PFS curve for the placebo + CT arm (2-knot hazards) can cross and exceed the pembrolizumab + CT arm (1-knot hazards) in the tail.

During the appraisal consultations, the company pointed out this logical inconsistency, arguing that it is clinically implausible for the control arm to have superior PFS at any point given the higher response rate and observed treatment effect in the pembrolizumab arm. The NHS England Cancer Drugs Fund clinical lead supported this view. Although the EAG defended the crossing as an artifact of long-term extrapolation uncertainty that did not invalidate the observed period fit, the NICE Appraisal Committee ultimately mandated that the placebo PFS curve be logically capped at the pembrolizumab curve.

In our reconstructed pseudo-IPD deterministic base case, the point-estimate Placebo PFS and Pembro PFS curves remain extremely close in the tail but do not cross. The cap is therefore inactive in the base case (crossover week = NA), yielding identical cost-effectiveness results for capped and uncapped runs. 

However, in HEOR modeling, cost-effectiveness must be evaluated under parameter uncertainty. During Probabilistic Sensitivity Analysis (PSA), we propagate the statistical estimation error of the fitted survival curves. Instead of varying a single parameter by a arbitrary range, we vary the **joint regression parameters (spline coefficients)** of the PFS models for both treatment arms:

- **Pembrolizumab + CT PFS:** The 3 spline coefficients (`gamma0`, `gamma1`, `gamma2`) of the 1-knot hazard spline model.
- **Placebo + CT PFS:** The 4 spline coefficients (`gamma0`, `gamma1`, `gamma2`, `gamma3`) of the 2-knot hazard spline model.

The variation range of these parameters is defined by their **joint multivariate normal distribution (MVN)**:
$$\boldsymbol{\gamma} \sim \mathcal{MVN}(\hat{\boldsymbol{\gamma}}, \boldsymbol{\Sigma})$$
where $\hat{\boldsymbol{\gamma}}$ represents the vector of estimated coefficients (the point estimates or means) and $\boldsymbol{\Sigma}$ is the estimated variance-covariance matrix (VCOV) of the coefficients from the regression model. This approach propagates both the fitting uncertainty of each parameter and their mathematical correlation. Due to the tail sensitivity of flexible spline models, these parameter draws frequently cause the Placebo PFS curve to cross and exceed the Pembro PFS curve. Implementing the logical cap is a **critical structural model guardrail** to ensure that every simulated PSA iteration remains clinically valid.

To demonstrate how this guardrail behaves under joint parameter uncertainty, we implement a mini-PSA loop (100 iterations) sampling the spline parameter vectors from their respective MVN distributions, and evaluate the frequency of crossovers and the cost-effectiveness impact of capping:

```{r pmmr-pfs-capping}
# Helper to run cost-effectiveness trace on a given PSM data frame
run_ce_trace <- function(psm_data) {
  psm_data |>
    group_by(arm) |>
    arrange(time_weeks, .by_group = TRUE) |>
    mutate(
      discount         = discount_factor(time_weeks, ce_params$annual_discount_rate),
      death_increment  = pmax(Dead - lag(Dead, default = 0), 0),
      pf_resource_cost = case_when(
        time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_initial,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          ce_params$weekly_cost_pf_pembro_maint,
        time_weeks <= 104 ~
          ce_params$weekly_cost_pf_offtreatment,
        TRUE ~
          ce_params$weekly_cost_pf_offtreatment_yr3
      ),
      qaly = (PF * ce_params$utility_pf + PD * ce_params$utility_pd) *
        ce_params$cycle_length_years * discount,
      disease_management_cost = (
        PF * pf_resource_cost + PD * ce_params$weekly_cost_pd
      ) * discount,
      treatment_cost = case_when(
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_combination_weeks ~
          PF * ce_params$weekly_cost_pembro_combination *
            (1 - ce_params$pembro_discount) * discount,
        arm == "Pembro + chemo" & time_weeks <= ce_params$pembro_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_pembro_maintenance *
            (1 - ce_params$pembro_discount) * discount,
        arm == "Placebo + chemo" & time_weeks <= ce_params$comparator_treatment_cap_weeks ~
          PF * ce_params$weekly_cost_comparator_chemo * discount,
        TRUE ~ 0
      ),
      ae_cost = case_when(
        arm == "Pembro + chemo"  & time_weeks == 0 ~ ce_params$one_off_ae_cost_pembro,
        arm == "Placebo + chemo" & time_weeks == 0 ~ ce_params$one_off_ae_cost_comparator,
        TRUE ~ 0
      ),
      terminal_care_cost = death_increment * ce_params$terminal_care_cost * discount,
      total_cost = disease_management_cost + treatment_cost + ae_cost + terminal_care_cost
    ) |>
    ungroup() |>
    group_by(arm) |>
    summarise(
      discounted_qalys = sum(qaly),
      discounted_costs = sum(total_cost),
      .groups = "drop"
    ) |>
    summarise(
      incremental_qalys = discounted_qalys[arm == "Pembro + chemo"] -
        discounted_qalys[arm == "Placebo + chemo"],
      incremental_costs = discounted_costs[arm == "Pembro + chemo"] -
        discounted_costs[arm == "Placebo + chemo"],
      icer = incremental_costs / incremental_qalys,
      .groups = "drop"
    )
}

# --- Deterministic Base Case (Inactive Cap) ---
res_base_det <- ce_incremental_pmmr_spline

# --- Probabilistic Uncertainty Analysis (Mini-PSA) ---
B <- 100
set.seed(42)

# Get spline fits for PFS
fit_pembro_pfs  <- eag_pmmr_spline_fits$fit[eag_pmmr_spline_fits$treatment == "pembro_chemo" & eag_pmmr_spline_fits$endpoint == "PFS"][[1]]
fit_placebo_pfs <- eag_pmmr_spline_fits$fit[eag_pmmr_spline_fits$treatment == "placebo_chemo" & eag_pmmr_spline_fits$endpoint == "PFS"][[1]]

# Sample parameters
draws_pembro_pfs  <- flexsurv::normboot.flexsurvreg(fit_pembro_pfs, B = B)
draws_placebo_pfs <- flexsurv::normboot.flexsurvreg(fit_placebo_pfs, B = B)

# Pull deterministic OS curves (retained as stable background for PFS analysis)
os_pembro_det  <- psm_pmmr_spline |> filter(arm == "Pembro + chemo") |> pull(OS)
os_placebo_det <- psm_pmmr_spline |> filter(arm == "Placebo + chemo") |> pull(OS)

results_psa_uncapped <- vector("list", B)
results_psa_capped <- vector("list", B)
crossovers <- logical(B)

for (b in 1:B) {
  # Calculate PFS survival curves for this draw
  pfs_placebo_b <- 1 - psurvspline(q = ce_time_grid$time_months, gamma = draws_placebo_pfs[b, ], knots = fit_placebo_pfs$knots, scale = "hazard")
  pfs_pembro_b  <- 1 - psurvspline(q = ce_time_grid$time_months, gamma = draws_pembro_pfs[b, ], knots = fit_pembro_pfs$knots, scale = "hazard")
  
  # Check if Placebo PFS > Pembro PFS at any point
  crossovers[b] <- any(pfs_placebo_b > pfs_pembro_b)
  
  # Build psm_data for uncapped
  psm_uncapped_b <- tibble(
    arm = rep(c("Pembro + chemo", "Placebo + chemo"), each = length(ce_time_grid$time_weeks)),
    time_weeks = rep(ce_time_grid$time_weeks, 2),
    time_months = rep(ce_time_grid$time_months, 2),
    PFS = c(pfs_pembro_b, pfs_placebo_b),
    OS = c(os_pembro_det, os_placebo_det)
  ) |>
    mutate(
      PFS  = pmin(PFS, OS),
      PF   = PFS,
      PD   = pmax(OS - PFS, 0),
      Dead = pmax(1 - OS, 0)
    )
  
  # Build psm_data for capped
  pfs_placebo_capped_b <- pmin(pfs_placebo_b, pfs_pembro_b)
  psm_capped_b <- tibble(
    arm = rep(c("Pembro + chemo", "Placebo + chemo"), each = length(ce_time_grid$time_weeks)),
    time_weeks = rep(ce_time_grid$time_weeks, 2),
    time_months = rep(ce_time_grid$time_months, 2),
    PFS = c(pfs_pembro_b, pfs_placebo_capped_b),
    OS = c(os_pembro_det, os_placebo_det)
  ) |>
    mutate(
      PFS  = pmin(PFS, OS),
      PF   = PFS,
      PD   = pmax(OS - PFS, 0),
      Dead = pmax(1 - OS, 0)
    )
  
  results_psa_uncapped[[b]] <- run_ce_trace(psm_uncapped_b)
  results_psa_capped[[b]]   <- run_ce_trace(psm_capped_b)
}

res_psa_uncapped_df <- bind_rows(results_psa_uncapped)
res_psa_capped_df   <- bind_rows(results_psa_capped)

res_psa_uncapped <- tibble(
  incremental_qalys = mean(res_psa_uncapped_df$incremental_qalys),
  incremental_costs = mean(res_psa_uncapped_df$incremental_costs),
  icer = incremental_costs / incremental_qalys
)

res_psa_capped <- tibble(
  incremental_qalys = mean(res_psa_capped_df$incremental_qalys),
  incremental_costs = mean(res_psa_capped_df$incremental_costs),
  icer = incremental_costs / incremental_qalys
)
```

In the deterministic point estimate (base case), the fitted curves do not cross (crossover frequency = 0%). However, under parameter uncertainty (PSA simulation), the placebo PFS curve crosses and exceeds the pembrolizumab curve in **`r sum(crossovers)`%** of the 100 simulated runs. 

The table below summarizes the cost-effectiveness results under deterministic and probabilistic scenarios, demonstrating the stabilizing effect of the committee-mandated logical cap:

```{r pmmr-capping-comparison-table}
bind_rows(
  res_base_det     |> mutate(Scenario = "Deterministic Base Case (Spline, Capped & Uncapped)"),
  res_psa_uncapped |> mutate(Scenario = "Probabilistic PSA Mean (Uncapped Splines)"),
  res_psa_capped   |> mutate(Scenario = "Probabilistic PSA Mean (Capped Placebo PFS)")
) |>
  select(Scenario, incremental_qalys, incremental_costs, icer) |>
  mutate(
    Crossover_Freq = c("0%", paste0(sum(crossovers), "%"), "0%"),
    incremental_qalys = round(incremental_qalys, 3),
    incremental_costs = round(incremental_costs, 0),
    icer = round(icer, 0)
  ) |>
  transmute(
    Scenario,
    `Crossover Freq.` = Crossover_Freq,
    `Incremental QALYs` = incremental_qalys,
    `Incremental Costs (£)` = incremental_costs,
    `ICER (£/QALY)` = icer
  ) |>
  kable(caption = "Cost-effectiveness comparison under deterministic and probabilistic capping scenarios", align = "l")
```

::: {.callout-note}
## A Note on Reconstructed IPD vs. Clinical Trial Database

In our reconstructed pseudo-IPD base case, the deterministic curves stay very close but do not cross, rendering the cap inactive. This is because flexible spline models are highly sensitive to the exact timing of event and censoring patterns in the tail, which are approximated during KM digitization.

However, implementing this capping logic remains a critical HTA requirement. In HEOR practice, this serves as a **structural model guardrail**. During Probabilistic Sensitivity Analysis (PSA) or bootstrap simulations, parameter uncertainty draws will frequently cause the Placebo PFS tail to rise and cross, violating clinical logic. Implementing the cap in the code ensures that every PSA iteration remains logically and clinically valid.
:::


### HTA Decision Interpretation & Consulting Takeaways

In HEOR modeling, the choice of survival models is rarely a purely statistical exercise. The "crossover and capping" debate in the pembrolizumab appraisal teaches several critical consulting lessons regarding the thinking process of HTA bodies:

1. **Primacy of Clinical Logic Over Statistical Fit**: Spline models (like the Royston-Parmar flexible parametric models) offer excellent empirical fit to the trial period due to their flexible knots. However, they lack structural constraints. When statistical optimization yields clinical absurdities—such as a placebo arm outperforming an active immunotherapy arm in the long term—clinical plausibility must serve as a hard boundary condition to restrict statistical extrapolation.
2. **Methodological Pragmatism (The Capping Solution)**: Instead of discarding the spline models entirely (which would require selecting a less fitting parametric distribution in the trial period), the NICE committee accepted the EAG's splines but mandated a post-hoc logical cap. This shows that HTA bodies favor pragmatic structural adjustments (capping or waning) that preserve the integrity of the trial period fit while ensuring safe tail behavior.
3. **The Role of Stakeholder Mediation**: The HTA process is a negotiation. While the EAG initially defended the curve crossing as a negligible mathematical artifact of uncertainty, the company and the Cancer Drugs Fund clinical lead successfully argued that clinical implausibility creates decision uncertainty. The committee's role is to mediate by validating the technical fit of the EAG while enforcing the clinical reality of the clinical lead.
4. **Subgroup-Specific Tail Plausibility**: For the dMMR cohort, the small sample size made complex splines or piecewise models unstable and prone to optimistic bias, leading the EAG to prefer simple parametric models. For the larger pMMR cohort, splines were justified to avoid the company's overly optimistic log-logistic tail, but they required logical capping.
5. **The Divergence Between Model Validity and ICER Impact**: As shown in the probabilistic analysis, the difference in the mean ICER between the capped and uncapped spline models is extremely small (around £8, or ~0.01%). This is because the crossover occurs in the long-term tail where PFS probability in both arms is already very close to zero, and the difference is further minimized by the 3.5% annual discount rate. However, in HTA appraisals, logical corrections are driven by **structural validity and scientific credibility**, not just by their numerical impact on the ICER. A model that allows placebo patients to have better progression-free survival than active treatment patients is structurally flawed, compromising the credibility of the entire submission. HTA committees will reject model structures that allow clinically impossible states, regardless of whether the numerical impact is £8 or £8,000. Therefore, implementing guardrails is a matter of professional model quality assurance.
6. **Strategic Alignment of Clinical Logic and Commercial Interests**: The crossover debate illustrates how clinical critique and commercial submission strategy can align. While the company framed its objection to the crossing curves on clinical and logical grounds, implementing the logical cap mathematically favored the cost-effectiveness case for pembrolizumab. Capping the Placebo PFS at the Pembro PFS curve reduces the Placebo PFS probability in the tail, which shifts simulated patient-time from the progression-free (PF) state to the progressed disease (PD) state. Because the PF health state is assigned a higher utility (0.814) than the PD state (0.720), this shift lowers the cumulative QALYs in the Placebo arm. This increases the incremental QALY benefit ($\Delta E$) in favor of the active treatment, which mathematically **reduces the ICER**. In HTA consulting, identifying areas where clinical realism arguments also happen to strengthen the cost-effectiveness case is a key element of strategic value demonstration.


#### Saving Reconstructed Model Outputs

This final block writes the same output files as the batch script.

```{r save-outputs}
write_csv(fit_summary, file.path(output_dir, "parametric_fit_summary.csv"))
write_csv(parameter_tbl, file.path(output_dir, "parametric_fit_parameters.csv"))
write_csv(extrapolated, file.path(output_dir, "best_fit_extrapolated_survival.csv"))
write_csv(top3_fit_summary, file.path(output_dir, "top3_fit_summary_by_curve.csv"))

tibble(
  output_file = c(
    "parametric_fit_summary.csv",
    "parametric_fit_parameters.csv",
    "best_fit_extrapolated_survival.csv",
    "top3_fit_summary_by_curve.csv"
  ),
  exists = file.exists(file.path(output_dir, output_file))
) |>
  kable()
```



## Appendices {.unnumbered}

### Appendix A: HTA Parameter Traceability & Reference Registry

All data inputs, parameters, and clinical evidence mapped in the de novo R replication model are registered against their primary public HTA and clinical sources:

| Source Document | Reference Scope | Key Evidence & Parameters Abstracted |
|---|---|---|
| **NICE Final Draft Guidance (FDG)** | TA1092, August 2025 | • Clinical recommendation criteria for first-line endometrial cancer.<br>• Definition of the 2-year pembrolizumab treatment stopping rule. |
| **EAG Evaluation Report & Supplementary Papers** | EAG Report Section 4.2 | • Manufacturer base case parametrics vs. EAG-preferred Royston-Parmar splines.<br>• EAG clinical expert estimates for starting age (67.1 years) and routine monitoring costs. |
| **Committee Consultation & Consultation Papers** | NICE Committee Papers | • Market shares for subsequent therapies (pembrolizumab + lenvatinib reallocation).<br>• Health state utilities (KEYNOTE-158 and KEYNOTE-B21) and AE management unit costs. |
| **Pivotal Trial Publication (NEJM)** | KEYNOTE-868 (NRG-GY018) | • Trial protocol, subgroup baseline demographics (mean age 65.4 years).<br>• Subgroup-specific PFS and OS Kaplan-Meier survival curves used for digitization. |
