---
title: "Rebuilding a Published Hypertension Markov Model — and Verifying It"
subtitle: "A four-state cost-effectiveness model from the ACCOMPLISH trial, reconstructed in R and checked against TECH-VER"
author: "Xiaoge Zhang, PhD"
date: today
page-layout: article
toc: true
toc-depth: 2
number-sections: false
format:
  html:
    code-fold: true
    self-contained: false
    grid:
      margin-width: 260px
---

::: {.callout-warning icon=false}
## What this is

A **replication and verification of a published model**, not a new clinical
finding. Nothing here is evidence about how these two antihypertensive regimens
compare; everything here is evidence about how one published comparison behaves
when it is rebuilt from the numbers the paper prints.

> Feng A, Lin W, Li L, Zeng M, Lyu J. **Pharmacoeconomic evaluation of stroke and
> myocardial infarction prevention in hypertensive patients: Markov model based
> on the ACCOMPLISH trial.** *Journal of International Medical Research*.
> 2023;51(12):1–15.
> [doi:10.1177/03000605231214921](https://doi.org/10.1177/03000605231214921) — CC BY-NC.

The rebuild is an independent R implementation of the paper's four-state annual
Markov cohort. The verification follows the five-stage structure of **TECH-VER**
(Büyükkaramikli et al., *PharmacoEconomics* 2019;37:1391–1408), with the
closing questions posed in the language of **AdViSHE** (Vemer et al.,
*PharmacoEconomics* 2016;34:349–361).

This is a methodological appraisal of an academic model. It was supported by a
Chinese provincial research laboratory grant, and its authors declare no
conflict of interest — there is no commercial sponsor and no product position at
stake here. Nothing below should be read as questioning anyone's good faith.
:::

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# R renders non-ASCII in data frames as <U+XXXX> placeholders unless the
# session is in a UTF-8 locale. Quarto does not guarantee one, so pin it.
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
source("R/00_eval.R")   # loads the engine, then the parameterised runner
options(scipen = 999)
kb <- function(x, ...) knitr::kable(x, ...)
```

## What this replication found

```{=html}
<style>
.finding { margin: 0 0 1.35rem 0; }
.finding > p { margin: 0 0 .3rem 0; }
.est { display:block; font-size:.86em; line-height:1.5; color:#5b6068;
       border-left:2px solid #d8dade; padding-left:.7rem; margin-top:.35rem; }
.est b { color:#3d424a; font-weight:600; }
.tablescroll { overflow-x: auto; margin: .4rem 0 .2rem 0; }
.tablescroll table { font-size: .82em; min-width: 52rem; }
.tablescroll td, .tablescroll th { vertical-align: top; }
/* a wide table may scroll; the page body never does */
.column-page > .tablescroll table { min-width: 0; }
</style>
```

::: {.finding}
**1 · The published results reproduce.** QALYs are exact at the precision
printed. On cost, hydrochlorothiazide reconciles immediately; amlodipine only
once the drug price is taken from the paper's own Figure 4, which prints a value
35 CNY a year below the one its Table 2 gives.
[**How it was established** — an independent R rebuild; cycle-10 cohort occupancy
matching the paper's text in six of six cells; the ranges printed on Figure 4
back-solved to their base values.]{.est}
:::

::: {.finding}
**2 · Much of the model cannot be checked from the paper at all.** No analytic
perspective, no base-case discount rate, no observation period for the rate
conversion, no distribution parameters behind the probabilistic analysis, no
account of what the event costs contain — and no validation of any kind
reported.
[**How it was established** — a reporting audit against TECH-VER's completeness
tests, with each absence confirmed across the full text rather than inferred
from a search.]{.est}
:::

::: {.finding}
**3 · The model runs at about three times the event rate its own trial observed.**
ACCOMPLISH's percentages are cumulative over a mean 36-month follow-up and are
used as one-year probabilities. The trial reports only cumulative proportions for
the individual endpoints, but gives its primary composite as a rate per 1000
patient-years as well — the distinction is one it draws itself. Run the model over
the trial's own three years and it returns 11.2% cumulative death where 4.1% went in.
[**How it was established** — a round-trip test on the transition matrix, using
no cost or utility value: the trial figure that went into the model does not come
back out of it.]{.est}
:::

::: {.finding}
**4 · The conclusion survives it.** Correcting the conversion halves the health
gain and narrows the cost gap by a third, and amlodipine still costs less and
delivers more. Everything found here is about the size of the answer, not its
sign.
[**How it was established** — the rebuild re-run on corrected transition
probabilities, engine otherwise untouched.]{.est}
:::

::: {.finding}
**5 · Half the modelled cost is an event charge levied every year**, from a
source that could not be obtained. The MI charge is applied on average 3.9 times
per patient who has one; if it is a first-year cost, the cost gap falls to a
third of the published one.
[**How it was established** — decomposing the rebuilt totals into drug and
event-state components, then re-running with the event figures charged once.]{.est}
:::

::: {.finding}
**6 · The paper's own table and Methods disagree about where the utilities came
from.** Table 3 attributes two of them to ACCOMPLISH, which reports no utility
outcome of any kind. The Methods name two economic evaluations instead; the two
attributions overlap in only one reference, and the trial the table cites is
absent from the Methods sentence entirely.
[**How it was established** — a full-text check of the trial, read against
Table 3's reference marks and the Methods sentence.]{.est}
:::

**The bottom line.** The analysis is competently built and it replicates. Its
*direction* — that amlodipine plus benazepril is the more economical of the two
regimens in China — survives every correction made here, including the largest.
Its *magnitudes* do not, and much of the model cannot be checked from the paper
at all. **The conclusion is usable; the numbers should not be quoted.**

# 1 · The model

A cohort of high-risk hypertensive patients — the ACCOMPLISH population, median
age 68, 60% diabetic, 23% with prior myocardial infarction — moves through four
health states in annual cycles for ten years. Two of those states are the events
the treatments exist to prevent: **myocardial infarction (MI, heart attack)** and
**stroke**.

::: {.column-page}
![The four states and every transition the model allows, with the annual probabilities as published. **Colour is the arm throughout this page** — blue for amlodipine, orange for hydrochlorothiazide. Where a transition carries two figures the arms differ there, which happens only on the three edges leaving the event-free state; everywhere else the two arms are identical. The self-loops are where most of the cohort actually sits.](visuals/state_diagram.png){#fig-states}
:::

Three structural choices deserve naming, because each is a modelling decision
rather than a piece of trial evidence.

- **No recurrent events, and no transition between MI and stroke.** A patient who
  has an infarction can only stay in that state or die — never have a second
  infarction, and never go on to have a stroke.
- **Post-event mortality is identical in both arms** (0.0611 from MI, 0.2381 from
  stroke). Treatment is modelled as acting only on *whether* an event happens,
  never on what happens afterwards.
- **The horizon is ten years, not lifetime.** Whatever befalls the patients still
  alive at cycle 10 is outside the model.

All three are ordinary simplifications. Whether the paper declares them, and
whether they hold up, is §3.

## The transition matrices

```{r inputs}
#| output: asis
mat <- function(arm) {
  p <- TPM_PUBLISHED[[arm]]; d <- "&mdash;"
  f <- function(x) sprintf("%.4f", x)
  m <- data.frame(
    From = c("Event-free", "Non-fatal MI", "Non-fatal stroke", "Death"),
    `Event-free` = c(f(p[["EE"]]), d, d, d),
    `Non-fatal MI` = c(f(p[["EM"]]), f(p[["MM"]]), d, d),
    `Non-fatal stroke` = c(f(p[["ES"]]), d, f(p[["SS"]]), d),
    Death = c(f(p[["ED"]]), f(p[["MD"]]), f(p[["SD"]]), "1.0000"),
    check.names = FALSE)
  rs <- c(sum(p[c("EE", "EM", "ES", "ED")]), sum(p[c("MM", "MD")]),
          sum(p[c("SS", "SD")]), 1)
  m$`Row sum` <- ifelse(abs(rs - 1) < 1e-9, sprintf("%.4f", rs),
                        sprintf("**%.4f**", rs))
  m
}
nm <- c(amlodipine = "Amlodipine", hctz = "Hydrochlorothiazide")
for (a in c("amlodipine", "hctz")) {
  print(knitr::kable(mat(a), escape = FALSE, align = "lrrrrr",
    caption = sprintf("**%s + benazepril.** Columns are the state moved *to*; a dash is a transition the model does not allow.", nm[[a]])))
  cat("\n")
}
```

Laid out as matrices, those structural choices *are* the dashes: the MI and
stroke rows lead nowhere except back to themselves or on to death, and the death
row leads only to itself. The event-free row sums to **1.0001** in the amlodipine
arm and **0.9999** in the hydrochlorothiazide arm, so the printed matrix is not
quite the matrix that was run.

## Costs and utilities

```{r costs-utils}
kb(data.frame(
  State = STATES,
  `Annual cost, amlodipine arm (CNY)` = c(COST_DRUG[["amlodipine"]],
    COST_EVENT[["MI"]], COST_EVENT[["Stroke"]], 0),
  `Annual cost, HCTZ arm (CNY)` = c(COST_DRUG[["hctz"]],
    COST_EVENT[["MI"]], COST_EVENT[["Stroke"]], 0),
  Utility = unname(UTIL[STATES]), check.names = FALSE),
  row.names = FALSE,
  caption = "Tables 2 and 3 of the paper. Every figure in Table 2 sits under the header Assignment/RMB/year.")
```

There is no micro-costing anywhere in this model. Cost enters as **state
occupancy × an annual state cost** — four numbers for the whole analysis. The
drug figures are a daily median tender price scaled to a year, with no dosing,
adherence, monitoring or follow-up resource use behind them; the two event
figures are taken whole from another study and adjusted to 2021 prices.

## Time, discounting and software

| | |
|---|---|
| Cycle | 1 year |
| Horizon | 10 cycles, not lifetime |
| Discount rate | 5% — recovered rather than reported; see §3 |
| Half-cycle correction | not applied, and not mentioned in the paper |
| Software | TreeAge Pro 2021 |

::: {.callout-note collapse="true" appearance="simple"}
## The data extraction document — every input, its source, and whether it checks out

Every transition probability traces to one table in the ACCOMPLISH trial through
a conversion the paper sets out in its Methods.

```{r chain}
kb(do.call(rbind, lapply(names(TRIAL), function(a) {
  t <- TRIAL[[a]]; p <- TPM_PUBLISHED[[a]]
  data.frame(Arm = a,
    `ACCOMPLISH MI` = sprintf("%.1f%%", 100 * t[["mi"]]),
    `1-exp(-p)` = sprintf("%.4f", 1 - exp(-t[["mi"]])),
    `Published PH-M` = sprintf("%.4f", p[["EM"]]),
    `ACCOMPLISH stroke` = sprintf("%.1f%%", 100 * t[["stroke"]]),
    `1-exp(-p) ` = sprintf("%.4f", 1 - exp(-t[["stroke"]])),
    `Published PH-S` = sprintf("%.4f", p[["ES"]]), check.names = FALSE) })),
  row.names = FALSE, caption = "The conversion, reproduced.")
```

The event-free death probability then follows by subtraction — all-cause annual
death less the deaths already routed through the two event states — reproducing
0.0344 and 0.0369 to four decimals. **The parameter chain is fully recovered.**
Whether the conversion itself is sound is a separate question, taken up in §3.

**The utility rows need reading together.** Table 3 marks the event-free and MI
utilities to reference 13 — the ACCOMPLISH trial, which reports no utility,
EQ-5D, SF-36 or quality-of-life outcome anywhere — and the stroke utility to
reference 14. The paper's Methods say something different: that the utilities "were
derived from published studies of the same or similar populations", citing
references **14 and 19**. Reference 13 does not appear in that sentence.

So the table and the text disagree, and the source the table names does not hold
the values. Both papers those Methods point to are Chinese-language and were not
obtainable here, which is exactly what a reviewer would need to settle it.
Recorded as a citation-consistency problem, not a data-integrity one.

The table below is the **data extraction document** for this rebuild: one row per
input, the source the paper cites for it, where that quantity actually sits in
the cited work, any transformation applied, and whether the source could be
located and checked. It is written for a reviewer who should not have to repeat
the search to judge whether an input is appropriate.

::: {.tablescroll}
```{r provenance}
kb(read.csv("data/provenance.csv", check.names = FALSE))
```
:::
:::

# 2 · Results

**Amlodipine plus benazepril costs less and delivers more.** Over ten years it
accrues about 8,400 CNY less and a tenth of a QALY more per patient, so it
**dominates** — there is no incremental cost-effectiveness ratio to compute, and
none that could be interpreted. That is the paper's conclusion, and the rebuild
reproduces it.

```{r recon}
reb <- run_both(drug_in = "event_free", event_cost = "prevalent")
reb <- reb[match(TARGET$arm, reb$arm), ]
kb(data.frame(
  Arm = TARGET$arm,
  `Cost, published` = sprintf("%.2f", TARGET$cost),
  `Cost, rebuilt`   = sprintf("%.2f", reb$cost),
  `Difference`      = sprintf("%+.2f%%", 100 * (reb$cost - TARGET$cost) / TARGET$cost),
  `QALY, published` = sprintf("%.2f", TARGET$qaly),
  `QALY, rebuilt`   = sprintf("%.4f", reb$qaly),
  `Difference `     = sprintf("%+.3f%%", 100 * (reb$qaly - TARGET$qaly) / TARGET$qaly),
  check.names = FALSE),
  caption = "Cell-by-cell reconciliation against Table 7.")
```

QALYs match to the precision the paper prints, and the incremental cost of
8,391.53 reproduces exactly. The 0.39% on the amlodipine total is not rounding;
it is traced to a specific input in §3, and closing it makes amlodipine slightly
cheaper still.

The paper nonetheless reports an ICER of −64,550.23 alongside a decision entry
of "Advantage plan". A negative ratio under dominance has no reading; §3 takes
that up, along with a second value the paper gives for the same quantity.

Everything below comes from the independent R rebuild. Two accounting rules and
the discount rate are not fully specified in the paper and had to be fitted to
the published totals; that is set out at the end of the page.

The paper carries five result figures, and it is worth being precise about what
each figure below is:

| | |
|---|---|
| @fig-trace | **reproduces** the paper's Figures 2 and 3 |
| @fig-tornado | answers the question of its Figure 4, on a **different outcome** — the one it uses has no interpretation under dominance |
| @fig-psa | **cannot** reproduce its Figures 5 and 6, because the parameters those need are not published; built here on stated assumptions instead |

Nothing on this page is copied from the paper. Every number and every mark is
computed from the published inputs by the code behind it, which is folded into
the page and can be opened chunk by chunk.

## The cohort over time

::: {.column-page}
![Where the cohort sits in each cycle, under the model as published. This reproduces the content of the paper's Figures 2 and 3, drawn as stacked areas rather than its four separate lines so that the four states are visibly one cohort. States take a sequential ramp because they are ordered by severity; the categorical blue and orange stay reserved for the arms.](visuals/cohort_trace.png){#fig-trace}
:::

By cycle 10 the amlodipine cohort is 45.9% event-free, 11.6% post-MI, 4.5%
post-stroke and 38.1% dead. The paper's own Results text reports 11.6%, 4.5% and
38.0% for the same three states — a match to the last digit it prints.

## Uncertainty

::: {.column-page}
![Probabilistic sensitivity analysis: 1000 simulations run on the rebuilt model. **Built here, not reproduced from the paper** — its Figures 5 and 6 depend on distribution parameters it does not report, so the spreads are mine, each set so the parameter's 95% interval matches the range the paper itself used in its one-way analysis. That choice sets the width of the cloud; it does not move the point estimate.](visuals/psa.png){#fig-psa}
:::

Under those assumptions amlodipine is cheaper *and* more effective in 87% of
draws, and cost-effective at the paper's own threshold in 98%. The 13% that fall
outside the dominant quadrant are worth naming: dominance is the base-case
result, not a certainty.

The reconstruction can be partly checked against the paper's own figures even
though it cannot reproduce them. At zero willingness to pay — where the question
reduces to "is amlodipine cheaper?" — the paper's Figure 5 reads **88%** and this
one gives **89%**, so the assumed spread on the cost side is close to whatever
the paper used. The effect side is not: the paper's Figure 6 runs its horizontal
axis from −0.30 to 0.50 incremental QALYs, while this cloud spans −0.06 to 0.31.
**The paper's effect uncertainty is materially wider than what these assumptions
generate**, which is the honest limit of a reconstruction: it can be anchored,
but not matched.

::: {.column-page}
![One-way sensitivity analysis, plotted twice from the same runs. The left panel uses the ICER, as the paper's Figure 4 does, and **reproduces its rank order on all sixteen parameters**. The right panel uses incremental net monetary benefit at the paper's own threshold, which is what a dominant comparison can actually be read on. Each arm's transition probabilities are moved separately, as the parameter names on the paper's figure show it does; ranges are the ones that figure prints.](visuals/tornado.png){#fig-tornado}
:::

**No parameter, moved to either end of its stated range, takes the net benefit
below zero.** The base-case conclusion is stable inside the ranges the paper
itself tested.

**The left panel is the strongest single verification test in this case.** Run
the same analysis on the paper's own axis and the rank order comes back
identical on all sixteen parameters, from the two event-free-to-death
transitions at the top to MI-to-death at the bottom. The bar extents line up
too: the widest run spans −220,000 to −34,000 against roughly −240,000 to
−35,000 as printed. A ratio with an effect difference of 0.125 in its
denominator amplifies any disagreement in the engine that produced it; there is
none to amplify.

**Which is why the two panels look nothing alike, and why that is not a
disagreement.** A tornado ranks parameters by the outcome it plots. An ICER
explodes as its denominator approaches zero, so anything that moves survival
dominates it — hence the long left tails, which are the ratio misbehaving rather
than the model being uncertain. Net benefit is linear in cost and in effect, so
the cost inputs lead instead, and the discount rate moves from near-invisible to
mid-table. Same model, same ranges, two defensible pictures.

⚠ The left panel also shows why the paper's own axis is the wrong one to decide
on: under dominance the ICER has no interpretation, and a bar that runs from
−220,000 to −34,000 tells a reader nothing about whether the conclusion holds.
The right panel does, and it never reaches zero.

# 3 · Verification and appraisal

Verification asks a narrow question: **does the model do what its description
says it does?** **TECH-VER** sets out a checklist for answering it, in five
stages worked in order — each one worth starting only once the stage before it
holds.

1. **Model inputs** — do the numbers going in match their sources, and each other?
2. **Event and state calculations** — does the cohort move between states correctly?
3. **Result calculations** — are the totals, the incremental figures and the decision rule right?
4. **Uncertainty analysis** — are the sensitivity analyses correctly built, and reported well enough to check?
5. **Overall validation** — does the model reproduce anything that can be verified from outside it?

Within each stage the checks escalate by how much work they cost. **Black-box
tests come first**: push an input to an extreme and see whether the output does
what it must — set every utility to zero and no QALYs should accrue; set
mortality to zero and nobody should die. They need no sight of the code, they
take minutes, and they catch a surprising share of what goes wrong. Only what
survives them is worth the expensive work: **white-box** inspection, reading the
calculation line by line, and then **replication**, rebuilding the model
independently to see whether the same numbers come back.

This page is a replication, so all three are open to it. Most of what follows is
black-box work even so, because that is where the cheap tests happened to bite.

## Stage 1 — the inputs

The model runs on twenty-four numbers. They fall into four groups, and the
verification result is different for each.

- **Sixteen transition probabilities.** Every one is derived inside the paper
  from the ACCOMPLISH trial, and the derivation reconciles exactly — all six
  annual probabilities and both event-free death probabilities come back to four
  decimals. **The arithmetic is right; the input to it looks wrong.** The trial
  percentages it starts from are cumulative over a mean follow-up of 36 months,
  and they are converted as though they were annual — the paper prints a
  conversion formula carrying an observation period and never gives that period
  a value. On the face of it that inflates every event rate in the matrix about
  threefold. **Stage 5 tests it rather than argues it.**

- **Four costs.** Two drug prices and two event-state costs. **None can be checked
  against its source**: the drug prices cite a tender website with no version or
  date, and the event costs cite a Chinese-language article that could not be
  obtained. One of the two drug prices also disagrees with the paper's own
  Figure 4, which Stage 3 takes up.
- **Four utilities.** **The paper's Table 3 and its Methods cite different sources
  for them**, and the source the table names contains no utility data at all. Set out
  in §1.
- **The settings around them** — cycle length, horizon, discount rate,
  perspective, half-cycle correction. Two are stated; the rest are not reported
  at all, which is collected below.

### What the paper does not report

Verification keeps running into the same obstacle, so it is worth setting out in
one place. None of the following appears anywhere in the paper — not in the
its Methods, its tables, its figures, its Discussion or its limitations.

- the **analytic perspective** — payer, health system or societal is never said,
  and which costs belong in scope follows from it
- the **base-case discount rate**, and whether costs and outcomes are discounted
  alike
- the **observation period $t_1$** in the rate conversion, and **ACCOMPLISH's
  follow-up duration** — the two things a reader would need to check it
- whether **drug cost continues after an event**, whether the **event figures are
  annual or one-off**, and **what resource use they contain**
- the **PSA distribution parameters** — the families are given, the numbers are not
- **half-cycle correction**
- that the model **forbids recurrent events** and holds **post-event mortality
  equal across arms** — two structural restrictions with real consequences over
  ten years
- **why ten years rather than a lifetime horizon**, which ISPOR-SMDM asks for
  where an intervention affects mortality
- any **model validation** — no face validity exercise, no internal verification,
  no external check against the trial, no cross-model comparison

On the two structural restrictions, it is worth being exact about what "not
declared" means here, because the paper does declare three assumptions of its
own. They are, in full:

> *"(1) the probability of occurrence of various events in the two patient groups
> at the end of the trial is the same; (2) the health status transition
> probability in each cycle is the same as the probability of clinical trial
> events; (3) because the cost of adverse drug reactions is low, the analysis
> does not include the occurrence of adverse reactions."*

The first is about event probabilities persisting past the trial, the second is
the conversion discussed above, the third is about adverse events. **Neither the
ban on recurrent events nor the equal post-event mortality appears in that list,
or anywhere else in the paper.** They are visible only in the transition matrix,
where the MI and stroke rows each sum to one across just two destinations.

## Stage 2 — the engine

The Results text reports the state distribution at cycle 10 for both arms — an
independent check that constrains the trace without passing through a single
cost or utility value.

```{r trace-check}
pub <- data.frame(arm = c("amlodipine", "hctz"),
                  Dead = c(38.0, 41.0), Stroke = c(4.5, 5.1), MI = c(11.6, 13.8))
reb <- do.call(rbind, lapply(c("amlodipine", "hctz"), function(a) {
  tr <- run_markov(a)$trace
  data.frame(arm = a, Dead = 100 * tr["10", "Dead"],
             Stroke = 100 * tr["10", "Stroke"], MI = 100 * tr["10", "MI"])
}))
p1 <- function(x) sprintf("%.1f%%", x)
kb(data.frame(Arm = pub$arm,
              `Dead, published` = p1(pub$Dead),  `Dead, rebuilt` = p1(reb$Dead),
              `Stroke, published` = p1(pub$Stroke), `Stroke, rebuilt` = p1(reb$Stroke),
              `MI, published` = p1(pub$MI), `MI, rebuilt` = p1(reb$MI),
              check.names = FALSE),
   caption = "Where the cohort sits at cycle 10, as reported in the paper's Results text and as rebuilt.")
```

Six cells, six matches to the last digit printed. The nine black-box tests
TECH-VER specifies for this stage also pass: cohort conservation, non-negativity,
monotone deaths, utilities set to one and to zero, costs set to zero, mortality
set to zero. **The transition engine is correct.**

## Stage 3 — the results

The reconciliation is in §2. QALYs are exact at the precision printed and the
incremental cost of 8,391.53 reproduces exactly; the amlodipine total comes out
0.39% high.

That residual is a reporting inconsistency rather than arithmetic. Fourteen of
the fifteen parameter ranges printed on the paper's Figure 4 back-solve to the
value in Tables 2 and 3. The fifteenth does not: the amlodipine drug price
implies **5,459.50** where Table 2 prints **5,494.53**. Run on the figure's
value, the arm total falls to 0.07% of the published one — within what the
transition matrix's four printed decimals allow.

```{r drugcost}
#| code-fold: true
set.seed(20260826); HALF <- 0.00005
srow <- function(q) repeat { v <- q + runif(length(q), -HALF, HALF)
  v[length(v)] <- 1 - sum(v[-length(v)])
  if (abs(v[length(v)] - q[length(q)]) <= HALF) return(v) }
band <- function(drug) replicate(3000, {
  q <- TPM_PUBLISHED$amlodipine
  e <- srow(c(q[["EE"]], q[["EM"]], q[["ES"]], q[["ED"]]))
  m <- srow(c(q[["MM"]], q[["MD"]])); k <- srow(c(q[["SS"]], q[["SD"]]))
  t <- TPM_PUBLISHED
  t$amlodipine <- c(EE=e[1], EM=e[2], ES=e[3], ED=e[4], MM=m[1], MD=m[2], SS=k[1], SD=k[2])
  run_model(tpm = t, drug = replace(COST_DRUG, "amlodipine", drug))$cost[["amlodipine"]] })
fig <- (4094.6 + 6824.4) / 2
kb(do.call(rbind, lapply(c(COST_DRUG[["amlodipine"]], fig), function(v) {
  tot <- run_model(drug = replace(COST_DRUG, "amlodipine", v))$cost[["amlodipine"]]
  bd <- band(v)
  data.frame(`Amlodipine drug cost` = sprintf("%.2f", v),
    Source = if (v > 5470) "Table 2" else "implied by Figure 4",
    `Rebuilt arm total` = sprintf("%.2f", tot),
    `vs published 66,196.97` = sprintf("%+.2f  (%+.2f%%)", tot - 66196.97,
                                       100 * (tot - 66196.97) / 66196.97),
    `Within four-decimal rounding` = if (66196.97 >= min(bd) && 66196.97 <= max(bd))
      "yes" else "no", check.names = FALSE) })), row.names = FALSE,
  caption = "The amlodipine total on each of the two drug prices the paper gives.")
```

A documented value and an implemented value differing by 35 CNY a year, findable
only by reading a figure's small print against a table. Nothing material turns on
it; every number elsewhere on this page uses Table 2's.

Dominance is reported as a ratio. Amlodipine is cheaper and more effective, so
the correct entry is *dominated*; Table 7 prints an ICER of −64,550.23 next to a
decision of "Advantage plan". The paper also gives a second value for the same
quantity — Figure 4 is centred on −67,133.56 — because Table 7 divides by the
effect difference **as printed**, 0.13, while the figure uses the unrounded
0.125.

## Stage 4 — uncertainty

The paper gives distribution families — normal for costs, beta for probabilities
and rates — but no standard deviations and no shape parameters, so its
probabilistic analysis cannot be reproduced; @fig-psa is a reconstruction on
stated assumptions, not a replication. The one-way analysis is a different matter: its
Figure 4 prints every parameter's name and its low and high value on the face of
the figure, which is better reporting than most published tornados manage and is
what made @fig-tornado checkable at all.

Worth noting for its own sake: those ranges perturb input *values*. The two
findings in Stages 1 and 5 are structural, and nothing inside the stated ranges
reaches them.

## Stage 5 — validation

The model's death transitions were derived *from* ACCOMPLISH's three-year
mortality, so running the model over ACCOMPLISH's own three years should give
that mortality back. It does not.

::: {.column-page}
![Cumulative all-cause death in the modelled cohort. Colour is the arm as before; **line style is the version** — solid as published, dashed with the trial percentages annualised over the 36 months actually observed. The shaded band is the trial's own follow-up, the only stretch where the model can be checked against evidence rather than extrapolated. The black marker is what ACCOMPLISH observed.](visuals/mortality_validity.png){#fig-validity}
:::

```{r validity}
kb(do.call(rbind, lapply(names(TRIAL), function(a) {
  run3 <- function(p) { m <- make_tpm(p)
    x <- matrix(0, 11, 4, dimnames = list(0:10, STATES)); x[1, "EventFree"] <- 1
    for (t in 1:10) x[t + 1, ] <- x[t, ] %*% m; x }
  pu <- run3(TPM_PUBLISHED[[a]]); co <- run3(corr_tpm(a))
  data.frame(Arm = a,
    `Trial observed` = sprintf("%.1f%%", 100 * TRIAL[[a]][["death"]]),
    `Model at cycle 3, as published` = sprintf("%.1f%%", 100 * pu["3", "Dead"]),
    `Model at cycle 3, annualised` = sprintf("%.1f%%", 100 * co["3", "Dead"]),
    check.names = FALSE) })), row.names = FALSE,
  caption = "The round trip. Column two is what went into the model; column three is what comes back out of it over the same three years; column four is what comes back once the conversion uses the period the trial actually ran.")
```

The round trip does not close. What went in as 4.1% comes back out as 11.2%.
Correcting the conversion closes it — 4.1% in, 4.1% out — which is what a sound
conversion has to do. Seven of the model's
ten years sit outside anything the trial observed — ordinary for a Markov model,
and not a criticism. What matters is the other three.

### Does any of it change the answer?

```{r survives}
#| code-fold: true
TPM_SAVE <- TPM_PUBLISHED
TPM_PUBLISHED <- lapply(names(TRIAL), corr_tpm); names(TPM_PUBLISHED) <- names(TRIAL)
cc <- run_both(drug_in = "event_free", event_cost = "prevalent")
cc <- cc[match(TARGET$arm, cc$arm), ]
TPM_PUBLISHED <- TPM_SAVE
kb(data.frame(
  Version = c("As published", "Conversion corrected"),
  `Incremental cost (CNY)` = sprintf("%+.0f", c(diff(TARGET$cost), diff(cc$cost))),
  `Incremental QALY` = sprintf("%+.4f", c(diff(TARGET$qaly), diff(cc$qaly))),
  `Amlodipine dominant` = c("yes", "yes"), check.names = FALSE), row.names = FALSE,
  caption = "Hydrochlorothiazide relative to amlodipine.")
```

Not the direction. Correcting the conversion halves the health gain and narrows
the cost gap by a third, and amlodipine still costs less and delivers more. Two
other things were tested for the same reason and behave the same way: charging
the event costs once rather than annually cuts the cost gap to a third of the
published one, and hydrochlorothiazide's price would have to fall 23.6% — below
amlodipine's — before the two arms cost the same.

**Everything found here is about the size of the answer, not its sign.** That is
worth being plain about, because it is the difference between a model that is
wrong and a model whose numbers should not be quoted.

## The checklist

This is the full record of what was checked on **the paper's model** — whether
it does what the paper says it does, and whether its inputs and its reporting
hold up. It is not a judgement on the clinical question.

The rows come from two places. Most are TECH-VER's own items, adapted to a
four-state cohort model: the black-box tests each stage prescribes, the
completeness and consistency tests that precede them, and the result and
uncertainty checks. The rest are checks the rebuild itself generated — questions
that could only be asked once there was a working model to ask them of, such as
whether the trial figures come back out of it, or whether the inputs printed in
the tables are the inputs it ran on. TECH-VER items that assume a spreadsheet
implementation — hidden sheets, external links, password protection — are left
out, because the model is a TreeAge file that was not available.

This is the **model checklist** for the rebuild. Each row carries one of four
verdicts — `pass`, `not pass`, `N/A`, `cant_verify` — and a single line of
reason. `cant_verify` marks a check that could not be completed, which is not
the same as one that failed; four rows carry it, and each names what was
missing.

```{r checklist-summary}
#| code-fold: true
ck <- read.csv("data/techver_checklist.csv", check.names = FALSE)
kb(as.data.frame(t(table(factor(ck$Verdict,
     levels = c("pass", "not pass", "cant_verify", "N/A")))))[, -1],
   col.names = c("Verdict", "Items"), row.names = FALSE)
```

::: {.callout-note collapse="true" appearance="simple"}
## The model checklist — all `r nrow(ck)` checks, with verdicts

::: {.tablescroll}
```{r checklist}
#| echo: false
kb(ck)
```
:::
:::

## Scope, and how this was built

An independent implementation in R of the model the paper describes, built from
the inputs the paper prints and from the trial it cites. Every implementation
choice — cycle timing, discount indexing, which states carry which cost — is a
switch rather than a hard-wired constant, so each could be tested against the
published totals rather than assumed.

**Covered.** The deterministic base case in full, the derivation of every
transition probability from the trial, the provenance of every input, and the
one-way analysis on the paper's own ranges. Three sensitivities were run far
enough to test the conclusion rather than describe it: the annualisation, the
basis of the event costs, and the drug price.

**Not covered.** The probabilistic analysis cannot be reproduced, because the
paper publishes distribution families and no parameters; @fig-psa is built on
stated assumptions instead. Cross-model comparison against other hypertension
models is validation rather than verification, and was declared out of scope
rather than attempted.

**Not obtainable.** Two Chinese-language articles — reference 14, cited for the
stroke utility and both event-state costs, and reference 19, named in the
Methods as a source of utilities but absent from Table 3. Neither could be
retrieved, so the rows depending on them are recorded as `cant_verify` rather
than assessed. The paper's own TreeAge file was likewise unavailable.

**Three inputs had to be fitted.** Whether drug cost continues after an event,
whether the MI and stroke figures are annual or one-off, and the base-case
discount rate are not reported. To make the rebuild runnable each was fitted to
the published totals, and one setting fits in each case: drug cost accrues only
while event-free, the event figures are annual state costs, the discount rate is
5%. The transition-probability findings do not use these settings.
