---
title: "Pairwise Meta-Analysis: The Foundation of NMA"
author: "Xiaoge Zhang, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "View the code"
    number-sections: true
---

*Pooling direct evidence for one comparison, from the likelihood up*

## Executive Summary: One Comparison, Many Trials

A pairwise meta-analysis pools the results of several trials that all compare the **same two treatments**, say a new drug B against standard of care A. Each trial on its own is a noisy estimate of the treatment effect; pooled, they give one estimate with a narrower interval.

It is also the building block of every method in this library. A [Standard Bayesian NMA](01_Standard_Bayesian_NMA_Guide.qmd) is several pairwise meta-analyses estimated together, tied by the consistency assumption so that evidence on A vs B and A vs C also informs B vs C. Everything an NMA adds sits on top of the two questions answered here:

1.  How should trials of different sizes be **weighted**?
2.  What if the trials do not all estimate the **same** effect?

## Theoretical Framework: From a 2×2 Table to a Pooled Estimate

### The Effect Size in a Single Trial

For a binary outcome such as response, each trial reduces to a 2×2 table:

|             | Responders | Non-responders |
|:------------|:----------:|:--------------:|
| Treatment B |    $a$     |      $b$       |
| Treatment A |    $c$     |      $d$       |

Index the trials by $i = 1, \dots, k$ (in the mock case below, $k = 5$). In trial $i$, arm B enrols $n_B$ patients and arm A enrols $n_A$. The responder counts are random variables, $a \sim \text{Binomial}(n_B, p_B)$ and $c \sim \text{Binomial}(n_A, p_A)$, where $p_B$ and $p_A$ are the true response rates; the non-responders are $b = n_B - a$ and $d = n_A - c$. The subscript $i$ is dropped from these quantities to keep the formulas readable.

The odds ratio and its logarithm are

$$OR_i = \frac{a / b}{c / d} = \frac{ad}{bc}, \qquad y_i = \ln OR_i$$

Meta-analysis works on the log scale because the OR is bounded below by 0 and asymmetric around 1, whereas $\ln OR$ is symmetric around 0 and close to Normal in reasonably sized trials.

The variance of $y_i$ is

$$v_i = \frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}$$

Each arm contributes the reciprocals of its own two counts. Larger trials have larger counts, hence smaller $v_i$.

::: {.callout-note collapse="true"}
# Where the variance formula comes from

**Step 1: the response rate in one arm.** Take arm B, and write $n = n_B$ and $p = p_B$ for short. Since $a \sim \text{Binomial}(n, p)$, the observed rate $\hat p = a / n$ has variance

$$\text{Var}(\hat p) = \frac{p(1-p)}{n}$$ {#eq-var-p}

**Step 2: from the rate to the log odds.** The log OR is built not from $\hat p$ but from its log odds, $g(\hat p) = \ln\frac{\hat p}{1 - \hat p} = \ln\frac{a}{b}$. The **delta method** handles this change of scale. Near $p$, the curve $g$ is close to its tangent line:

$$g(\hat p) \approx g(p) + g'(p)\,(\hat p - p)$$ {#eq-taylor}

On the right of ([-@eq-taylor]), $g(p)$ and $g'(p)$ are fixed numbers; only $\hat p$ is random. Multiplying a random quantity by a constant multiplies its variance by the constant squared, and adding a constant leaves it unchanged, so

$$\text{Var}\big(g(\hat p)\big) \approx g'(p)^2 \, \text{Var}(\hat p)$$ {#eq-delta}

**Step 3: the slope of the log odds.** Writing $g(p) = \ln p - \ln(1 - p)$ and differentiating:

$$g'(p) = \frac{1}{p} + \frac{1}{1-p} = \frac{1}{p(1-p)}$$ {#eq-slope}

**Step 4: combine.** Substituting ([-@eq-slope]) and ([-@eq-var-p]) into ([-@eq-delta]):

$$\text{Var}\left(\ln\frac{a}{b}\right) \approx \frac{1}{p^2(1-p)^2} \cdot \frac{p(1-p)}{n} = \frac{1}{np(1-p)}$$ {#eq-var-logodds}

**Step 5: back to the counts.** Because $(1 - p) + p = 1$, the right-hand side of ([-@eq-var-logodds]) splits into two fractions:

$$\frac{1}{np(1-p)} = \frac{(1 - p) + p}{np(1-p)} = \frac{1}{np} + \frac{1}{n(1-p)}$$ {#eq-split}

In ([-@eq-split]), $np$ and $n(1-p)$ are the expected numbers of responders and non-responders. Replacing them with the observed counts $a$ and $b$ gives the variance for arm B:

$$\text{Var}\left(\ln\frac{a}{b}\right) \approx \frac{1}{a} + \frac{1}{b}$$ {#eq-arm-B}

Steps 1 to 5 apply unchanged to arm A, with its counts $c$ and $d$:

$$\text{Var}\left(\ln\frac{c}{d}\right) \approx \frac{1}{c} + \frac{1}{d}$$ {#eq-arm-A}

**Step 6: two arms.** The log OR is a difference of the two log odds, $y_i = \ln\frac{a}{b} - \ln\frac{c}{d}$. For independent quantities, $\text{Var}(X - Y) = \text{Var}(X) + \text{Var}(Y)$, and the two arms enrol different patients, so adding ([-@eq-arm-B]) and ([-@eq-arm-A]):

$$v_i = \text{Var}(y_i) \approx \frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}$$ {#eq-vi}

which is the formula above.

**What it implies.** The smallest cell dominates ([-@eq-vi]). A trial of 1,000 patients per arm with only 5 responders on A still has $v_i > 1/5$, so a large trial with a rare event can be less precise than a small trial with a common one. When a cell is zero, $v_i$ is infinite, which is why a continuity correction, typically adding 0.5 to every cell, is applied. That correction is itself an assumption, and one that can move results when events are sparse.
:::

### Fixed Effect: Inverse-Variance Weighting

The **fixed-effect** model assumes every trial estimates one common effect $\theta$, and differs from it only through sampling error:

$$y_i \sim N(\theta, v_i), \qquad i = 1, \dots, k$$

treating each $v_i$ as known. Strictly, the variance of $y_i$ depends on the true response rates $p_B$ and $p_A$, which are unknown; $v_i$ replaces them with the observed rates, so it is an estimate. Treating it as a known constant is an approximation that works well with large counts, but with few events it makes both the weights and the pooled estimate unreliable, because $y_i$ and $v_i$ are computed from the same counts.

The log-likelihood is

$$\ell(\theta) = -\frac{1}{2} \sum_{i=1}^{k} \frac{(y_i - \theta)^2}{v_i} + \text{const}$$

Setting its derivative to zero gives the maximum likelihood estimate:

$$\sum_{i=1}^{k} \frac{y_i - \theta}{v_i} = 0 \quad \Rightarrow \quad \hat\theta_{FE} = \frac{\sum w_i y_i}{\sum w_i}, \qquad w_i = \frac{1}{v_i}$$

So the pooled estimate is a weighted average, and each trial's weight is the inverse of its variance: precise trials count for more. The second derivative of $\ell(\theta)$ is $-\sum w_i$, which gives

$$\text{Var}(\hat\theta_{FE}) = \frac{1}{\sum w_i}$$

This is the same curvature argument as in [The Core of Bayesian Estimation](10_Bayesian_Estimation.qmd): every trial adds its weight to the curvature of the log-likelihood, so the interval narrows as trials accumulate.

### Random Effects: When Trials Estimate Different Effects

Trials differ in populations, doses and settings, so the true effect may differ between them. The **random-effects** model lets each trial have its own true effect $\theta_i$, drawn from a common distribution:

$$y_i \sim N(\theta_i, v_i), \qquad \theta_i \sim N(\mu, \tau^2)$$

Here $\mu$ is the average effect across settings and $\tau^2$ the between-study variance. $v_i$ is the within-trial variance computed from $a, b, c, d$: it describes how $y_i$ varies around trial $i$'s own true effect $\theta_i$. A single trial has only one $\theta_i$, so it cannot show how true effects differ between trials; $\tau^2$ is estimated from how much the trials scatter beyond what their own $v_i$ can explain, as described in @sec-tau2.

Integrating out $\theta_i$ gives

$$y_i \sim N(\mu, \; v_i + \tau^2)$$

which is the fixed-effect model with every variance increased by $\tau^2$. The same algebra therefore applies, with weights $w_i^* = 1 / (v_i + \tau^2)$:

$$\hat\mu_{RE} = \frac{\sum w_i^* y_i}{\sum w_i^*}, \qquad \text{Var}(\hat\mu_{RE}) = \frac{1}{\sum w_i^*}$$

Adding the same $\tau^2$ to every trial matters most for the trials whose own $v_i$ is small. Large trials lose weight, small trials gain it, and the weights become more equal.

### Measuring Heterogeneity: Q, τ² and I²

The random-effects model needs a value for $\tau^2$, and the choice between the two models depends on how much the trials disagree. Three quantities describe that disagreement, and each answers a different question:

| Quantity | Question it answers | Scale |
|:---|:---|:---|
| Cochran's $Q$ | Do the trials disagree more than chance alone would explain? | A test statistic |
| $\tau^2$ | How far apart are the true effects? | Variance of the log OR |
| $I^2$ | What share of the observed scatter is real rather than chance? | A proportion, 0 to 100% |

::: {#nte-choose .callout-note collapse="true"}
# Choosing between the two models

In order of what should carry the most weight:

1.  **Clinical judgement, first.** Do the trials plausibly estimate one common effect: similar populations, doses, comparators, outcome definitions and follow-up? If they clearly do not, the fixed-effect assumption is hard to defend and the random-effects model is the honest default. A fixed-effect analysis then needs a positive argument, not just a favourable statistic.
2.  **The size of the heterogeneity, not its significance.** Read $\tau$ and the prediction interval (@sec-pi), and ask whether the spread they describe is large enough to change a decision. A statistically significant $Q$ with a prediction interval that stays well away from no effect is a different situation from a non-significant $Q$ with a prediction interval that crosses it.
3.  **Model fit.** Compare the two models on residual deviance and DIC: lower is better, and the DIC already penalises the extra parameter the random-effects model brings. NICE DSU TSD 2 recommends consulting both and sets no numerical threshold for how large a difference counts; the figure of 3 to 5 points often quoted is a rule of thumb from the wider DIC literature, not from TSD 2. When the two models are close, TSD 2's advice elsewhere is to decide on the data-generating process and the interpretability of the results rather than on the statistic.
4.  **How many trials there are.** With few trials $\tau$ is badly estimated, and a DerSimonian–Laird random-effects interval can be too narrow (@sec-tau2). The remedies are the Hartung–Knapp–Sidik–Jonkman adjustment, REML, or a Bayesian model with an informative prior on $\tau$. Switching to a fixed-effect model is not a remedy: it assumes the problem away.

Three things not to do: choose by whether $Q$ is significant, choose by an $I^2$ threshold alone, or choose the model that happens to give the narrower interval. NICE DSU TSD 2 sets out the fit-based comparison; the [Standard Bayesian NMA](01_Standard_Bayesian_NMA_Guide.qmd) guide lists the same judgement in a network setting.
:::

#### Cochran's Q: More Scatter Than Chance?

$Q$ adds up how far each trial sits from the fixed-effect estimate:

$$Q = \sum_{i=1}^{k} w_i (y_i - \hat\theta_{FE})^2 = \sum_{i=1}^{k} \left( \frac{y_i - \hat\theta_{FE}}{\sqrt{v_i}} \right)^2$$

The second form shows what each term is: the distance of trial $i$ from the pooled estimate, measured in that trial's own standard errors, then squared. A small trial far from the pooled estimate adds little, because its standard error is large; a large trial the same distance away adds a lot.

If every trial estimates one common effect, each standardised distance behaves like a draw from $N(0, 1)$, and the sum of $k$ squared standard Normal draws follows a $\chi^2$ distribution. One degree of freedom is lost because $\hat\theta_{FE}$ is itself estimated from the same trials, so

$$Q \sim \chi^2_{k-1}, \qquad E(Q) = k - 1 \quad \text{under homogeneity}$$

That gives a test: a $Q$ well above $k - 1$ is unlikely if the trials share one effect. The test has two weaknesses that pull in opposite directions:

- **Few trials, low power.** With five trials, $Q$ has 4 degrees of freedom and real heterogeneity often fails to reach significance. A threshold of $p < 0.10$ rather than $0.05$ is commonly used for this reason, and a non-significant $Q$ is not evidence that the trials agree.
- **Many large trials, too much power.** With precise trials, differences too small to matter clinically still produce a significant $Q$.

Either way, $Q$ says whether heterogeneity is detectable, not how large it is.

#### τ²: How Far Apart Are the True Effects? {#sec-tau2}

$\tau^2$ is the variance of the true log ORs across settings, and $\tau$ their standard deviation. It is the quantity with a direct clinical reading: under the random-effects model, about 95% of true effects lie within $\mu \pm 1.96\,\tau$ on the log scale. On the OR scale that is a multiplicative range. With $\tau = 0.3$, for example, true ORs span a factor of $e^{1.96 \times 0.3} \approx 1.8$ either side of the average OR: an average OR of 2 would be consistent with true ORs from about 1.1 to 3.6 in different settings.

The **DerSimonian–Laird** estimator obtains $\tau^2$ from $Q$. Under the random-effects model, the expected value of $Q$ rises in a straight line with $\tau^2$:

$$E(Q) = (k - 1) + \tau^2 \left( \sum w_i - \frac{\sum w_i^2}{\sum w_i} \right)$$

The first term is the scatter chance alone produces; the second is the extra scatter from real differences. Setting the observed $Q$ equal to its expected value and solving gives

$$\hat\tau^2_{DL} = \max\left(0, \; \frac{Q - (k - 1)}{\sum w_i - \sum w_i^2 / \sum w_i}\right)$$

This is a method-of-moments estimator: it matches one observed statistic to its expectation, with no likelihood involved. Two consequences follow:

- **Truncation at zero.** By chance, $Q$ can fall below $k - 1$, which would give a negative variance. The estimate is then set to 0. An estimate of $\hat\tau^2 = 0$ therefore means the scatter did not exceed chance, not that the true effects are identical.
- **One number, treated as known.** The random-effects weights and interval use $\hat\tau^2$ as if it were the true value. With few trials the estimate is very imprecise, and that imprecision is not carried into the pooled interval. Restricted maximum likelihood (REML) and Paule–Mandel are common alternative estimators; the Bayesian model in this guide instead gives $\tau$ a posterior distribution.

::: {.callout-note collapse="true"}
# Deriving E(Q) and the DerSimonian–Laird estimator

Write $W = \sum w_i$ and $\hat\theta = \sum w_i y_i / W$. Under the random-effects model, $y_i = \mu + (\theta_i - \mu) + \varepsilon_i$ with $\text{Var}(y_i) = v_i + \tau^2$, and the trials are independent.

**Step 1: split Q.** Adding and subtracting $\mu$ inside the square, the cross term cancels because $\sum w_i (y_i - \hat\theta) = 0$:

$$Q = \sum w_i (y_i - \mu)^2 - W (\hat\theta - \mu)^2$$

**Step 2: the first term.** Each $(y_i - \mu)^2$ has expectation $v_i + \tau^2$, and $w_i v_i = 1$:

$$E\left[\sum w_i (y_i - \mu)^2\right] = \sum w_i (v_i + \tau^2) = k + \tau^2 W$$

**Step 3: the second term.** $\hat\theta$ is a weighted average of independent $y_i$, so

$$\text{Var}(\hat\theta) = \frac{\sum w_i^2 (v_i + \tau^2)}{W^2} = \frac{W + \tau^2 \sum w_i^2}{W^2}$$

using $w_i^2 v_i = w_i$. Since $\hat\theta$ is unbiased for $\mu$, $E[W(\hat\theta - \mu)^2] = W\,\text{Var}(\hat\theta) = 1 + \tau^2 \sum w_i^2 / W$.

**Step 4: combine.**

$$E(Q) = (k + \tau^2 W) - \left(1 + \tau^2 \frac{\sum w_i^2}{W}\right) = (k - 1) + \tau^2 \left( W - \frac{\sum w_i^2}{W} \right)$$

With $\tau^2 = 0$ this is the $k - 1$ of the homogeneity test. Replacing $E(Q)$ with the observed $Q$ and solving for $\tau^2$ gives the DerSimonian–Laird estimator.
:::

#### I²: What Share of the Scatter Is Real? {#sec-i2}

$$I^2 = \max\left(0, \; \frac{Q - (k - 1)}{Q}\right)$$

$Q$ is the total scatter, $k - 1$ is the part chance would produce, and the difference is the part attributed to real differences. $I^2$ is that difference as a share of the total. The Cochrane Handbook gives rough bands, deliberately overlapping: 0–40% might not be important, 30–60% may be moderate, 50–90% may be substantial, and 75–100% considerable.

The most common misreading is to treat $I^2$ as a measure of how different the true effects are. It is not, because it compares $\tau^2$ with the precision of the trials. The same $I^2$ can be written as

$$I^2 = \frac{\hat\tau^2}{\hat\tau^2 + s^2}$$

where $s^2$ is a typical within-trial variance. The same true spread therefore gives very different values of $I^2$ depending on trial size. With $\tau^2 = 0.09$ ($\tau = 0.3$):

| Trials | Typical within-trial variance $s^2$ | $I^2$ |
|:---|:---:|:---:|
| Small | 0.20 | $0.09 / 0.29 \approx 31\%$ |
| Large | 0.02 | $0.09 / 0.11 \approx 82\%$ |

Large trials leave little room for chance, so almost all the scatter they show counts as real. Two further cautions apply:

- **I² is itself uncertain.** With few trials its confidence interval commonly spans most of the 0–100% range, so a single value such as 59% should not be read to the nearest percentage point.
- **I² does not tell you what to expect in a new setting.** For that, $\tau$ and the prediction interval below are the relevant quantities.

::: {.callout-note collapse="true"}
# Why I² equals τ² / (τ² + s²)

Write $c = W - \sum w_i^2 / W$, the denominator of the DerSimonian–Laird estimator, so that $\hat\tau^2 = (Q - (k - 1)) / c$. Higgins and Thompson (2002) define the typical within-trial variance as

$$s^2 = \frac{(k - 1)\,W}{W^2 - \sum w_i^2} = \frac{k - 1}{c}$$

Then, whenever $Q > k - 1$,

$$\frac{\hat\tau^2}{\hat\tau^2 + s^2} = \frac{(Q - (k - 1)) / c}{(Q - (k - 1)) / c + (k - 1) / c} = \frac{Q - (k - 1)}{Q} = I^2$$

So $I^2$ and $\hat\tau^2_{DL}$ carry the same information from $Q$; $I^2$ simply rescales it by the precision of the trials.
:::

### The Prediction Interval {#sec-pi}

The confidence interval for $\mu$ describes uncertainty about the **average** effect. A decision about a new population needs the range of effects to expect in **one new setting**, which also includes the spread $\tau^2$:

$$\hat\mu_{RE} \pm 1.96 \sqrt{\hat\tau^2 + \text{Var}(\hat\mu_{RE})}$$

With substantial heterogeneity, the prediction interval can include no effect even when the confidence interval does not.

### The Bayesian Version: The Same Model, Estimated Differently {#sec-bayes}

The models above reduce each trial to two numbers, $y_i$ and $v_i$, and treat $y_i$ as Normal with known variance. The Bayesian model used for NMA skips that reduction and works directly with the number of responders and patients in each arm, through a Binomial likelihood:

$$r_{iA} \sim \text{Binomial}(n_{iA}, p_{iA}), \qquad r_{iB} \sim \text{Binomial}(n_{iB}, p_{iB})$$

$$\text{logit}(p_{iA}) = \mu_i, \qquad \text{logit}(p_{iB}) = \mu_i + \delta_i, \qquad \delta_i \sim N(d, \tau^2)$$

This is the model of the [Standard Bayesian NMA](01_Standard_Bayesian_NMA_Guide.qmd) guide with a single comparison, and its unknowns are the ones already met above under different names: $d$ is the average effect ($\mu$ in the random-effects model), $\delta_i$ is trial $i$'s own true effect ($\theta_i$), and $\tau$ is the spread between trials. Only the baselines $\mu_i$ are new, because reducing each trial to $y_i$ had already removed them by taking a difference within the trial. Note that $\mu_i$ here is a baseline log odds, not the average effect $\mu$ of the random-effects model; the clash of letters is the convention used in the NMA literature.

What changes is not the model but what is treated as unknown. The frequentist approach treats $\mu$ and $\tau$ as fixed constants to be estimated, and $\theta_i$ as a random draw that is never itself reported. The Bayesian approach treats all of them as unknown quantities with a joint posterior distribution:

$$
\begin{aligned}
p(d, \tau, \delta, \mu \mid \text{counts}) \;\propto\;
& \prod_{i=1}^{k} P(r_{iA}, r_{iB} \mid \mu_i, \delta_i) && \text{likelihood} \\[4pt]
\times \; & \prod_{i=1}^{k} N(\delta_i \mid d, \tau^2) && \text{between-trial distribution} \\[4pt]
\times \; & \; p(d)\; p(\tau) \prod_{i=1}^{k} p(\mu_i) && \text{priors}
\end{aligned}
$$ {#eq-post}

There is nothing to solve in ([-@eq-post]); MCMC maps it by sampling, exactly as in [The Core of Bayesian Estimation](10_Bayesian_Estimation.qmd).

Counting the unknowns: with $k$ trials there are $2k + 2$ of them, a baseline and a true effect for each trial plus $d$ and $\tau$. The baselines are nuisance parameters, each identified by its own trial's control arm and never pooled; the frequentist model removes them by taking a difference within each trial. The $\delta_i$ are not free either: the between-trial distribution holds them together, which is what makes $d$ and $\tau$ estimable and what stops each $\delta_i$ from simply reproducing that trial's own log OR.

Two things follow, and both show up in the results below.

::: {.callout-note collapse="true"}
# With τ held fixed, the two approaches give the same answer

Take the Normal approximation of the earlier sections, $y_i \mid \theta_i \sim N(\theta_i, v_i)$ and $\theta_i \sim N(\mu, \tau^2)$, so that marginally $y_i \sim N(\mu, v_i + \tau^2)$. Hold $\tau^2$ fixed and put a flat prior on $\mu$, $p(\mu) \propto 1$. The posterior is then proportional to the likelihood:

$$p(\mu \mid y) \;\propto\; \prod_{i=1}^{k} \exp\left(-\frac{(y_i - \mu)^2}{2(v_i + \tau^2)}\right) = \exp\left(-\frac{1}{2}\sum_{i=1}^{k} w_i^* (y_i - \mu)^2\right)$$ {#eq-post-mu}

with $w_i^* = 1/(v_i + \tau^2)$ as before. Expanding the sum in ([-@eq-post-mu]) and completing the square in $\mu$ gives a Normal posterior:

$$\mu \mid y \;\sim\; N\left( \frac{\sum w_i^* y_i}{\sum w_i^*}, \; \frac{1}{\sum w_i^*} \right)$$ {#eq-post-normal}

The posterior mean in ([-@eq-post-normal]) is exactly the random-effects estimate $\hat\mu_{RE}$, and the posterior variance is exactly $\text{Var}(\hat\mu_{RE})$. So the DerSimonian–Laird interval is the credible interval of this Bayesian model under a flat prior with $\tau$ fixed at $\hat\tau$.

The Bayesian model in this guide differs from that special case in two ways: it gives $\tau$ a prior and integrates over it rather than fixing it, and it uses the Binomial counts rather than the Normal approximation. The first difference is what widens the interval in the results below.
:::

::: {.callout-note collapse="true"}
# Shrinkage: each trial's own effect is revised

The posterior also reports each $\delta_i$, which the frequentist model never estimates. Using the same Normal approximation, with $\mu$ and $\tau^2$ treated as known, the posterior for one trial's true effect combines two sources: its own result $y_i$, with precision $1/v_i$, and the distribution the trials come from, with precision $1/\tau^2$. Weighting each by its precision,

$$E(\theta_i \mid y_i) = B_i \, y_i + (1 - B_i)\, \mu, \qquad B_i = \frac{\tau^2}{\tau^2 + v_i}$$ {#eq-shrink}

Each trial's estimate is pulled from its own result towards the average of the trials. How far it is pulled depends on the comparison in $B_i$:

- A **small trial** has a large $v_i$, so $B_i$ is small and its result is pulled a long way towards $\mu$. An extreme result from a small trial is largely discounted.
- A **large trial** has a small $v_i$, so $B_i$ is close to 1 and its result stands nearly as observed.
- If $\tau^2 \to 0$, every $B_i \to 0$ and all trials collapse onto $\mu$, which is the fixed-effect model.
- If $\tau^2$ is large, $B_i \to 1$ and the trials are left alone, as if each were analysed separately.

In an NMA this is the mechanism behind a question EAGs often ask: how much of the estimate for a treatment supported by one small trial comes from that trial, and how much is borrowed from the rest of the network.
:::

## The Minimal Mock Case: Five Trials of B vs A

- **Treatment A**: Standard of Care
- **Treatment B**: New Drug
- **Outcome**: response (higher is better), so an OR above 1 favours B

Five trials of different sizes compared B with A:

```{r data}
library(knitr)

trials <- data.frame(
    study = paste("Study", 1:5),
    n_A = c(120, 60, 300, 80, 150),
    r_A = c(24, 15, 75, 16, 30),
    n_B = c(120, 60, 300, 80, 150),
    r_B = c(45, 28, 88, 34, 60)
)

kable(trials, col.names = c("Study", "Patients (A)", "Responders (A)", "Patients (B)", "Responders (B)"))
```

Study 3 is by far the largest trial, and, as the next section shows, the one closest to no effect. That combination is what makes the choice between fixed and random effects visible.

## Implementation: From Scratch, `metafor` and Stan

### From Scratch in Base R

Every quantity from the theoretical framework, computed directly:

```{r from-scratch}
fmt <- function(x) sprintf("%.2f", x)
ci_text <- function(est, se) sprintf("%.2f (%.2f, %.2f)", exp(est), exp(est - 1.96 * se), exp(est + 1.96 * se))

# Cells of each 2x2 table
resp_B <- trials$r_B
non_B <- trials$n_B - trials$r_B
resp_A <- trials$r_A
non_A <- trials$n_A - trials$r_A

# Effect size and variance per trial
y <- log((resp_B / non_B) / (resp_A / non_A)) # log OR
v <- 1 / resp_B + 1 / non_B + 1 / resp_A + 1 / non_A
k <- length(y)

# Fixed effect
w <- 1 / v
lor_fe <- sum(w * y) / sum(w)
se_fe <- sqrt(1 / sum(w))

# Heterogeneity
Q <- sum(w * (y - lor_fe)^2)
Q_p <- pchisq(Q, df = k - 1, lower.tail = FALSE)
tau2 <- max(0, (Q - (k - 1)) / (sum(w) - sum(w^2) / sum(w))) # DerSimonian-Laird
I2 <- max(0, (Q - (k - 1)) / Q)

# Random effects
w_re <- 1 / (v + tau2)
lor_re <- sum(w_re * y) / sum(w_re)
se_re <- sqrt(1 / sum(w_re))

# 95% prediction interval, on the OR scale
pi_half <- 1.96 * sqrt(tau2 + se_re^2)
pi_lo <- exp(lor_re - pi_half)
pi_hi <- exp(lor_re + pi_half)

w_fe_pct <- 100 * w / sum(w)
w_re_pct <- 100 * w_re / sum(w_re)

kable(
    data.frame(trials$study, exp(y), y, v, w_fe_pct, w_re_pct),
    digits = c(0, 2, 3, 3, 1, 1),
    col.names = c("Study", "OR", "log OR", "Variance", "Weight, FE (%)", "Weight, RE (%)")
)

kable(
    data.frame(Q, k - 1, Q_p, tau2, 100 * I2),
    digits = c(2, 0, 3, 3, 1),
    col.names = c("Cochran's Q", "df", "p-value", "$\\tau^2$", "$I^2$ (%)")
)

kable(
    data.frame(
        c("Fixed effect", "Random effects (DerSimonian-Laird)"),
        c(ci_text(lor_fe, se_fe), ci_text(lor_re, se_re)),
        c("&ndash;", sprintf("%.2f to %.2f", pi_lo, pi_hi))
    ),
    col.names = c("Model", "Pooled OR (95% CI)", "95% prediction interval")
)
```

Study 3 has the smallest variance, so under the fixed-effect model it carries `r sprintf("%.0f%%", w_fe_pct[3])` of the weight. With $\hat\tau^2 =$ `r sprintf("%.3f", tau2)` added to every variance, its share falls to `r sprintf("%.0f%%", w_re_pct[3])`, while the smallest trial, Study 2, rises from `r sprintf("%.0f%%", w_fe_pct[2])` to `r sprintf("%.0f%%", w_re_pct[2])`.

### Cross-Check With `metafor`

The standard R package gives the same numbers:

```{r metafor-check, message=FALSE}
library(metafor)

es <- escalc(measure = "OR", ai = r_B, n1i = n_B, ci = r_A, n2i = n_A, data = trials)
fit_fe <- rma(yi, vi, data = es, method = "FE")
fit_re <- rma(yi, vi, data = es, method = "DL")

kable(
    data.frame(
        c("log OR, fixed effect", "SE, fixed effect", "log OR, random effects", "SE, random effects", "$\\tau^2$", "$I^2$ (%)"),
        c(lor_fe, se_fe, lor_re, se_re, tau2, 100 * I2),
        c(coef(fit_fe), fit_fe$se, coef(fit_re), fit_re$se, fit_re$tau2, fit_re$I2)
    ),
    digits = 4,
    col.names = c("Quantity", "From scratch", "metafor")
)
```

The forest plot shows each trial with its 95% CI, the random-effects pooled estimate as a diamond, and the prediction interval as the dotted line through it:

```{r forest, fig.height=4}
forest(
    fit_re,
    slab = trials$study,
    atransf = exp,
    at = log(c(0.5, 1, 2, 4, 8)),
    addpred = TRUE,
    header = c("Study", "OR [95% CI]")
)
```

### Bayesian Random Effects in Stan

::: {.callout-note collapse="true" appearance="minimal"}
# Stan model: pairwise_re.stan

```{r show-stan, echo=FALSE, class.output="stan"}
cat(readLines("pairwise_re.stan"), sep = "\n")
```
:::

Three choices in the model are worth stating:

- **Priors**: $N(0, 10^2)$ for the baselines $\mu_i$ and the pooled log OR $d$, which is vague on the log-odds scale; half-Normal$(0, 1)$ for $\tau$, which keeps implausibly large heterogeneity unlikely without ruling it out. NICE DSU TSD 2 uses wider conventions, $N(0, 100^2)$ for $\mu_i$ and $d$ and $\sigma \sim \text{Uniform}(0, 2)$ for the between-trial standard deviation; the tighter priors here are chosen to sample stably with only five trials, and a submission would report a sensitivity analysis over both.
- **Non-centred parameterisation**: $\delta_i = d + \tau z_i$ with $z_i \sim N(0, 1)$. It describes the same model as $\delta_i \sim N(d, \tau^2)$, but samples far more reliably when $\tau$ is poorly identified, as it is with five trials.
- **Prediction**: `OR_new` draws a log OR for a new study from $N(d, \tau^2)$ at every iteration, giving the Bayesian counterpart of the prediction interval.

```{r stan-fit, message=FALSE, warning=FALSE}
library(rstan)

stan_data <- list(
    S = nrow(trials),
    n_A = trials$n_A, r_A = trials$r_A,
    n_B = trials$n_B, r_B = trials$r_B
)

fit_bayes <- stan(
    file = "pairwise_re.stan",
    data = stan_data,
    chains = 4,
    iter = 4000,
    warmup = 2000,
    seed = 42,
    control = list(adapt_delta = 0.99),
    refresh = 0
)

bayes_sum <- summary(fit_bayes, pars = c("d", "OR", "tau", "OR_new"))$summary
kable(
    bayes_sum[, c("mean", "sd", "2.5%", "50%", "97.5%", "n_eff", "Rhat")],
    digits = c(3, 3, 3, 3, 3, 0, 3),
    col.names = c("Mean", "SD", "2.5%", "Median", "97.5%", "ESS", "R-hat")
)

n_div <- sum(sapply(get_sampler_params(fit_bayes, inc_warmup = FALSE), function(x) sum(x[, "divergent__"])))
```

All R-hat values are at or below `r sprintf("%.3f", max(bayes_sum[, "Rhat"]))`, the smallest ESS is `r round(min(bayes_sum[, "n_eff"]))`, and the sampler reported `r n_div` divergent transitions.

## Results & Interpretation {#results}

```{r compare}
post <- rstan::extract(fit_bayes, pars = c("OR", "OR_new", "tau"))
b_or <- quantile(post$OR, c(0.5, 0.025, 0.975))
b_new <- quantile(post$OR_new, c(0.025, 0.975))
b_tau <- quantile(post$tau, c(0.5, 0.025, 0.975))

kable(
    data.frame(
        c("Fixed effect", "Random effects (DerSimonian-Laird)", "Random effects (Bayesian)"),
        c(ci_text(lor_fe, se_fe), ci_text(lor_re, se_re), sprintf("%.2f (%.2f, %.2f)", b_or[1], b_or[2], b_or[3])),
        c("&ndash;", sprintf("%.2f to %.2f", pi_lo, pi_hi), sprintf("%.2f to %.2f", b_new[1], b_new[2])),
        c("0 (assumed)", sprintf("%.2f", sqrt(tau2)), sprintf("%.2f (%.2f, %.2f)", b_tau[1], b_tau[2], b_tau[3]))
    ),
    col.names = c("Model", "Pooled OR (95% interval)", "95% prediction interval", "$\\tau$ (between-study SD)")
)
```

Three readings of this table matter for an HTA submission.

**Fixed vs random effects.** The fixed-effect OR (`r fmt(exp(lor_fe))`) sits closer to Study 3 than the random-effects OR (`r fmt(exp(lor_re))`), because the fixed-effect model gives the largest, least favourable trial `r sprintf("%.0f%%", w_fe_pct[3])` of the weight. Which is right is not a statistical question alone: if the trials plausibly differ in population or setting, the fixed-effect assumption of one common effect is hard to defend, and its narrower interval is not a real gain in precision.

**Confidence vs prediction interval.** The random-effects 95% CI (`r sprintf("%.2f to %.2f", exp(lor_re - 1.96 * se_re), exp(lor_re + 1.96 * se_re))`) describes the average effect across settings like these trials. The prediction interval (`r sprintf("%.2f to %.2f", pi_lo, pi_hi)`) describes the effect to expect in one new setting, and its lower end sits at `r fmt(pi_lo)`. B helps on average; whether it helps in a particular new population is far less certain. A cost-effectiveness model that uses only the pooled estimate carries the first statement, not the second.

**Frequentist vs Bayesian random effects.** The Bayesian 95% interval for the pooled OR (`r sprintf("%.2f to %.2f", b_or[2], b_or[3])`) is wider than the DerSimonian–Laird CI. The two are not different models: as @sec-bayes shows, holding $\tau$ fixed and using a flat prior reproduces the DerSimonian–Laird estimate and interval exactly. The difference is that DerSimonian–Laird plugs in one value of $\tau$ (`r fmt(sqrt(tau2))`) as if it were known, while the Bayesian model carries the uncertainty about $\tau$ itself, whose 95% interval runs from `r fmt(b_tau[2])` to `r fmt(b_tau[3])`. With five trials, that uncertainty is large, and ignoring it overstates precision.

## HTA Audit Points & Interview Q&A {#audit}

::: {.callout-important collapse="true"}
# Should I choose fixed or random effects based on I²?

No. I² is a supporting signal, not a switch, and a low I² with few trials is weak evidence of homogeneity, because $Q$ has little power to detect heterogeneity when $k$ is small. The judgement starts from whether the trials plausibly estimate one common effect; see @nte-choose for the full order.
:::

::: {.callout-important collapse="true"}
# What does an I² of around 60% actually tell you?

That around 60% of the scatter between trials is real rather than chance, not how far apart the true effects are: that depends on trial size as well, and $\tau$ or the prediction interval is the measure to use. See @sec-i2.
:::

::: {.callout-important collapse="true"}
# Why is DerSimonian–Laird criticised when there are few trials?

It estimates $\tau^2$ from one moment equation and then treats the estimate as fixed, so the confidence interval ignores the uncertainty in $\tau^2$. With few trials that uncertainty is large, and the interval is too narrow. Common remedies are the Hartung–Knapp–Sidik–Jonkman adjustment, which rescales the variance and uses a $t$ distribution, usually widening the interval, restricted maximum likelihood for $\tau^2$, or a Bayesian model that carries the uncertainty in $\tau$ through, as above.
:::

::: {.callout-important collapse="true"}
# When is a pairwise meta-analysis enough, and when do you need an NMA?

A pairwise meta-analysis is enough when the decision problem is one comparison and there are head-to-head trials of it. Once the decision involves several treatments, or the comparator of interest has never been tested directly against the new treatment, the evidence has to be connected through common comparators, which is what an [NMA](01_Standard_Bayesian_NMA_Guide.qmd) does.
:::

## Conclusion {#conclusion}

Pairwise meta-analysis comes down to a weighted average with two decisions attached: how to weight the trials, and whether they share one effect. The fixed-effect model weights by precision alone; the random-effects model adds between-study variance, which evens out the weights, widens the interval and makes the prediction interval available. The Bayesian version works from the counts and carries the uncertainty in $\tau$, and it is the same model that a [Standard Bayesian NMA](01_Standard_Bayesian_NMA_Guide.qmd) extends to a network of comparisons.

**Methodological References**: NICE DSU TSD 2 (Dias et al., 2011, updated 2016); *Cochrane Handbook for Systematic Reviews of Interventions*, Chapter 10; DerSimonian & Laird (1986); Higgins, Thompson & Spiegelhalter (2009) on prediction intervals.
